### 2008-07-29 thesis/abstract.tex
Checked in work on thesis.

\documentclass{article}

\usepackage[english]{babel}
\usepackage{qtree}
\usepackage[utf8x]{inputenc}
\usepackage{linguex}
\usepackage{natbib}

\bibliographystyle{plainnat}

\author{Aleksandar Dimitrov}
\title{Questioning Discourse in Prolog: From Questions to Queries}

\begin{document}

\maketitle

Discourse Representation Theory (DRT), as presented in \citet{kampreyle:drt}
provides a powerful and flexible approach to capturing the meaning of natural
language and storing information in tidy data structures.
\citep{blackburnbos:cl2} further provide us with an implementation of DRT in
Prolog. Although unfinished, their results may already prove useful to an
attempt of an implementation of a generic way to retrieve information from
Discourse Representation Structures (DRS) in form of either absolute truth
values (answers to alternative questions) in questions like \ref{qex:yesno} or a
possible set of individuals as in \ref{qex:set}.

\ex. 
\a. Is Mary pregnant? \label{qex:yesno}
\b. Who is coming to the party? \label{qex:set}

The problem of treating questions has generally been recognized as being
non-trivial and there exist several attempts to explain the semantics of
questions in natural language, the classical examples being \citet{hamblin:1973}
with a first rather simple attempt, \citet{karttunen:1973} and \citet{gs:1984a},
who both take a much more sophisticated approach to the proper treatment of
questions.

One of the main challenges of this work will be the unification of classical
linguistic theories with the newer DRT. Both originate from entirely different
mindsets. While the classic theory according to \citep{karttunen:1973} is based
on Montague's analysis of English (in \citet{ptq}) and thus presupposes
intensional logic, DRT introduces its own formalism and does so far not treat
intensionality. Trying to let the classic theory operate on DRT in a naïve way
will thus fail. This makes it imperative to find a feasible approach to let
question operators work with DRSs in a well-defined way.

As a further challenge, implementing presuppositions and/or modality might prove
to be an interesting task.

\ex. \label{presups}
\a. Is Jon pregnant? \label{presup}
\b. Can Jon be pregnant? \label{modal}

In \ref{presups} we see two sentences which should be answered on grounds of
presuppositions - Jon is presumably male and can thus not be pregnant. In
\ref{presup}, an answer to the question may not even need to be proved against
the current model, while a sentences as \emph{Jon became pregnant} should be
rejected by the system. Cases like \ref{modal} might be much harder to properly
treat, since they involve modal verbs.

While treating presuppositions in DRT is not new\footnote{\citet[Chap.
4]{blackburnbos:cl2} provide a discussion on the implementation in Prolog},
modality has so far not received great attention\footnote{Albeit a few papers do
seem to exist. See \citet{frank96context}.}. Since questions in natural language
rely quite a lot on presuppositions, implementing at least a primitive method
seems to be an enticing opportunity to further explore the treatment of natural
language questions to DRSs.

Based on \citet{jaeger:1997} and possibly others, one might also want to
consider implementing \emph{informative questions}, that is, questions that not
only query for informations, but also \emph{provide} information to the system.

\bibliography{aleksbib}

\end{document}

### 2008-07-29 thesis/paper.tex
Checked in work on thesis.

\documentclass{scrartcl}

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{bbm,qtree,natbib,stmaryrd,aleks,linguex}

\newcommand{\abbr}{\textsf}
\newcommand{\pn}{\textsc}
\newcommand{\url}[1]{\texttt{http://#1}}
%names
\newcommand{\curt}{\pn{Curt}}
\newcommand{\prol}{\pn{Prolog}}

\bibliographystyle{plainnat}

\author{Aleksandar Dimitrov}
\title{From Questions to Queries -- Enhancing the Curt System by a Theory and Implementation of Embedded WH-Questions}
\begin{document}

\maketitle
\tableofcontents

\section{Choosing Grounds} % FIXME title sucks

\subsection{The Curt System}

\subsection{General Comparison of the Curt System}

This section will briefly discuss how the features of the Curt system compare to
other similar systems or systems with the same goal, namely implementing a
system that is able to model a logic representation of human language.

\subsection{Early Attempts: \pn{Shrdlu} and \pn{Cyc}}
%TODO: need some reference for this section!

Interest in deep semantic parsing of natural language has existed since the very
beginning of the informational age. One of the first systems to achieve public
recognition, the \pn{Shrdlu} system

\section{Concepts}

After having elaborated on the grounds of the implementation, we will now
display the implementation's underlying concepts. % FIXME this sentence is too short

%OUTLINE Presenting a conceptual overview, while arguing for and against stuff.

First we will present a brief overview of the considerations regarding the
choice of an appropriate formalism for the system in section \ref{sec:formal}
to then compare it with already existing implementations of question answering
systems in \ref{sec:othercrap}.

\subsection{Formalism}\label{sec:formal}

Most of the traditional literature in natural language semantics follows
\cite{ptq} and thus also uses its underlying formalism, \emph{Intensional Logic}
(henceforth \abbr{IL}). \cite{gs:sqpa} (and subsequent publications) however use
two-sorted type logic (henceforth \abbr{Ty2}) as proposed in
\cite{gallin:ty2}. \cite{z:ilty2} discusses the differences and similarities in expressive power
between \abbr{Ty2} and \abbr{IL} and comes to the conclusion that the set of
expressions expressible in the former but not in the latter might not be
relevant for the treatment of natural language semantics. Although he proposes
a translation scheme between both languages, translation from \abbr{Ty2} to
\abbr{IL} seems to sometimes produce very complex \abbr{IL} terms.

Since this paper is not concerned with possible extensions of the underlying
theories of treating natural language questions, but rather with their
implementation in the programming language \pn{Prolog} several additional
constraints have to be considered. Among them the \emph{feasibility} of any
possible approach takes highest precedence.

While the \pn{Curt} system is unique in its approach to deep semantic analysis, its
architecture has been crafted in a pragmatic manner. The authors chose to use
external tools where it was possible in order to avoid reimplementation of
existing functionality and be able to focus on novel parts of the
implementation\footnote{The name of the \pn{Curt} system reflects this choice.
It stands for \emph{\textbf{C}lever \textbf{u}se of \textbf{r}easoning
\textbf{t}ools}.}. While there exist inferencing tools for modal
logic\footnote{See %todo todo and todo 
for discussions. The interested reader is also referred to \pn{Molle}
(\url{molle.sf.net}), a modal theorem prover that comes with a graphical user
interface that allows interactive investigation of the constructed indices.},
a restructuring of the \pn{Curt} system would to take advantage of them would be
beyond the scope of this thesis. Another design decision of the \pn{Curt} system
was to use \prol as an implementation language. While \prol is a general
purpose logic programming language, it is rooted in first-order predicate logic
and does not explicitly support modal logic\footnote{But see%todo, todo and todo
for a discussion of the implementation of modal logic into \prol}.

It was therefore decided to avoid modal logic and implement a pragmatic approach
to representing the indices of higher-order logic in \prol in section
\ref{sec:indices}.

\cite{g:inqs} presents an approach to implementing inquisitive semantics in
predicate logic.

\subsection{Existing Implementations of Similar Systems}\label{sec:othercrap}

Godot, all that statistical crap and Pereira's CHAT-80.

\section{Implementation}

\subsection{An Ontology of Embedding Verbs}

Karttunen's List.

\subsubsection{Extensional WH-embedding Verbs}

Like \emph{know}.

\subsubsection{Intensional WH-embedding Verbs}

Like \emph{believe}.

\subsection{Modality and Possible Worlds}
\label{sec:indices}
\subsection{Treatment of Baker-Ambiguities}

\section{Discussion}

\bibliography{aleksbib}

\end{document}

### 2008-07-29 thesis/pres.tex
Checked in work on thesis.

\documentclass{beamer}

\mode<beamer>{
\usetheme{Singapore}
%Berlin
%\usecolortheme{rose}
}

\usepackage[english]{babel}
\usepackage{qtree}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{amsfonts,amsmath}
\usepackage{linguex}
\usepackage{color}

\newcommand{\stress}{\textbf}
\title{From Questions to Queries: Asking Prolog the Natural Way}
\institute{Universität Tübingen}
\author{Aleksandar L. Dimitrov}

\begin{document}
\begin{frame}
  \titlepage
\end{frame}

\begin{frame}
  \frametitle{Outline}
  \tableofcontents
\end{frame}

\section{What We Have So Far}

\subsection{The Curt System}
\begin{frame}
\frametitle{The Curt System so Far}
  \begin{itemize}
    \item[$\to$] Newest version: Sensitive Curt
    \item \stress{Features:}
    \begin{itemize}
      \item Uses external tools to provide enhanced feature set
      \item Can thus \stress{detect inconsistent additions} to existing discourse
      \item and \stress{detect uninformative additions} to existing discourse
    \end{itemize}
    \item \stress{Shortcomings:}
    \begin{itemize}
      \item Does not account for $\alpha$-equivalent readings
      \item Does not know about worlds and situations
      \item Has no background knowledge apart from the discourse
      \item Is hard to interface with as a normal human
    \end{itemize}
  \end{itemize}
\end{frame}

\section{Blackburn \& Bos: Extending Curt}

\subsection{Scrupulous Curt}
\begin{frame}
\frametitle{Scrupulous Curt}
  \begin{itemize}
    \item Recognizes ``a man likes a hash bar'' is not ambiguous
    \item \ldots by proving ``$\leftrightarrow$'' between the two readings
    \item Only disambiguates at request
    \item Does not produce models after disambiguation
    \item Is unable to construct a model where one holds and the other doesn't
    \item \emph{Failure to prove something does not prove the non-existance of a proof}
  \end{itemize}
\end{frame}

\subsection{Knowledgeable Curt}
\begin{frame}
\frametitle{Knowledgeable Curt: Know What's Going on in the Human World}
  \begin{itemize}
    \item Knowledgeable Curt introduces \stress{background knowledge}
    \item Features
    \begin{itemize}
      \item Distinguishes between \emph{plants} and \emph{humans}
      \item Adds new types of knowledge to our discourse: \stress{it takes knowledge}
      \item[$\to$] from \pause the lexicon\pause, the world\pause{} and the situational
      context
    \end{itemize}
    \item Our new discourse:
    \begin{equation}
    LK \wedge WK \wedge SK \wedge DK
    \end{equation}
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Lexical Knowledge}
  Keywords: \emph{Lexical Semantics}, \emph{ontology}
  \begin{itemize}
    \item Knowledgeable Curt introduces \emph{Ontologies}
    \item \Tree [.Organism Animal Plant [.Person Man Woman ] ]
    \item Ontologies can be \emph{subject of dispute}
    \item \texttt{lexicalKnowledge.pl} uses \alert{triggers} to enhance
      efficiency.
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Lexical Knowledge, Cntd.}
  \begin{itemize}
    \item A particular ontology can only be applicable to a certain domain
    \item Ontologies can help distinguish \stress{homonyms}
    \item[$\to$] $\forall X [\texttt{grunt}(X) \wedge \texttt{person}(X)]$ vs.  $\forall X [ \texttt{grunt}(X) \wedge \texttt{utterance}(X) ]$
    vs. $\forall X [\texttt{grunt}(X) \wedge \texttt{fish}(X)]$
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{World Knowledge}
  \begin{itemize}
    \item Being a \texttt{wife} implies being \texttt{married}
    \item Distinction between \stress{world knowledge} and \stress{lexical
    knowledge} can be fuzzy:
    \begin{itemize}
      \item being a \texttt{student} implies being an \texttt{undergraduate}.
      \item \ldots{} but couldn't a \texttt{student} be a child of \texttt{undergraduate}?
    \end{itemize}
    \item \stress{Inheritance relationships:} no thing is a subpart of itself, no thing is a subpart of two distinct things.
    \item \stress{Really?} No thing is owned by itself? No thing is owned by two
    distinct things?
    \item[$\to$] Distinguish between notions of \emph{have}.
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Situational Knowledge}
  \begin{itemize}
    \item Discourse knowledge
    \begin{itemize}
      \item When we talk about two different things, we don't talk about the
	same thing.
    \end{itemize}
    \item All three types of knowledge can interact in interesting
      ways\ldots{}\pause
    \item consider: \emph{every car has a radio}.
  \end{itemize}
\end{frame}

\subsection{Helpful Curt}
\begin{frame}
\frametitle{Helpful Curt: Asking the Natural Way}
  \begin{itemize}
    \item For now, \texttt{readings/0} and \texttt{models/0} were the only way of
    querying Curt
    \item How about natural language questions?
    \begin{enumerate}
      \item Is Mia married?
      \item Who owns a hash bar?
      \item Does every man like a gun?
      \item Who did not shoot Marvin?
    \end{enumerate}
    \item Building models where possible answers hold
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Helping Curt Help}
  \begin{itemize}
    \item Introducing the \texttt{que/3} predicate.
    \begin{itemize}
      \item basically a \emph{variable binder}:
      \item[$\to$] $que(X, person(X), not(shot(X. marvin))).$
      \item[$\to$] $que(X, customer(X), some(Y, and(fdshake(Y), order(X,Y)))).$
    \end{itemize}
    \item No first order terms or formulas
    \item[$\to$] Translate:
    \begin{equation}
      que(X,R,N) \to some(X, and (R,N))
    \end{equation}
    \item[$\to$] Now we know whether the question makes sense - and if we can
    answer it.
    \item Next: free the variable and unify it with the model: $satisfy(X, and(R,N))$
  \end{itemize}
\end{frame}

\section{Excursion: Question Semantics}
\begin{frame}
\frametitle{Question Semantics}
  Two main theories:
  \begin{enumerate}
    \item<1-| alert@3> Lauri Karttunen
    \item<2-> Groenendijk \& Stokhof
  \end{enumerate}
\end{frame}

\subsection{Karttunen 77}
\begin{frame}
  \frametitle{Karttunen 77: Types of Questions}
  \begin{itemize}
    \item<1-> Two (syntactic) types of questions:
      \begin{enumerate}
	\item<2-|alert@4> Direct questions
	\item<3-|alert@5> Indirect questions
      \end{enumerate}
    \item<4-> \emph{Which book did Mary read?}, \emph{Who is John meeting?}
    \item<5-> I ask\ldots{} \emph{\dots{} whether it is raining},
      \emph{\dots{} who John met}
    \item<6-> Karttunen treats only \alert{indirect questions}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Karttunen 77: Types of Questions, contd.}
  \begin{itemize}
    \item<1-> Two (semantic) types of questions:
      \begin{enumerate}
	\item<2-|alert@4> Alternative questions \only<5->{\stress<5>{Yes/No questions}}
	\item<3-|alert@6> Wh-questions
      \end{enumerate}
    \item<4-> \emph{Does Mary smoke?}, \emph{Will you come to the party?}
    \item<6-> \emph{Who owns a hash bar?}, \emph{Who is Mia married to?}
    \item<7-> Karttunen considers \stress{alternative questions} syntactically
      degenerate variants of \stress{Wh-questions}.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{A Simplistic Overview of Karttunen 77}
  \begin{itemize}
    \item Karttunen generalises all questions to \emph{proto-questions}
      \begin{itemize}
	\item ?Mary reads
      \end{itemize}
    \item He introduces three question rules: PQ, AQ, YNQ (and some more\dots)
    \item PQ: For every phrase $\phi$ of type $t$, ?$\phi$ is of type Q
    \item ?$\phi$ is the set of propositions that are the intension of $\phi$
      and for which the extension of $\phi$ holds
    \item AQ: For every disjunction of $\phi_1$ to $\phi_n$, ?$\phi_1$ to
      $\phi_n$ is the set of propositions for which either proto-question holds.
      (whether Mia snorts or Butch walks)
    \item YNQ: For every phrase $\phi$, ?$\phi$ is a set of propositions equal
      to either the intension of $\phi$ or the intension of $\phi$
      \emph{negated} \alert{and} the extension of $\phi$
      (whether Mia snorts or not)
  \end{itemize}
\end{frame}

%\begin{frame}
  %\frametitle{An Ontology of Embedding Verbs}
  %Verbs of\ldots
  %\begin{itemize}
    %\item \emph{retaining knowledge}: know, recall, forget
    %\item \emph{acquiring knowledge}: learn find out
    %\item \emph{conjecture}: guess, estimate
    %\item \emph{decision}: decide, determine, agree on
    %\item \emph{relevance}: matter, care, be significant
    %\item \emph{dependency}: depend on, make a difference to
    %\item \ldots
  %\end{itemize}
%\end{frame}

\subsection{Intensional Logic}

\begin{frame}
  \frametitle{Intensional Logic}
  Intensional logic is needed to interpret
  \begin{itemize}
    \item tense
    \item aspect
    \item modality
    \item object intensional verbs like \emph{seek}
    \item \ldots
  \end{itemize}
\end{frame}

\section{Thesis Topic: More Curt}
\begin{frame}
  \frametitle{Thesis Topic: Extend Curt}
  \begin{itemize}
    \item Curt's way of inferring background knowledge is demanding
    \item Curt is unaware of the informative value of questoins
    \item Curt cannot handle more than one situation
    \item Curt has no idea of accessible worlds
    \item Curt can only handle very simple questoins
    \item Curt's background knowledge is static
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Meaning of Life: Make Curt Even More Awesome}
  \begin{itemize}
    \item Let Curt handle more and more syntactical constructions
    \item Implement plurals
    \item Implement tense \& aspect
    \item Make him feel \texttt{:-)}
  \end{itemize}
\end{frame}

\subsection{Ideas: Advertent Curt}
\begin{frame}
\frametitle{Advertent Curt: Be Smart About Information}
  \begin{itemize}
    \item Curt should handle \emph{informative questions}.
    \item[$\to$] \emph{Does George Bush like Jaques Chirac?}
    \item What does Mia do?
    \item What can Mia do?
  \end{itemize}
\end{frame}

\subsection{More Ideas: Studious Curt}
\begin{frame}
\frametitle{Spiritual Curt: Learning from Discourse}
  \begin{itemize}
    \item Knowledgebale Curt is stubborn: provide a way to re-learn
    \item Curt should be able to start blank and improve from there
    \item Awarness of distinction between \emph{world} knowledge and
    \emph{situational} knowledge
  \end{itemize}
\end{frame}

\subsection{Even More Ideas: Spiritual Curt}
\begin{frame}
\frametitle{Spiritual Curt: Finding Possible and Accessible Worlds}
  \begin{itemize}
    \item Support \emph{more than one} situation
    \item \emph{believe}, \emph{may}, \emph{can}, \emph{seek}
    \item John seeks a unicorn? Every unicorn is a king of France!
    \item Extend Curt to be \stress{intensional}
  \end{itemize}
\end{frame}

\subsection{Even More More Ideas: Inquisitive Curt}
\begin{frame}
  \frametitle{Inqusitive Curt: Detect Inconsistencies and Unadequate Discourse}
  \begin{itemize}
    \item What if we don't say something about an entity in either
    \emph{world}, \emph{situational} or \emph{lexical} knowledge?
    \item What if for an item not every relationship it \emph{can} engage on is
    fulfilled?
    \begin{itemize}
      \item Is Mia married?
    \end{itemize}
  \end{itemize}
\end{frame}

\end{document}

### 2008-07-29 src/intensionalModels.pl
Added intensional models stub

@@ -1,0 +1,8 @@
+% Implementation of intensional models.
+
+module (intensionalModels).
+
+% unit tests
+
+% give a list of models given Ty2 expression containing indices
+

### 2008-07-29 thesis/paper.tex
added more outlines.

@@
 \usepackage[utf8x]{inputenc}
-\usepackage{bbm,qtree,natbib,stmaryrd,aleks,linguex}
+\usepackage{bbm,qtree,natbib,stmaryrd,amsmath,amssymb,aleks,linguex}
 
@@
 \newcommand{\curt}{\pn{Curt}}
-\newcommand{\prol}{\pn{Prolog}}
+\newcommand{\prol}{\pn{Prolog} }
 
@@
 
+\pn{Curt}, a system written in \prol that is able to interpret basic utterances
+in natural language and express them in terms of first order logic, is developed
+throughout chapter six of \cite{blackburnbos:cl1}. While relatively simple in
+terms of lexical and grammatical coverage it is able to perform basic checks for 
+\emph{consistency} and \emph{informativity} of its input\footnotemark, within
+the bounds of the tools it uses.
+
+\footnotetext{Here, the notion of \emph{consistency} is meant as \emph{validity}
+of the input and \emph{informativity} denotes \emph{validity}. Thus, if
+$\mathcal{D}$, the discourse so far and $\phi$ a new input token (fully parsed
+sentence), then consistency means $\mathcal{D} \to \phi$ holds and informativity
+means $\mathcal{D} \nvDash \phi$.}
 \subsection{General Comparison of the Curt System}
@@
 
+We first derive know(john,all(X,imp(woman(X),love(X,john)))) from the syntax of
+john knows whether every woman loves john, then we see if
+all(X,imp(woman(X),love(X,john))) is valid (uninformative) and ridicule John, if
+it's not valid and not consistent, assert his knowledge of the fact that it's
+not true. If it's true, we assert he knows that.
+
+For whether we can ridicule him if it's not true (inconsistent).
+\newcommand{\discourse}{\ensuremath{\mathcal{D}}}
+The MB/TP do a comparison \discourse to all(X,imp(woman(X),love(X,john)))
+
 \subsubsection{Intensional WH-embedding Verbs}
@@
 \label{sec:indices}
+
+We have seen that some embedding verbs may not be interpreted without
+implementing a higher order formalism that is able to model the notion of
+\emph{possible} and \emph{accessible} worlds. While foobar %fixme! Nelken &
+% Francez 2001
+argue for an extensional treatment of questions, they already % n & f 2006
+retract their claims and admit that a ``minimal account of intensionality'' has
+to be assumed. While they propose a two-world system later adopted in
+\cite{g:is}, this is not going to suffice if we want to model each individual's
+knowledge in order to be able to discriminate people's knowledge and beliefs.
+%fixme bla bla
+
 \subsection{Treatment of Baker-Ambiguities}
+
+\subsection{Alternative Questions -- Inquisitive Semantics}
+
+\cite{g:is} presents a novel approach to treating questions in natural
+semantics: \pn{Inquisitive Semantics}. Groenendijk presents a new logic
+formalism aimed at providing a wider coverage of the intentions of question in
+human language. The main source of the ``inquisitiveness'' of statements is
+given by the interpretation of \emph{disjunction}. We will now show how to
+incorporate this novel approach into the \pn{Curt} system.
+
+We will, however depart from the path we have been following throughout this
+paper and first implement a theory for direct questions. Only then will we extend
+it to also cover embedded alternative questions.
 

### 2008-08-06 thesis/paper.tex
Further work. Structure should be OK
will need some changes still, but I'm content with the general concept.
Tuscany is as beautiful as ever. Hope I'll come back when I don't have
to read so many papers...

@@
-\documentclass{scrartcl}
+\documentclass[a4paper,fontsize=12pt]{article}
 
@@
 \usepackage[utf8x]{inputenc}
-\usepackage{bbm,qtree,natbib,stmaryrd,amsmath,amssymb,aleks,linguex}
+\usepackage{amsthm,bbm,qtree,natbib,stmaryrd,amsmath,amssymb,aleks,linguex}
 
-\newcommand{\abbr}{\textsf}
-\newcommand{\pn}{\textsc}
-\newcommand{\url}[1]{\texttt{http://#1}}
+\newcommand{\abbr}{\textsf} % how to render abbreviations like IL or Ty2
+\newcommand{\stress}{\textbf} % put stress somewhere (as opposed to \emph)
+\newcommand{\term}{\textsf} % introduce a term
+\newcommand{\code}{\texttt} % layout of in-line code snippets
+%\newcommand{\quote}[1]{\textit{``#1''}} % a direct quote from a paper
+\newcommand{\pn}{\textsc} % how to render proper names like Shrdlu
+\newcommand{\url}[1]{\code{http://#1}} % how urls appear
+
+% commonly used phrases
+\newcommand{\Disc}{\ensuremath{\mathcal{D}}} % the discourse
+\newcommand{\wh}{\textsc{wh}} %
+
 %names
-\newcommand{\curt}{\pn{Curt}}
-\newcommand{\prol}{\pn{Prolog} }
+\newcommand{\curt}{\pn{Curt}\mbox{ }}
+\newcommand{\acurt}{\pn{Advertent Curt}\mbox{ }}
+\newcommand{\prol}{\pn{Prolog}\mbox{ }}
+
+%theorem environments
+\theoremstyle{remark} \newtheorem*{termin}{Definition} % terminology. Could use a
+%better title fixme
 
@@
 
-\author{Aleksandar Dimitrov}
+\author{Aleksandar Lyubenov Dimitrov\\
+$\langle$\href{mailto:aleks.dimitrov@googlemail.com}{aleks.dimitrov@googlemail.com}$\rangle$}
 \title{From Questions to Queries -- Enhancing the Curt System by a Theory and Implementation of Embedded WH-Questions}
+
+\usepackage[pdftex,colorlinks,citecolor=magenta]{hyperref}
 \begin{document}
-
 \maketitle
 \tableofcontents
+\newpage
 
-\section{Choosing Grounds} % FIXME title sucks
+\section*{Introduction}
 
-\subsection{The Curt System}
+%fixme: to be formalized
+Where we use ``natural language'', usually the English language will be treated
+in practice, though it would be nice to come up with treatments of other
+languages. We abstract away from pragmatic affairs and concern ourselves with
+mostly the purely semantic/logic domain, so this should be feasible to achieve.
+\subsection*{Acknowledgements}
+
+\section{Concepts}
+
+% fixme: formalize
+A brief overview of the premisses for the theory (which itself lays ground to
+the implementation).
+
+\subsection{The Curt System} \label{sec:curt}
 
 \pn{Curt}, a system written in \prol that is able to interpret basic utterances
-in natural language and express them in terms of first order logic, is developed
-throughout chapter six of \cite{blackburnbos:cl1}. While relatively simple in
+in natural language and store them in expressions of first order logic, is presented
+in \cite{blackburnbos:cl1}. While relatively limited when compared to other
+systems presented in \ref{sec:comparision} in
 terms of lexical and grammatical coverage it is able to perform basic checks for
@@
 
+% FIXME: don't really like this footnote
 \footnotetext{Here, the notion of \emph{consistency} is meant as \emph{validity}
 of the input and \emph{informativity} denotes \emph{validity}. Thus, if
-$\mathcal{D}$, the discourse so far and $\phi$ a new input token (fully parsed
-sentence), then consistency means $\mathcal{D} \to \phi$ holds and informativity
-means $\mathcal{D} \nvDash \phi$.}
-\subsection{General Comparison of the Curt System}
+\Disc, the discourse so far and $\phi$ a new input token (fully parsed
+sentence), then consistency means $\Disc \to \phi$ holds and informativity
+means $\Disc \nvDash \phi$.}
+
+It can thus detect incoherent discourse and would reject a sequence of
+sentences such as \emph{``Mary likes every man. Peter is a man. Mary does not
+like Peter''}, claiming it found a contradiction in its input. For every input
+token, \pn{Curt} will also try to infer whether the current discourse would
+entail the information given in its newest token and is thus irrelevant with
+regard to previous utterances -- effectively allowing it to
+reduce the size of its internal representation of the discourse by not adding
+clauses to it which would not contribute to the overall informational value
+already present in the program.
+
+The system is introduced during the final chapter of \cite{blackburnbos:cl1},
+incrementally increasing in features and complexity. Support for external
+inferencing via theorem provers and model builders is added in a step-by-step
+fashion and finally an ontology and a straight-forward way of parsing and
+answering direct questions for noun phrases conclude its development. The theory
+and implementation of this paper is meant as a replacement and enhancement for
+the latter feature. This will also happen in an incremental manner and new
+features will be introduced one at a time.
+
+\subsection{General Comparison of the Curt System}\label{sec:comparision}
 
@@
 
-\subsection{Early Attempts: \pn{Shrdlu} and \pn{Cyc}}
+\subsubsection{Early Attempts: \pn{Shrdlu} and \pn{Cyc}}
 %TODO: need some reference for this section!
@@
 
-\section{Concepts}
+While the \pn{Curt} system is unique in its approach to deep semantic analysis, its
+architecture has been crafted in a pragmatic manner. The authors chose to use
+external tools where it was possible in order to avoid reimplementation of
+existing functionality and be able to focus on novel parts of the
+implementation\footnote{The name of the \pn{Curt} system reflects this choice.
+It stands for \emph{\textbf{C}lever \textbf{u}se of \textbf{r}easoning
+\textbf{t}ools}.}. This also allows for great flexibility in adopting new
+solutions and improvements made in the field of inferencing tools for first
+order predicate logic.
+
+\subsection{Extending the Curt System}
+
+At the end of \cite{blackburnbos:cl1}, \curt is able to cope with a small
+grammar and provides some features for interacting with humans in natural
+language. It does only understand basic extensional transitive verbs - modal
+verbs or the treatment of questions are not implemented. Anaphora resolution
+still remains an open issue. The grammar is rather limited and \curt only comes
+with a basic vocabulary oriented around the setting of a popular movie.
+
+The aim of this paper is to introduce the reader into question semantics as it
+will be relevant for implementing a basic capability of treating intensional
+verbs like `to \stress{know} whether something is the case' and `to
+\stress{believe} that something might be the case' and then present such an
+implementation within the framework of the \curt system. Further and more
+advanced investigations will be dealt with at the end of this discussion or left
+for further research.
+
+\section{Theory}
+
+\subsection{Terminology}
+
+Following \cite{gs:q}, we will not use the natural natural language word
+``question'' but distinguish between \term{questions},
+\term{interrogative sentences} and \term{interrogative acts} in the following
+way:
+
+\begin{termin}
+  An \term{interrogative} is a certain type of sentence in natural language
+  distinguished by certain features\footnote{To quote \cite{gs:q}: \quote{[An
+  interrogative sentence is] characterized by word order, intonation, question
+  mark, the occurrence of interrogative pronouns}.}
+  whereas an \term{interrogative act} refers to the utterance of an
+  \emph{interrogative}. Finally, a 
+  \term{question} will denote the semantic entity uttered by the
+  \emph{interrogative act}, whichever form it may have.
+\end{termin}
+
+\abbr{Ty2} is two-sorted type logic\footnote{See \cite{gallin:ty2} for a
+discussion and \cite{gs:sqpa} for its usage in describing the semantics of
+questions}. \abbr{IL} is intensional logic\footnote{As used by \cite{ptq}}. 
+An embedded \wh-phrase looks
+like:%fixme
+
+\ex. John \stress{knows} \emph{whether} good examples are hard to come up with.
 
@@
 
+% FIXME if we have a `Terminology' section, all those `henceforths' aren't really
+% necessary anymore.
+
 Most of the traditional literature in natural language semantics follows
-\cite{ptq} and thus also uses its underlying formalism, \emph{Intensional Logic}
-(henceforth \abbr{IL}). \cite{gs:sqpa} (and subsequent publications) however use
-two-sorted type logic (henceforth \abbr{Ty2}) as proposed in
-\cite{gallin:ty2}. \cite{z:ilty2} discusses the differences and similarities in expressive power
-between \abbr{Ty2} and \abbr{IL} and comes to the conclusion that the set of
-expressions expressible in the former but not in the latter might not be
-relevant for the treatment of natural language semantics. Although he proposes
-a translation scheme between both languages, translation from \abbr{Ty2} to
-\abbr{IL} seems to sometimes produce very complex \abbr{IL} terms.
+\cite{ptq} and thus also makes use of its underlying formalism,
+\emph{Intensional Logic} (henceforth \abbr{IL}). \cite{gs:sqpa} (and many subsequent
+publications) however also use two-sorted type logic (henceforth \abbr{Ty2}) as
+proposed in \cite{gallin:ty2}. \cite{z:ilty2} discusses the differences and
+similarities in expressive power between \abbr{Ty2} and \abbr{IL} and comes to
+the conclusion that the set of expressions contained in the former but not in
+the latter might not be relevant for the treatment of natural language
+semantics. Although he proposes a translation scheme between both languages,
+translation from \abbr{Ty2} to \abbr{IL} seems to sometimes produce very complex
+\abbr{IL} terms and is therefore avoided here. In fact, \abbr{IL} will be only
+marginally relevant for the presentation of \acurt -- only the type system
+presented in \cite{ptq} is carried over.
+
+% OK till now, but the rest is a fixme :-(
 
@@
 
-While the \pn{Curt} system is unique in its approach to deep semantic analysis, its
-architecture has been crafted in a pragmatic manner. The authors chose to use
-external tools where it was possible in order to avoid reimplementation of
-existing functionality and be able to focus on novel parts of the
-implementation\footnote{The name of the \pn{Curt} system reflects this choice.
-It stands for \emph{\textbf{C}lever \textbf{u}se of \textbf{r}easoning
-\textbf{t}ools}.}. While there exist inferencing tools for modal
-logic\footnote{See %todo todo and todo 
+As was already mentioned in \ref{sec:curt}, Curt is based on an
+implementation and the use of external tools only capable of treating first
+order logic -- an extensional framework. \prol on the other hand, is a language
+that can very well make use of higher order formalisms for
+calculation\footnote{The two most popular predicates in \prol used for
+implementing higher-order logic are
+\code{call/n} and \code{apply/3}. This paper is based on \pn{SWI-Prolog} which
+provides an implementation of the former, but not of the latter. See
+\cite{naish:prolhio} for a discussion of higher order functional programming in
+\prol.}.
+
+While there exist inferencing tools for modal logic\footnote{See %todo todo and todo 
 for discussions. The interested reader is also referred to \pn{Molle}
-(\url{molle.sf.net}), a modal theorem prover that comes with a graphical user
+(\url{http://molle.sf.net}), a modal theorem prover that comes with a graphical user
 interface that allows interactive investigation of the constructed indices.},
-a restructuring of the \pn{Curt} system would to take advantage of them would be
-beyond the scope of this thesis. Another design decision of the \pn{Curt} system
-was to use \prol as an implementation language. While \prol is a general
-purpose logic programming language, it is rooted in first-order predicate logic
-and does not explicitly support modal logic\footnote{But see%todo, todo and todo
-for a discussion of the implementation of modal logic into \prol}.
+a restructuring of the \pn{Curt} system to take advantage of them would be
+beyond the scope of this thesis.
 
@@
 
-\cite{g:inqs} presents an approach to implementing inquisitive semantics in
+\cite{g:is} presents an approach to implementing inquisitive semantics in
 predicate logic.
 
-\subsection{Existing Implementations of Similar Systems}\label{sec:othercrap}
+\subsection{On the Proper Treatment of Questions}
 
-Godot, all that statistical crap and Pereira's CHAT-80.
+Question Semantics is a much-disputed but often overlooked field of modern
+semantic research. Often question semantics has been reduced to an issue of
+pragmatics, treating direct questions (interrogatives) as nothing more than a
+paraphrase of indirect questions (embedded interrogatives).\cite{tichy}
+even goes as far as claiming equality in semantic value between
+indicatives and interrogatives. The benefit of such an approach would be the
+strict reliance on traditional analyses, where interrogatives would only infer a
+different `attitude' in the speaker's utterance. Putting this attitude into
+context and making it yield the correct results would then be left to
+pragmatics.
+This approach has been challenged and proven wrong by \cite{gs:q},
+who also argue that the proper treatment of questions would require extensions
+to the existing logic frameworks used in natural language semantics semantics.
+More recent publications, such as \cite{gal:tmbi}
+address this topic and aim to introduce notions of \term{information change
+potential} as opposed to the traditional focus of natural language semantics on
+\term{truth values}.
 
-\section{Implementation}
+Nevertheless, progress has been very lively since the first
+formal approaches to the field started with \cite{hamblin:q} and other
+publications such as % todo
+. \cite{gs:q} provide a very descriptive overview of the field of question
+semantics
+
+\subsubsection{Classification of Interrogatives}
+
+When treating questions one will be confronted with many different
+categorizations of questions. One of the first distinctions to make is the one
+between \term{embedded questions} as in \ref{ex:embed} and \term{direct
+questions} shown in \ref{ex:direct}.
+
+\ex. \a. John knows \emph{whether this sentence contains an embedded
+question}.\label{ex:embed}
+\b. Does this sentence pose a direct question?\label{ex:direct}
+
+As indicated already by the title, this paper will concern itself with the
+treatment of the former and only briefly comment on the latter. The reasons for
+this are numerous  %fixme I need a synonym for that
+and this section is going to name the most important ones.
+
+Early literature, such as \cite{karttunen:1977} only regards embedded
+\wh-phrases, treating direct questions as semantically equivalent. To quote
+Karttunen:
+
+\begin{quote}A direct question can be treated as semantically equivalent to a
+certain kind of declarative sentence containing the corresponding indirect
+question embedded under  a suitable `performative' verb.\end{quote}
+
+Using this strategy, Karttunen escapes the treatment of direct  questions and
+thus gives an approach for indirect questions only.
 
@@
 
-Karttunen's List.
+\cite{karttunen:1977} introduces means of classification for \wh-embedding
+verbs, the nature of which \cite{lahiri:diss} covers in a much broader scope.
+The different verb classes ask for different strategies of treating the
+\wh-clauses they embed and also pose different restrictions on the type of
+clause they expect.
+
+Our aim is to improve coverage of grammatical structures and thus when faced
+with the necessity to make compromise about the complexity of our grammar, we
+will choose overgeneration in favor of unmotivated complexity of the grammar
+itself.
 
@@
 For whether we can ridicule him if it's not true (inconsistent).
-\newcommand{\discourse}{\ensuremath{\mathcal{D}}}
-The MB/TP do a comparison \discourse to all(X,imp(woman(X),love(X,john)))
+The MB/TP do a comparison \Disc to all(X,imp(woman(X),love(X,john)))
 
@@
 Like \emph{believe}.
+
+\section{Implementation}
+
+\subsection{Existing Implementations of Similar Systems}\label{sec:othercrap}
+
+Enhancing \curt by an implementation dealing with questions in natural language
+sparks interest in whether there have been similar efforts made in the past. No
+further extensions to the \curt system aiming at comparable goals are known to the
+author at the time of writing. However, there is a lively interest in providing
+so-called question answering or QA-systems since the early stages of computer
+science. % fixme; cite some ancient 60's paper
+This section discusses a few of the more elaborate approaches: first an early
+but already quite impressive implementation is shown, followed by a newer approach
+using embodied agents and finally an overview over the many database-based
+QA-systems is given. Finally, all this is compared what the current extension to
+\curt can do.
+
+\subsubsection{Godot, the Talking Robot}
+
+\subsubsection{CHAT-80}
+
+\subsubsection{Modern Intensional Database Approaches}
+
+\subsubsection{Advertent Curt}
+
+\acurt lends its name from the fact that it is not only aware of its own
+knowledge and presuppositions, but `knows' about what individuals in its
+model of the world are aware of. It is not in this sense a question
+\emph{answering} system, but it may be seen as an early effort towards providing
+such a system on a more sophisticated scale. While section \ref{sec:altq} shows
+how direct questions can be integrated into the system with relative ease, focus
+is still on providing a feasible method of implementing intensionality in what
+is mostly a first-order logic computing environment.
+
+It is also not constrained to being useful in a particular domain, like
+\pn{CHAT-80} is.
 
@@
 
-\subsection{Treatment of Baker-Ambiguities}
+\subsubsection{Enhancing Storage}
 
-\subsection{Alternative Questions -- Inquisitive Semantics}
+\curt parses sentences of natural language using an enhanced version of \pn{Cooper
+storage}\footnote{See \cite{cooper:storage2} for the original implementation.} as
+presented in \cite{keller:storage}. \cite{blackburnbos:cl1} call this version
+\pn{Keller storage}. That name will be used from here % FIXME sucks. Please change
+to refer to this particular implementation as presented in
+\cite{blackburnbos:cl1}. Although \pn{Keller storage} is already quite capable
+of dealing with more complex constituents such as nested noun phrases, it needs
+some enhancements in order to be able to treat verbs embedding whole
+sentences or \wh-phrases correctly. The aim is to enable \curt's implementation of
+\pn{Keller storage} to deal with examples such as \ref{ex:kel1}\footnotemark.
 
-\cite{g:is} presents a novel approach to treating questions in natural
-semantics: \pn{Inquisitive Semantics}. Groenendijk presents a new logic
+\ex. \label{ex:kel1} Mia knows whether Vincent snorts.\\
+\Tree [.t [.T Mia ] [.IV [.{IV/$\bar{t}$} knows ] [.{$\bar{t}$} whether [.t [.T
+Vincent ] [.IV snorts ] ] ] ] ]
+
+\footnotetext{Note that \ref{ex:kel1} is shown using the same syntactic type system used in
+\cite{gs:sawhq}, which extends the system given in \cite{ptq} by $\bar{t}$ to
+denote embedded \wh-constituents.}
+
+\subsection{Treatment of Baker Ambiguities}
+
+\subsection{Alternative Questions -- Inquisitive Semantics}\label{sec:altq}
+
+\cite{g:is} introduces a novel approach to treating questions in natural
+semantics: \pn{Inquisitive Semantics}. He presents a new logic
 formalism aimed at providing a wider coverage of the intentions of question in
@@
 
+We have seen, the Curt system is quite capable of treating questions this way.
+Having implemented modality and possible worlds in such a fashion, let's go on
+and implement some even cooler stuff (tense, aspect, negation, comparatives,
+anaphora). Basing all this on a nice implementation of DRT (discussed in
+\cite{kampreyle:drt}) like the one given in \cite{blackburnbos:cl2} might be
+interesting.
+
+\section{Appendix}
+
+\subsection{The Source Code}
+
+\subsection{Whatnot}
+
 \bibliography{aleksbib}

### 2008-08-09 src/englishGrammar.pl
incremental commit.

@@ -25,15 +25,15 @@
    Texts
 ========================================================================*/
 
-t([sem:T])--> 
+t([sem:T])-->
    s([coord:no,sem:S]),
    {combine(t:T,[s:S])}.
 
-t([sem:T])--> 
+t([sem:T])-->
    s([coord:yes,sem:S]),
    {combine(t:T,[s:S])}.
 
-t([sem:T])--> 
+t([sem:T])-->
    q([sem:Q]),
    {combine(t:T,[q:Q])}.
 
@@ -87,7 +87,6 @@ sinv([gap:G,sem:S])-->
    vp([coord:_,inf:inf,num:Num,gap:G,sem:VP]), 
    {combine(sinv:S,[av:Sem,np:NP,vp:VP])}.
 
-
 /*========================================================================
    Questions
 ========================================================================*/
@@ -101,7 +100,6 @@ q([sem:Sem])-->
    whnp([num:_,sem:NP]), 
    sinv([gap:[np:NP],sem:S]),
    {combine(q:Sem,[sinv:S])}.
-
 
 /*========================================================================
    Noun Phrases
@@ -211,11 +209,29 @@ vp([coord:no,inf:Inf,num:Num,gap:[],sem:VP])-->
    iv([inf:Inf,num:Num,sem:IV]), 
    {combine(vp:VP,[iv:IV])}.
 
-vp([coord:no,inf:I,num:Num,gap:G,sem:VP])-->   
+vp([coord:no,inf:I,num:Num,gap:G,sem:VP])-->
    tv([inf:I,num:Num,sem:TV]), 
    np([coord:_,num:_,gap:G,sem:NP]), 
    {combine(vp:VP,[tv:TV,np:NP])}.
 
+/*
+ * WH-Embedding verb phrases: take an IV/t-bar verb and a tbar
+ * and produce a verb phrase. We check for type agreement, in order to prevent
+ * something like 'believe if'. 
+ *
+ * We produce finite iv.
+ */
+vp([coord:no,inf:fin,num:Num,gap:[],sem:VP])
+-->	ivt([coord:_,type:Type,num:Num,sem:IVT])
+	, tbar([coord:_,type:Type,sem:TBAR])
+	, {combine(iv:VP,[ivt:IVT,tbar:TBAR])}
+.
+% coord:no to exclude a faux pas like 'whether either'
+% TODO: but 'John knows that either Bill likes Mary or …'? is this grammatical?
+tbar([coord:no,type:Type,sem:TBAR])
[…]

### 2008-08-09 src/englishLexicon.pl

@@ -105,6 +105,8 @@ lexEntry(pn,[symbol:thewolf,syntax:[the,wolf]]).
 lexEntry(pn,[symbol:vincent,syntax:[vincent]]).
 lexEntry(pn,[symbol:vincent,syntax:[vincent,vega]]).
 lexEntry(pn,[symbol:yolanda,syntax:[yolanda]]).
+lexEntry(pn,[symbol:john,syntax:[john]]).
+lexEntry(pn,[symbol:mary,syntax:[mary]]).
 
 
 /*========================================================================
@@ -113,7 +115,6 @@ lexEntry(pn,[symbol:yolanda,syntax:[yolanda]]).
 
 lexEntry(qnp,[symbol:person,syntax:[who],mood:int,type:wh]).
 lexEntry(qnp,[symbol:thing,syntax:[what],mood:int,type:wh]).
-
 
 /*========================================================================
    Intransitive Verbs
@@ -296,18 +297,26 @@ lexEntry(whemb,[syntax:[that],mood:int,type:ass]).
    WH-embedding verbs - extensional
 ========================================================================*/
 
-lexEntry(ivt,[symbol:know,type:extensional,syntax:[know],inf:inf:num:sg]).
-lexEntry(ivt,[symbol:know,type:extensional,syntax:[knows],inf:fin:num:sg]).
+lexEntry(ivt,[type:ex,symbol:know,syntax:[know],inf:inf,num:sg]).
+lexEntry(ivt,[type:ex,symbol:know,syntax:[knows],inf:fin,num:sg]).
+lexEntry(ivt,[type:ex,symbol:know,syntax:[know],inf:inf,num:pl]).
 
-lexEntry(ivt,[symbol:tell,type:extensional,syntax:[tell],inf:inf:num:sg]).
-lexEntry(ivt,[symbol:tell,type:extensional,syntax:[tells],inf:fin:num:sg]).
+lexEntry(ivt,[type:ex,symbol:tell,syntax:[tell],inf:inf,num:sg]).
+lexEntry(ivt,[type:ex,symbol:tell,syntax:[tells],inf:fin,num:sg]).
+lexEntry(ivt,[type:ex,symbol:tell,syntax:[tell],inf:inf,num:pl]).
 
 /*========================================================================
    WH-embedding verbs - intensional
[…]

### 2008-08-09 src/semLexStorage.pl

@@ -46,7 +46,7 @@ semLex(noun,M):-
         sem:[lam(X,Formula)]],
    compose(Formula,Sym,[X]).
 
-semLex(iv,M):-
+semLex(ivM):-
    M = [symbol:Sym,
         sem:[lam(X,Formula)]],
    compose(Formula,Sym,[X]).

### 2008-08-09 src/semRulesLambda.pl

@@ -56,6 +56,8 @@ combine(n:app(A,B),[adj:A,n:B]).
 combine(n:A,[noun:A]).
 combine(n:app(B,A),[noun:A,nmod:B]).
 
+%combine().
+
 combine(nmod:A,[pp:A]).
 combine(nmod:A,[rc:A]).
 combine(nmod:lam(P,app(A,app(B,P))),[pp:A,nmod:B]).

### 2008-08-15 thesis/paper.tex
changed bibtex markup
trying to put it all into xetex
... for better looks ;-)

@@
-\documentclass[a4paper,fontsize=12pt]{article}
+\documentclass[a4paper]{article}
 
 \usepackage[english]{babel}
-\usepackage[utf8x]{inputenc}
-\usepackage{amsthm,bbm,qtree,natbib,stmaryrd,amsmath,amssymb,aleks,linguex}
+
+\usepackage{amssymb,bbm,qtree,graphicx,natbib,aleks,linguex}
+
+%% note that this file depends on being processed with XeTeX
+\usepackage{xltxtra,xunicode,fontspec}
+
+% font specifications (don't overuse ;-)
+\defaultfontfeatures{Mapping=tex-text} % converts LaTeX specials (``quotes'' --- dashes etc.) to unicode
+\setromanfont [Ligatures={Common,Rare}]{Gentium}
 
@@
 \newcommand{\code}{\texttt} % layout of in-line code snippets
-%\newcommand{\quote}[1]{\textit{``#1''}} % a direct quote from a paper
 \newcommand{\pn}{\textsc} % how to render proper names like Shrdlu
@@
 %theorem environments
-\theoremstyle{remark} \newtheorem*{termin}{Definition} % terminology. Could use a
+%\theoremstyle{remark} \newtheorem*{termin}{Definition} % terminology. Could use a
 %better title fixme
@@
 \bibliographystyle{plainnat}
+\bibpunct{(}{)}{;}{a}{,}{,}
 
@@
 $\langle$\href{mailto:aleks.dimitrov@googlemail.com}{aleks.dimitrov@googlemail.com}$\rangle$}
-\title{From Questions to Queries -- Enhancing the Curt System by a Theory and Implementation of Embedded WH-Questions}
+\title{From Questions to Queries – Enhancing the Curt System by a Theory and Implementation of Embedded WH-Questions}
 
-\usepackage[pdftex,colorlinks,citecolor=magenta]{hyperref}
+\usepackage{fontspec}
+\usepackage[xetex,colorlinks,citecolor=magenta]{hyperref}
 \begin{document}
+
 \maketitle
@@
 of the input and \emph{informativity} denotes \emph{validity}. Thus, if
-\Disc, the discourse so far and $\phi$ a new input token (fully parsed
-sentence), then consistency means $\Disc \to \phi$ holds and informativity
-means $\Disc \nvDash \phi$.}
+\Disc, the discourse so far and $φ$ a new input token (fully parsed
+sentence), then consistency means $\Disc → \phi$ holds and informativity
+means $\Disc \nvDash$ φ.}
 
@@
 entail the information given in its newest token and is thus irrelevant with
-regard to previous utterances -- effectively allowing it to
+regard to previous utterances – effectively allowing it to
 reduce the size of its internal representation of the discourse by not adding
@@
 
-\begin{termin}
+\begin{quote}
   An \term{interrogative} is a certain type of sentence in natural language
@@
   \emph{interrogative act}, whichever form it may have.
-\end{termin}
+\end{quote}
 
@@
 \abbr{IL} terms and is therefore avoided here. In fact, \abbr{IL} will be only
-marginally relevant for the presentation of \acurt -- only the type system
+marginally relevant for the presentation of \acurt – only the type system
 presented in \cite{ptq} is carried over.
@@
 implementation and the use of external tools only capable of treating first
-order logic -- an extensional framework. \prol on the other hand, is a language
+order logic – an extensional framework. \prol on the other hand, is a language
 that can very well make use of higher order formalisms for
@@
 
-\subsubsection{Classification of Interrogatives}
+\subsubsection{Pragmatics and Semantics of Questions}
+
+In choosing to treat only embedded questions, we can escape most pragmatic
+considerations and focus on the (implementation of) pure semantics of
+interrogatives.
+
+\subsection{Classification of Interrogatives}
 
@@
 
-\subsection{An Ontology of Embedding Verbs}
+\subsubsection{An Ontology of Embedding Verbs}
 
@@
 
-\subsection{Alternative Questions -- Inquisitive Semantics}\label{sec:altq}
+\subsection{Alternative Questions – Inquisitive Semantics}\label{sec:altq}
 

### 2008-08-18 src/englishGrammar.pl
Preliminary working state for Grammar.
Only lambda for now. I still get un-reduced terms, let's see how to fix this.

@@ -102,6 +102,28 @@ q([sem:Sem])-->
    sinv([gap:[np:NP],sem:S]),
    {combine(q:Sem,[sinv:S])}.
 
+/*
+ * Embedded Questions
+ */
+
+%todo: really coord:no for s? Bill knows that either Jane loves Mary or …
+tbar([sem:TBar])
+-->	whemb([sem:WHemb])
+	, s([coord:no,sem:S])
+	, { combine(tbar:TBar,[whemb:WHemb,s:S]) }
+.
+
+% for empty complementizers
+tbar([sem:TBar])
+-->	s([coord:no,sem:S])
+	, { combine(tbar:TBar,[s:S]) }
+.
+
+% for 'mia knows who likes vincent'
+tbar([sem:TBar])
+-->	q([sem:Q])
+	, { combine(tbar:TBar,[q:Q]) }
+.
 
 /*========================================================================
    Noun Phrases
@@ -191,6 +213,12 @@ nmod([sem:Sem])-->
    Verb Phrases
 ========================================================================*/
 
+vp([coord:no,inf:Inf,num:Num,gap:[],sem:VP])
+-->	ivtbar([inf:Inf,num:Num,sem:IVTBAR]) % verb
+	, tbar([sem:TBAR]) % embedded sentence
+	, {combine(vp:VP,[ivtbar:IVTBAR,tbar:TBAR])}
+.
+
 vp([coord:yes,inf:Inf,num:Num,gap:[],sem:VP])--> 
    vp([coord:no,inf:Inf,num:Num,gap:[],sem:VP1]), 
    coord([type:_,sem:C]), 
@@ -246,6 +274,12 @@ iv([inf:Inf,num:Num,sem:Sem])-->
    Word,
    {semLex(iv,[symbol:Sym,sem:Sem])}.
 
+ivtbar([inf:Inf,num:Num,sem:Sem])
+-->	{lexEntry(ivtbar,[symbol:Sym,syntax:Word,inf:Inf,num:Num])}
+	, Word
+	, {semLex(ivtbar,[symbol:Sym,sem:Sem])}
+.
+
 tv([inf:Inf,num:Num,sem:Sem])--> 
    {lexEntry(tv,[symbol:Sym,syntax:Word,inf:Inf,num:Num])},
    Word,
@@ -263,13 +297,25 @@ det([mood:M,type:Type,sem:Det])-->
 
 pn([sem:Sem])--> 
[…]

### 2008-08-18 src/englishLexicon.pl

@@ -21,6 +21,22 @@
 
 *************************************************************************/
 
+/*
+ * wh-embedding verbs
+ * cf karttunen 77 for the complete classification according to 'type'
+ */
+lexEntry(ivtbar,[symbol:knowi,syntax:[know],inf:inf,num:sg]).
+lexEntry(ivtbar,[symbol:knowi,syntax:[knows],inf:fin,num:sg]).
+lexEntry(ivtbar,[symbol:knowi,syntax:[know],inf:fin,num:pl]).
+
+lexEntry(ivtbar,[symbol:believei,syntax:[believe],inf:inf,num:sg]).
+lexEntry(ivtbar,[symbol:believei,syntax:[believes],inf:fin,num:sg]).
+lexEntry(ivtbar,[symbol:believei,syntax:[believe],inf:fin,num:pl]).
+
+lexEntry(ivtbar,[symbol:thinki,syntax:[think],inf:inf,num:sg]).
+lexEntry(ivtbar,[symbol:thinki,syntax:[thinks],inf:fin,num:sg]).
+lexEntry(ivtbar,[symbol:thinki,syntax:[think],inf:fin,num:pl]).
+
 /*========================================================================
    Determiners
 ========================================================================*/
@@ -205,7 +221,7 @@ lexEntry(tv,[symbol:pickup,syntax:[picks,up],inf:fin,num:sg]).
 lexEntry(tv,[symbol:pickup,syntax:[pick,up],inf:fin,num:pl]).
 
 lexEntry(tv,[symbol:shoot,syntax:[shot],inf:inf,num:sg]).
-lexEntry(tv,[symbol:shoot,syntax:[shot],inf:fin,num:sg]).
+lexEntry(tv,[symbol:shoot,syntax:[shot],inf:fin,num:sg]). % interesting. they allow past tense :-)
 lexEntry(tv,[symbol:shoot,syntax:[shoots],inf:fin,num:sg]).
 lexEntry(tv,[symbol:shoot,syntax:[shoot],inf:fin,num:pl]).
 
@@ -254,6 +270,12 @@ lexEntry(adj,[symbol:tall,syntax:[tall]]).
 lexEntry(relpro,[syntax:[who]]).
 lexEntry(relpro,[syntax:[that]]).
 
+/*
+ * wh-embedders
+ */
+lexEntry(whemb,[syntax:[that],type:assert]).
+lexEntry(whemb,[syntax:[whether],type:altern]).
+lexEntry(whemb,[syntax:[if],type:altern]).
 
[…]

### 2008-08-18 src/semLexLambda.pl

@@ -52,6 +52,21 @@ semLex(tv,M):-
         sem:lam(K,lam(Y,app(K,lam(X,Formula))))], 
    compose(Formula,Sym,[Y,X]).
 
+semLex(ivtbar,M)
+:-	M = [symbol:Sym,sem:lam(S,lam(X,app(S,lam(Y,Formula))))]
+	, compose(Formula,Sym,[X,Y])
+.
+
+% a simple unit rule for assertive embedders
+semLex(whemb,M)
+:-	M = [sem:lam(S,S),type:assert]
+.
+
+% a more complex rule for yes/no embedders
+semLex(whemb,M)
+:-	M = [sem:lam(S,or(S,not(S))),type:altern]
+.
+
 semLex(qnp,M):-
    M = [type:wh,
         symbol:Sym,

### 2008-08-18 src/semRulesLambda.pl

@@ -31,6 +31,9 @@ combine(t:Converted,[s:Sem]):-
 combine(t:Converted,[q:Sem]):- 
    betaConvert(Sem,Converted).
 
+combine(tbar:Converted,[s:Sem]):- 
+   betaConvert(Sem,Converted).
+
 combine(s:app(A,B),[np:A,vp:B]).
 combine(s:app(A,B),[s:A,s:B]).
 combine(s:lam(B,imp(S,B)),[if:S]).
@@ -65,6 +68,20 @@ combine(vp:app(A,B),[av:A,vp:B]).
 combine(vp:app(A,B),[cop:A,np:B]).
 combine(vp:A,[iv:A]).
 combine(vp:app(A,B),[tv:A,np:B]).
+
+% wh-embedding verbs need to beta-reduce the sentence they embed first.
+combine(vp:app(A,B),[ivtbar:A,tbar:B]).
+
+combine(tbar:app(WH,SRed),[whemb:WH,s:S])
+:-	betaConvert(S,SRed)
+.
+
+combine(tbar:SRed,[s:S])
+:-	betaConvert(S,SRed)
+.
+combine(tbar:QRed,[q:Q])
+:-	betaConvert(Q,QRed)
+.
 
 combine(pp:app(A,B),[prep:A,np:B]).
 

### 2008-08-19 src/englishGrammar.pl
Removed a few extraneous comments and predicates.

@@ -305,17 +305,14 @@ relpro([sem:Sem])-->
    Word,
    {semLex(relpro,[sem:Sem])}.
 
-/*
- * WH-embedding
- * We don't use symbols because wh-embedders are translated to
- * logic right away (using the respective storage method)
- */
+% WH-embedding
+% We don't use symbols because wh-embedders are translated to
+% logic right away (using the respective storage method)
 whemb([sem:Sem])
 -->	{lexEntry(whemb,[syntax:Word,type:T])}
 	, Word
 	, {semLex(whemb,[sem:Sem,type:T])}
 .
-
 
 prep([sem:Sem])--> 
    {lexEntry(prep,[symbol:Sym,syntax:Word])},

### 2008-08-19 src/semRulesLambda.pl

@@ -31,9 +31,6 @@ combine(t:Converted,[s:Sem]):-
 combine(t:Converted,[q:Sem]):- 
    betaConvert(Sem,Converted).
 
-combine(tbar:Converted,[s:Sem]):- 
-   betaConvert(Sem,Converted).
-
 combine(s:app(A,B),[np:A,vp:B]).
 combine(s:app(A,B),[s:A,s:B]).
 combine(s:lam(B,imp(S,B)),[if:S]).
@@ -69,20 +66,17 @@ combine(vp:app(A,B),[cop:A,np:B]).
 combine(vp:A,[iv:A]).
 combine(vp:app(A,B),[tv:A,np:B]).
 
-% wh-embedding verbs need to beta-reduce the sentence they embed first.
 combine(vp:app(A,B),[ivtbar:A,tbar:B]).
 
 combine(tbar:app(WH,SRed),[whemb:WH,s:S])
 :-	betaConvert(S,SRed)
 .
-
 combine(tbar:SRed,[s:S])
 :-	betaConvert(S,SRed)
 .
 combine(tbar:QRed,[q:Q])
 :-	betaConvert(Q,QRed)
 .
-
 combine(pp:app(A,B),[prep:A,np:B]).
 
 combine(rc:app(A,B),[relpro:A,vp:B]).

### 2008-08-19 src/semLexLambda.pl
Fixed that damn application bug.

@@ -53,7 +53,7 @@ semLex(tv,M):-
    compose(Formula,Sym,[Y,X]).
 
 semLex(ivtbar,M)
-:-	M = [symbol:Sym,sem:lam(S,lam(X,app(S,lam(Y,Formula))))]
+:-	M = [symbol:Sym,sem:lam(S,lam(X,app(lam(Y,Formula),S)))]
 	, compose(Formula,Sym,[X,Y])
 .
 

### 2008-08-19 src/semLexStorage.pl
Ported over to KellerStorage.

@@ -56,6 +56,21 @@ semLex(tv,M):-
         sem:[lam(K,lam(Y,app(K,lam(X,Formula))))]], 
    compose(Formula,Sym,[Y,X]).
 
+semLex(ivtbar,M)
+:-	M = [symbol:Sym,sem:[lam(S,lam(X,app(lam(Y,Formula),S)))]]
+	, compose(Formula,Sym,[X,Y])
+.
+
+% a simple unit rule for assertive embedders
+semLex(whemb,M)
+:-	M = [sem:[lam(S,S),type:assert]]
+.
+
+% a more complex rule for yes/no embedders
+semLex(whemb,M)
+:-	M = [sem:[lam(S,or(S,not(S))),type:altern]]
+.
+
 semLex(qnp,M):-
    M = [type:wh,
         symbol:Sym,

### 2008-08-19 src/semRulesKeller.pl

@@ -54,6 +54,19 @@ combine(q:Q,[whnp:[A|S1],vp:[B|S2]]):-
 combine(q:Q,[sinv:Q]):-
    Q=que(_,_,_).
 
+
+combine(tbar:[app(WH,SRed)|S1],[whemb:[WH],s:[S|S1]])
+:-	betaConvert(S,SRed)
+.
+
+combine(tbar:[SRed],[s:S])
+:-	betaConvert(S,SRed)
+.
+
+combine(tbar:[QRed],[q:Q])
+:-	betaConvert(Q,QRed)
+.
+
 combine(np:A,[pn:A]).
 combine(np:A,[qnp:A]).
 combine(np:[lam(P,app(P,X)),bo([app(A,B)|S],X)],[det:[A],n:[B|S]]).
@@ -81,6 +94,8 @@ combine(vp:A,[iv:A]).
 combine(vp:[app(A,B)|S],[tv:[A],np:[B|S]]).
 combine(vp:[app(app(B,A),C)|S3],[vp:[A|S1],coord:[B],vp:[C|S2]]):-
 	appendLists(S1,S2,S3).
+
+combine(vp:[app(A,B)|S],[ivtbar:[A],tbar:[B|S]]).
 
 combine(pp:[app(A,B)|S],[prep:[A],np:[B|S]]).
 

### 2008-08-19 src/englishGrammar.pl
Added proper treatment of assertive verbs.
... using a type system. This will have to be integrated with whemb's types

@@ -275,9 +275,9 @@ iv([inf:Inf,num:Num,sem:Sem])-->
    {semLex(iv,[symbol:Sym,sem:Sem])}.
 
 ivtbar([inf:Inf,num:Num,sem:Sem])
--->	{lexEntry(ivtbar,[symbol:Sym,syntax:Word,inf:Inf,num:Num])}
+-->	{lexEntry(ivtbar,[symbol:Sym,syntax:Word,inf:Inf,num:Num,type:Type])}
 	, Word
-	, {semLex(ivtbar,[symbol:Sym,sem:Sem])}
+	, {semLex(ivtbar,[symbol:Sym,sem:Sem,type:Type])}
 .
 
 tv([inf:Inf,num:Num,sem:Sem])--> 

### 2008-08-19 src/englishLexicon.pl

@@ -25,17 +25,17 @@
  * wh-embedding verbs
  * cf karttunen 77 for the complete classification according to 'type'
  */
-lexEntry(ivtbar,[symbol:knowi,syntax:[know],inf:inf,num:sg]).
-lexEntry(ivtbar,[symbol:knowi,syntax:[knows],inf:fin,num:sg]).
-lexEntry(ivtbar,[symbol:knowi,syntax:[know],inf:fin,num:pl]).
+lexEntry(ivtbar,[symbol:knowi,syntax:[know],inf:inf,num:sg,type:assert]).
+lexEntry(ivtbar,[symbol:knowi,syntax:[knows],inf:fin,num:sg,type:assert]).
+lexEntry(ivtbar,[symbol:knowi,syntax:[know],inf:fin,num:pl,type:assert]).
 
-lexEntry(ivtbar,[symbol:believei,syntax:[believe],inf:inf,num:sg]).
-lexEntry(ivtbar,[symbol:believei,syntax:[believes],inf:fin,num:sg]).
-lexEntry(ivtbar,[symbol:believei,syntax:[believe],inf:fin,num:pl]).
+lexEntry(ivtbar,[symbol:believei,syntax:[believe],inf:inf,num:sg,type:stipul]).
+lexEntry(ivtbar,[symbol:believei,syntax:[believes],inf:fin,num:sg,type:stipul]).
+lexEntry(ivtbar,[symbol:believei,syntax:[believe],inf:fin,num:pl,type:stipul]).
 
-lexEntry(ivtbar,[symbol:thinki,syntax:[think],inf:inf,num:sg]).
-lexEntry(ivtbar,[symbol:thinki,syntax:[thinks],inf:fin,num:sg]).
-lexEntry(ivtbar,[symbol:thinki,syntax:[think],inf:fin,num:pl]).
+lexEntry(ivtbar,[symbol:thinki,syntax:[think],inf:inf,num:sg,type:stipul]).
+lexEntry(ivtbar,[symbol:thinki,syntax:[thinks],inf:fin,num:sg,type:stipul]).
+lexEntry(ivtbar,[symbol:thinki,syntax:[think],inf:fin,num:pl,type:stipul]).
 
 /*========================================================================
    Determiners

### 2008-08-19 src/semLexLambda.pl

@@ -53,7 +53,12 @@ semLex(tv,M):-
    compose(Formula,Sym,[Y,X]).
 
 semLex(ivtbar,M)
-:-	M = [symbol:Sym,sem:lam(S,lam(X,app(lam(Y,Formula),S)))]
+:-	M = [symbol:Sym,sem:lam(S,lam(X,app(lam(Y,and(Formula,S)),S))),type:assert]
+	, compose(Formula,Sym,[X,Y])
+.
+
+semLex(ivtbar,M)
+:-	M = [symbol:Sym,sem:lam(S,lam(X,app(lam(Y,Formula),S))),type:_]
 	, compose(Formula,Sym,[X,Y])
 .
 

### 2008-08-19 src/semLexStorage.pl

@@ -57,7 +57,12 @@ semLex(tv,M):-
    compose(Formula,Sym,[Y,X]).
 
 semLex(ivtbar,M)
-:-	M = [symbol:Sym,sem:[lam(S,lam(X,app(lam(Y,Formula),S)))]]
+:-	M = [symbol:Sym,sem:[lam(S,lam(X,app(lam(Y,and(S,Formula)),S)))],type:assert]
+	, !
+	, compose(Formula,Sym,[X,Y])
+.
+semLex(ivtbar,M)
+:-	M = [symbol:Sym,sem:[lam(S,lam(X,app(lam(Y,Formula),S)))],type:_T]
 	, compose(Formula,Sym,[X,Y])
 .
 

### 2008-08-20 src/englishLexicon.pl
experimental treatment of uniform knowledge

@@ -25,17 +25,17 @@
  * wh-embedding verbs
  * cf karttunen 77 for the complete classification according to 'type'
  */
-lexEntry(ivtbar,[symbol:knowi,syntax:[know],inf:inf,num:sg,type:assert]).
-lexEntry(ivtbar,[symbol:knowi,syntax:[knows],inf:fin,num:sg,type:assert]).
-lexEntry(ivtbar,[symbol:knowi,syntax:[know],inf:fin,num:pl,type:assert]).
+lexEntry(ivtbar,[symbol:knowledge,syntax:[know],inf:inf,num:sg,type:assert]).
+lexEntry(ivtbar,[symbol:knowledge,syntax:[knows],inf:fin,num:sg,type:assert]).
+lexEntry(ivtbar,[symbol:knowledge,syntax:[know],inf:fin,num:pl,type:assert]).
 
-lexEntry(ivtbar,[symbol:believei,syntax:[believe],inf:inf,num:sg,type:stipul]).
-lexEntry(ivtbar,[symbol:believei,syntax:[believes],inf:fin,num:sg,type:stipul]).
-lexEntry(ivtbar,[symbol:believei,syntax:[believe],inf:fin,num:pl,type:stipul]).
+lexEntry(ivtbar,[symbol:knowledge,syntax:[believe],inf:inf,num:sg,type:stipul]).
+lexEntry(ivtbar,[symbol:knowledge,syntax:[believes],inf:fin,num:sg,type:stipul]).
+lexEntry(ivtbar,[symbol:knowledge,syntax:[believe],inf:fin,num:pl,type:stipul]).
 
-lexEntry(ivtbar,[symbol:thinki,syntax:[think],inf:inf,num:sg,type:stipul]).
-lexEntry(ivtbar,[symbol:thinki,syntax:[thinks],inf:fin,num:sg,type:stipul]).
-lexEntry(ivtbar,[symbol:thinki,syntax:[think],inf:fin,num:pl,type:stipul]).
+lexEntry(ivtbar,[symbol:knowledge,syntax:[think],inf:inf,num:sg,type:stipul]).
+lexEntry(ivtbar,[symbol:knowledge,syntax:[thinks],inf:fin,num:sg,type:stipul]).
+lexEntry(ivtbar,[symbol:knowledge,syntax:[think],inf:fin,num:pl,type:stipul]).
 
 /*========================================================================
    Determiners

### 2008-08-20 src/modelChecker1.pl
just removed some spaces :-)

@@ -56,7 +56,6 @@ evaluate(Formula,Example,Assignment):-
    example(Example,Model),
    satisfy(Formula,Model,Assignment,Result),
    printStatus(Result).
-   
 
 /*========================================================================
    Test Suite

### 2008-08-20 thesis/paper.tex
more work

@@
-\documentclass[a4paper]{article}
+\documentclass[12pt,a4paper]{article}
 
-\usepackage[english]{babel}
+%\usepackage[english]{babel}
 
-\usepackage{amssymb,bbm,qtree,graphicx,natbib,aleks,linguex}
+\usepackage{amssymb,qtree,natbib,aleks,linguex}
+
+\usepackage{amsthm}
 
 %% note that this file depends on being processed with XeTeX
-\usepackage{xltxtra,xunicode,fontspec}
+\usepackage{fontspec}
+\usepackage{xltxtra,xunicode}
 
+\usepackage{mathpazo}
 % font specifications (don't overuse ;-)
 \defaultfontfeatures{Mapping=tex-text} % converts LaTeX specials (``quotes'' --- dashes etc.) to unicode
-\setromanfont [Ligatures={Common,Rare}]{Gentium}
+\setromanfont [Ligatures={Common,Rare}, BoldFont={Fontin Bold}, ItalicFont={Fontin Italic}]{Fontin Regular}
+\setsansfont [Ligatures={Common}, BoldFont={Fontin Sans Bold}, ItalicFont={Fontin Sans Italic}]{Fontin Sans}
 
@@
 \newcommand{\code}{\texttt} % layout of in-line code snippets
-\newcommand{\pn}{\textsc} % how to render proper names like Shrdlu
+\newcommand{\pn}{\textsf} % how to render proper names like Shrdlu
 \newcommand{\url}[1]{\code{http://#1}} % how urls appear
+\newcommand{\example}{\textit} % in-text examples
 
@@
 %theorem environments
-%\theoremstyle{remark} \newtheorem*{termin}{Definition} % terminology. Could use a
+\theoremstyle{remark} \newtheorem*{termin}{Definition} % terminology. Could use a
 %better title fixme
@@
 of the input and \emph{informativity} denotes \emph{validity}. Thus, if
-\Disc, the discourse so far and $φ$ a new input token (fully parsed
-sentence), then consistency means $\Disc → \phi$ holds and informativity
-means $\Disc \nvDash$ φ.}
+\Disc, the discourse so far and $\phi$ a new input token (fully parsed
+sentence), then consistency means $\Disc \to \phi$ holds and informativity
+means $\Disc \nvDash \phi$.}
 
@@
 
+This section first gives an overview over the terminology that will be used for
+the discussion of the formal treatment of questions. It then elaborates on the
+differences between different logical formalisms employed in the treatment of
+natural language questions. After a small overview on the types of questions
+(and pointing out the subset of questions that will be treated), %fixme maybe a
+% bit sloppy
+the discussion
+turns to how those questions are best treated in the given formalism.
+
+The reader should note that all choices presented in this section have been made
+with the implementation of section \label{sec:implementation} in mind.
+% todo talk a bit more
+
 \subsection{Terminology}
@@
 
-\begin{quote}
+\begin{termin}
   An \term{interrogative} is a certain type of sentence in natural language
@@
   \emph{interrogative act}, whichever form it may have.
-\end{quote}
+\end{termin}
 
@@
 (\url{http://molle.sf.net}), a modal theorem prover that comes with a graphical user
-interface that allows interactive investigation of the constructed indices.},
+interface which allows interactive investigation of the constructed indices.},
 a restructuring of the \pn{Curt} system to take advantage of them would be
@@
 to representing the indices of higher-order logic in \prol in section
-\ref{sec:indices}.
+\ref{sec:indices}. This logic however has to be augmented with question
+operators as already used by \cite{karttunen:1977}. % todo: writeme!
 
@@
 
-Question Semantics is a much-disputed but often overlooked field of modern
-semantic research. Often question semantics has been reduced to an issue of
-pragmatics, treating direct questions (interrogatives) as nothing more than a
-paraphrase of indirect questions (embedded interrogatives).\cite{tichy}
-even goes as far as claiming equality in semantic value between
+Question semantics are an interesting subproblem of modern research on natural
+language semantics. While traditional approaches based on \abbr{IL} and newer
+accounts also employ plain extensional logic, % todo: which ones 
+the treatment
+of questions seems to demand that we introduce a new kind of logic formalism.
+But as interesting topics go, the semantics of questions has attracted many
+different views on this issue.
+
+Often question semantics has been reduced to a problem of
+pragmatics, treating direct interrogatives as nothing more than a
+paraphrase of embedded interrogatives.\cite{tichy}
+even goes as far as claiming equality in semantic (truth) value between
 indicatives and interrogatives. The benefit of such an approach would be the
@@
 pragmatics.
-This approach has been challenged and proven wrong by \cite{gs:q},
-who also argue that the proper treatment of questions would require extensions
-to the existing logic frameworks used in natural language semantics semantics.
+
+This approach has been challenged and proven unfeasible by \cite{gs:q},
+who have also proposed interesting extensions
+to the existing logic frameworks used in natural language semantics.
 More recent publications, such as \cite{gal:tmbi}
@@
 
-Nevertheless, progress has been very lively since the first
-formal approaches to the field started with \cite{hamblin:q} and other
-publications such as % todo
-. \cite{gs:q} provide a very descriptive overview of the field of question
-semantics
+The present work will rely on \cite{gs:q} for providing the main guidelines for
+the theoretic aspects and also lend from earlier works, most importantly
+\cite{gs:sqpa}\footnotemark. Later on, in section \ref{sec:altq} we will also
+discuss the very newest (at the time of writing) developments in question
+semantics, depicted in \cite{g:is}.
+
+\footnotetext{\cite{gs:sqpa} is actually a series of three articles
+that have appeared before. Among them, we will mostly consider \cite{gs:sawhq}.
+Both, the earlier paper and the later dissertation (and many other papers
+originating in Amsterdam) have been kindly made
+available electronically, free of charge, by the Universiteit van Amsterdam at
+the following address: \url{http://dare.uva.nl}}
 
@@
 
-Our aim is to improve coverage of grammatical structures and thus when faced
-with the necessity to make compromise about the complexity of our grammar, we
-will choose overgeneration in favor of unmotivated complexity of the grammar
-itself.
+The present work will not consider a full analysis of \wh-embedding verbs, but
+confine itself on a few ones, namely \example{know}, \example{think} and
+\example{believe}. The grammar implementation of \acurt allows for \wh-embedding
+verbs to have a \code{type} field, determining their semantic denotation. The two
+types used so far are \code{assert} for assertive verbs such as \example{know}
+and \code{stipul} for stipulating verbs, such as \example{believe} and
+\example{think}.
+
+\subsubsection{Grammatic Constraints}
+
+With this ontology in mind, the next step is to analyse and enumerate the sorts
+of grammatic constructions that are to be treated. The following examples will
+illustrate them.
+
+\begin{itemize}
+  \item Alternative Questions
+  \ex. \a. Mia \emph{knows} whether Vincent snorts.
+  \b. * Mia \emph{believes} whether Vincent snorts.
+  \b. * Mia \emph{thinks} whether Vincent snorts.
+
+  \item Enumerative Questions
+  \ex. \a.  Mia \emph{knows} who snorts.
+  \b. * Mia \emph{believes} who snorts.
+  \b. * Mia \emph{thinks} who snorts.
+
+  \item Embedding Propositions
+  \ex. \label{ex:stipul}
+  \a. Mia \emph{believes} Vincent snorts.
+  \b. * Mia \emph{believes} who snorts.
+
+\end{itemize}
+
+An important observation to be made here is that only \emph{assertive} verbs can
+embed real questions. \emph{Stipulating} verbs only embed \emph{propositions} as
+in \ref{ex:stipul}.
+Nevertheless, both are treated in the following analysis, but with a focus on
+the treatment of assertive \wh-embedding verbs.
 
@@
 
-\subsubsection{Enhancing Storage}
+The underlying theory for
+
+\subsubsection{Defining the Grammar Rules}
 
@@
 to refer to this particular implementation as presented in
-\cite{blackburnbos:cl1}. Although \pn{Keller storage} is already quite capable
-of dealing with more complex constituents such as nested noun phrases, it needs
-some enhancements in order to be able to treat verbs embedding whole
-sentences or \wh-phrases correctly. The aim is to enable \curt's implementation of
-\pn{Keller storage} to deal with examples such as \ref{ex:kel1}\footnotemark.
+\cite{blackburnbos:cl1}.
+
+\pn{Keller storage} is already quite capable
+of dealing with more complex constituents such as nested noun phrases, but it
+has its limitations. In particular, it can only handle quantifier scope
+ambiguities and does not take negation into account. Moreover, it overgenerates
+with some quantifier scope ambiguities\footnote{\cite{blackburnbos:cl1} mention
+sentences such as \ref{ex:kellersucks}, where Keller storage produces six
+instead of the desired five readings.
+\ex. One criminal knows every owner of a hash bar\label{ex:kellersucks}
+
+Clearly, \example{a hash bar} cannot take scope over \example{every owner}, but
+Keller storage predicts this reading.}.
+
+In order to cope with more complicated quantifier configurations, where
+constraints need to be enforced, and with other kinds of ambiguities,
+Blackburn and Bos present an approach utilizing underspecified representations and
+dominance relations, \pn{Hole Semantics}\footnote{Also presented in % todo:cite
+%Bos' paper
+}. Since, however, none of the cases discussed so far would require the enhanced
+expressive power of \pn{Hole Semantics} it was decided to go on using \pn{Keller
+Storage} instead, since this is what the original implementation of the \curt
+system relies on.
+%todo: fix capitalization in Keller storage all over the document
+
+Enhancing the grammatic rules of the \curt system to cope with
+proposition-embedding verbs is made trivial by its elaborate grammatical back
+end. \ref{ex:kel1} is an example of the logic behind the grammar that is put to
+use.
 
 \ex. \label{ex:kel1} Mia knows whether Vincent snorts.\\
-\Tree [.t [.T Mia ] [.IV [.{IV/$\bar{t}$} knows ] [.{$\bar{t}$} whether [.t [.T
-Vincent ] [.IV snorts ] ] ] ] ]
+\Tree
+[.$t$ [.$T$ { $\lambda X.X(Mia)$\\Mia } ]
+[.$IV$ [.{$IV/\bar{t}$} { $\lambda X.know(X)$\\knows } ]
+[.{$\bar{t}$} [.$\bar{t}/t$ whether ] [.{ $t$\\$snort(Vincent)$ } 
+[.$T$ { $\lambda X.X(Vincent)$\\Vincent } ]
+[.$IV$ { $\lambda X.snort(X)$\\snorts } ] ] ] ] ]
 

### 2008-08-21 src/intensionalExampleModels.pl
Wow. First steps in intensionality complete.
We have a model framework and the like.
But validation is so far only extensional.

@@ -1,0 +1,29 @@
+:- module(intensionalExampleModels,[example/2]).
+
+
+/*========================================================================
+   Example Models
+
+   model(Individuals,Worlds,AccessibleRelations).
+========================================================================*/
+
+example(1,model([d1,d2,d3,d4,d5]
+	, [ (w0,world(
+		[ f(0,mia,d1)
+		, f(0,vincent,d2)
+		, f(0,pumpkin,d3)
+		, f(0,honey_bunny,d4)
+		, f(0,yolanda,d5)
+		, f(1,snort,[d1])
+		, f(1,customer,[d1,d2])
+		, f(1,robber,[d3,d4])
+		, f(2,love,[(d3,d4)])
+		]))
+	  , (w1,world(
+		[ f(1,snort,[d2])
+		, f(2,love,[(d2,d5)])
+		]))
+	  ]
+	, [(d1,w2)]
+	))
+	.

### 2008-08-21 src/intensionalModels.pl

@@ -1,8 +1,441 @@
-% Implementation of intensional models.
 
-module (intensionalModels).
+/*************************************************************************
 
-% unit tests
+    File: modelChecker2.pl
+    Copyright (C) 2004,2005,2006 Patrick Blackburn & Johan Bos
 
-% give a list of models given Ty2 expression containing indices
+    This file is part of BB1, version 1.3 (November 2006).
+
+    BB1 is free software; you can redistribute it and/or modify
+    it under the terms of the GNU General Public License as published by
+    the Free Software Foundation; either version 2 of the License, or
+    (at your option) any later version.
+
+    BB1 is distributed in the hope that it will be useful,
+    but WITHOUT ANY WARRANTY; without even the implied warranty of
+    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+    GNU General Public License for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with BB1; if not, write to the Free Software Foundation, Inc., 
+    59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
+
+*************************************************************************/
+
+:- module(modelChecker2,[modelCheckerTestSuite/0,
+                         info/0,
+                         infix/0,
+                         prefix/0,
+			 evaluate/2,
+			 evaluate/3,
+			 satisfy/4]).
+
+:- use_module(comsemPredicates,[memberList/2,
+                                compose/3, 
+                                proveOnce/1,
+                                infix/0,
+                                prefix/0,
+                                printRepresentations/1]).
+
+:- use_module(intensionalExampleModels,[example/2]).
+
+:- use_module(modelCheckerTestSuite,[test/4]).
+
+
+/*========================================================================
+   Evaluate a formula in an example model
+========================================================================*/
+
+evaluate(Formula,Example):-
+   evaluate(Formula,Example,[]).
+
+
[…]

### 2008-08-22 thesis/paper.tex
more work.

@@
 
+Growing efforts have thus been made to provide a technology that is (going to
+be) able to extract logical formulae from sentences of ordinary language and
+establish relations between objects of its logical representation. In addition,
+providing a means to compute with said representations is will eventually enable
+the machine-processing of meaning.  In doing so, the ultimate goal may be to
+enable a man-made system to \emph{understand} what a humans communicate.
+
+Exploration of this  interesting field of linguistics has shown that
+the notion of \emph{understanding} is far more than a linguistic concern, but
+involves at least findings in the domain of logics, (cognitive) psychology and
+probably even philosophy.
+
+Development of exhaustive systems in this area appears to be a daunting task of
+quite some scope. Therefore most approaches have been centered on providing
+coverage of at least a fragment of natural language. Many such systems exist and
+they prove to be increasingly successful. Some of them will be discussed here
+and presenting an extension to one system in particular is the main matter of
+this Bachelor's thesis.
+
+In doing so, the present work will aim to provide the \curt system of
+\cite{blackburnbos:cl1} with the mechanics of dealing with more than one
+partition of its logical domain, viz. the ability to handle more than one
+\term{possible} or \term{accessible world}. As a ground for development of this
+work, the treatment of \term{embedded questions} was chosen as one that
+naturally demands a possibility to account for more than one division in a
+system's \emph{understanding} or representation of a universe of objects and
+relations.
+
+As
+
+The first section gives a conceptual overview, introducing the \curt
+system and comparing it to other similar systems in existence. It then goes on
+to explain the target domain of natural language the implementation has to
+account for.
+The next section concerns itself with a more detailed analysis of the theoretic
+assumptions of this implementation and justifies some choices and restrictions
+that had to be made because of the limited scope of this analysis. While the
+third section will display and clarify the actual implementation on top of the
+\curt system, the last section will discuss the results of this work and also
+give a prospect on possible further work on the subject.
+
+\subsection*{Acknowledgements}
+
+Frank... who else?
+
+\section{Concepts}
+
+The basis of
+
 %fixme: to be formalized
@@
 mostly the purely semantic/logic domain, so this should be feasible to achieve.
-\subsection*{Acknowledgements}
 
-\section{Concepts}
-
-% fixme: formalize
 A brief overview of the premisses for the theory (which itself lays ground to
@@
 \pn{Curt}, a system written in \prol that is able to interpret basic utterances
-in natural language and store them in expressions of first order logic, is presented
-in \cite{blackburnbos:cl1}. While relatively limited when compared to other
-systems presented in \ref{sec:comparision} in
-terms of lexical and grammatical coverage it is able to perform basic checks for
-\emph{consistency} and \emph{informativity} of its input\footnotemark, within
-the bounds of the tools it uses.
+in natural language and store them in expressions of first order logic, is
+presented in \cite[chapter 6]{blackburnbos:cl1}. While relatively limited when
+compared to other systems presented in \ref{sec:comparision} in terms of lexical
+and grammatical coverage, it provides a sound basis for extended work because of
+its coherent design and relative simplicity.
+
+Besides parsing and translating utterances into logic, \curt is able to perform
+basic checks for \emph{consistency} and \emph{informativity} on the resulting
+formulae -- within the bounds of the tools it uses.
 
@@
 possible approach takes highest precedence.
+
+To quote \cite{blackburnbos:cl1}:
+
+\begin{quote}
+  Questions, unlike assertions, don't have truth-values, so it would be
+  misleading to represent them as ordinary formulas.
+\end{quote}
+
+It is important to note that this claim is not undisputed and
+\cite{karttunen:1977}, argues against it. The present approach, however will
+base itself on this assumption and extend it by more formal approaches than the
+one given by Blackburn \& Bos. Specifically, the work by J. Groenendijk and M. Stokhof
+will lay ground to the implementation. But in order to realise their formal
+discussions computationally, modifications will have to be made, and even
+compromises.
 

### 2008-08-22 thesis/paper.tex
added section about disjunction

@@
 
+\subsubsection{Representing Intensionality in \prol}
+
+\prol programs can be self-modifying, or dynamic. This means that the program is
+able to change its own state and thus store information.
+
+\subsubsection{Disjunction and Ambiguity}
+Consider the following sentences:
+
+\ex. \label{ex:disjunction} \a. Mia or Jody dances
+\b. Either Butch snores or Vincent is using the chainsaw
+
+\pn{Helpful Curt}, with its unpartitioned logical space and its lack of
+the awareness of more than one (possible) world, would generate only one model
+to represent the facts of \ref{ex:disjunction}. It is therefore only able to
+generate \emph{one} possible answer to both questions, effectively accounting only
+for one possible constituent of the disjunction, prefixing it with a `maybe' to
+indicate its uncertainty.
+
+Another example where \curt has to sacrifice exhaustiveness over its one-world
+representation of the world is given in \ref{ex:ambig}
+
+\ex.\label{ex:ambig} Every boxer likes a woman.
+
+This will generally generate two readings, but only one model -- with only one
+world. And further discourse will have to choose between one of the readings and
+stick with the minimal model constructed by the model builder. This is not
+really satisfactory and extending the dimensions of knowledge within \curt seems
+necessary in order to treat such examples.
+
+The previous work on \acurt enables it to generate a \term{choice point} -- and
+moreover, not pursue only one path of it, but both at the same time, by
+generating more than one world in its model.
+Unfortunately, this will build up very large models over time and computations
+with it might prove to be very complex and resource demanding. Adding such a
+functionality to the system, however, is not a hard task and hopefully the
+technique employed here can be further refined. %todo: maybe add a footnote
+%about future prospects or so
+
 \subsubsection{Defining the Grammar Rules}

### 2008-08-22 src/intensionalModels.pl
Changed info

@@ -421,7 +421,7 @@ answer(protoq(X,Formula),model(D,M),G,Answers) :-
 
 info:-
    format('~n> ------------------------------------------------------------------------- <',[]),
-   format('~n> modelChecker2.pl, by Patrick Blackburn and Johan Bos                      <',[]),
+   format('~n> intensionalModels.pl, by Aleksandar Dimitrov, based on modelChecker2.pl   <',[]),
    format('~n>                                                                           <',[]),
    format('~n> ?- evaluate(F,E).          - evaluate a formula in a model                <',[]),
    format('~n> ?- evaluate(F,E,A).        - evaluate a formula in a model wrt assignment <',[]),

### 2008-08-23 src/or.pl
Added or.pl to do disjunktion [sic]

@@ -1,0 +1,76 @@
+% or.pl
+% takes a formula and gives back a list of possible formulas
+% after eliminating disjunction
+% Aleksandar Dimitrov
+
+module(or,[disjunk/2]).
+
+term(Sym,Args,Term) :-
+	(
+		is_list(Args), !, Term =.. [Sym|Args]
+	;
+		Term =.. [Sym|[Args]]
+	)
+	.
+
+disjunk(X,[X]) :- (var(X); atom(X)) , ! .
+
+disjunk(or(A,B),Formulas) :-
+	disjunk(A,FA)
+	, disjunk(B,FB)
+	, !
+	, append(FA,FB,Formulas)
+	.
+
+disjunk(Formula,Result) :-
+	Formula =.. [Sym|Args]
+	, maplist(disjunk,Args,DArgs)
+	, args(DArgs,PDArgs)
+	, !
+	, maplist(term(Sym),PDArgs,Result)
+	.
+
+
+% args/2
+% takes a list of lists and returns all permutations of each list's elements
+args([],[[]]).
+
+args([X],X).
+
+args([X,Y],Perms) :- 
+	args(X,Y,Perms)
+	.
+
+args([Xs,Ys|Zs],Perms) :-
+	args(Xs,Ys,P)
+	, args([P|Zs],Perms)
+	.
+
+% args/3 
+% helper for args/2 - two lists (both of which may be nested)
+% and returs all permutations on both lists
+args([],_,[]).
+
+args(As,Bs,Lists) :- 
+	As = [X|Xs]
+	, args(Xs,Bs,More)
+	, maplist(pairs(X),Bs,New)
+	, append(New,More,Lists)
+	.
+
+pairs(X,Y,R) :-
+	(
+		(
+			is_list(X),
+			(
+				is_list(Y), append(X,Y,R)
+			;
+				append(X,[Y],R)
+			)
+		;
+			is_list(Y), append([X],Y,R)
+		)
+	;
+		R = [X,Y]
+	)
+	.

### 2008-08-23 src/advertentCurt.pl
Added helpfulCurt stuff to advertentCurt.

@@ -1,1 +1,336 @@
 %% Advertent Curt. Here to answer your questions.
+% Aleksandar Dimitrov
+% based on Helpful Curt
+
+:- module(curt,[curt/0,infix/0,prefix/0]).
+
+:- use_module(callInference,[callTP/3,
+                             callTPandMB/6]).
+
+:- use_module(readLine,[readLine/1]).
+
+:- use_module(comsemPredicates,[infix/0,
+                                prefix/0,
+                                memberList/2,
+                                compose/3,
+				selectFromList/3,
+				printRepresentations/1]).
+
+:- use_module(kellerStorage,[kellerStorage/2]).
+
+:- use_module(intensionalModels,[satisfy/4]).
+
+:- use_module(backgroundKnowledge,[backgroundKnowledge/2]).
+
+:- use_module(elimEquivReadings,[elimEquivReadings/2]).
+
+:- use_module(curtPredicates,[curtHelp/0,
+                              curtOutput/1,
+                              updateReadings/1,
+                              updateModels/1,
+                              updateHistory/1,
+                              clearHistory/0,
+                              list2string/2,
+                              selectReadings/3]).
+
+
+/*========================================================================
+   Dynamic Predicates
+========================================================================*/
+
+:- dynamic history/1, readings/1, models/1.
+
+history([]).
+readings([]).
+models([]).
+
+
+/*========================================================================
+   Start Curt
+========================================================================*/
+
+curt:- 
+   curtTalk(run).
+
+
+/*========================================================================
+   Control
+========================================================================*/
+
+curtTalk(quit).
+
+curtTalk(run):-
+   readLine(Input),
+   curtUpdate(Input,CurtsMoves,State), 
+   curtOutput(CurtsMoves),
+   curtTalk(State).
+
+
+/*========================================================================
+   Update Curt's Information State
+========================================================================*/
+
+curtUpdate([],[clarify],run):- !.
+
+curtUpdate([bye],[bye],quit):- !,
+   updateReadings([]),
+   updateModels([]),
+   clearHistory.
+
+curtUpdate([new],[],run):- !,
+   updateReadings([]),
+   updateModels([]),
+   clearHistory.
+
+curtUpdate([help],[],run):- !,
+   curtHelp.
+
[…]

### 2008-08-24 src/or.pl
Fixed bug in or.pl

@@ -3,7 +3,7 @@
 % after eliminating disjunction
 % Aleksandar Dimitrov
 
-module(or,[disjunk/2]).
+:- module(or,[disjunk/2]).
 
 term(Sym,Args,Term) :-
 	(

### 2008-08-24 src/advertentCurt.pl
Reworked some of acurt. added interpretation stubs

@@ -16,6 +16,8 @@
 				selectFromList/3,
 				printRepresentations/1]).
 
+:- use_module(or,[disjunk/2]).
+
 :- use_module(kellerStorage,[kellerStorage/2]).
 
 :- use_module(intensionalModels,[satisfy/4]).
@@ -107,10 +109,10 @@ curtUpdate([summary],[],run):-
    updateModels([]).
 
 curtUpdate([knowledge],[],run):-
-   readings(R), 
+   readings(R),
    findall(K,(memberList(F,R),backgroundKnowledge(F,K)),L),
    printRepresentations(L).
-   
+
 curtUpdate([readings],[],run):- !,
    readings(R),
    printRepresentations(R).
@@ -123,40 +125,31 @@ curtUpdate([history],[],run):- !,
    history(H),
    printRepresentations(H).
 
-curtUpdate(Input,Moves,run):-
-   kellerStorage(Input,Readings), !,
-   updateHistory(Input),
-   (
-      Readings=[que(X,R,S)|_],
-      models(OldModels),
-      answerQuestion(que(X,R,S),OldModels,Moves)
-   ;
-      \+ Readings=[que(_,_,_)|_],
-      consistentReadings(Readings,[]-ConsReadings,[]-Models),
-      (
-         ConsReadings=[],
-         Moves=[contradiction]
-      ;
-         \+ ConsReadings=[],
-         informativeReadings(ConsReadings,[]-InfReadings),
-         (
-            InfReadings=[],
-            Moves=[obvious]
-         ;
-            \+ InfReadings=[],
-            Moves=[accept]
-         ),
-         combine(ConsReadings,CombinedReadings), 
-         updateReadings(CombinedReadings),
-         updateModels(Models)
-      )
-   ).
+curtUpdate(Input,Moves,run) :-
+	kellerStorage(Input,Rs)
+	, !
+	, maplist(disjunk,Rs,Nested)
+	, flatten(Nested,Readings)
+	, updateHistory(Input)
+	,
+	% TODO: Write Answer Code
+	interpret(Readings,Model,Moves)
+	.
 
 curtUpdate(_,[noparse],run).
+
+%%% interpret/3
+% take some readings and generate appropriate models based on what's already
+% there. Also inform the rest of Curt about what we do next based on the result.
+
+interpret([],[],[noparse]).
+
+interpret(Readings,Model,Move).
 
 
 /*========================================================================
    Combine New Utterances with History
+   TODO: add a call to the summary command
[…]

### 2008-08-24 src/intensionalExampleModels.pl
changed the model format slightly.
Probably going to change a lot when I start doing the stuff on embedding.

@@ -4,11 +4,13 @@
 /*========================================================================
    Example Models
 
-   model(Individuals,Worlds,AccessibleRelations).
+   model([Individuals],[Worlds],[AccessibleRelations]).
+   world([AcessibleBy],[Functions])
 ========================================================================*/
 
 example(1,model([d1,d2,d3,d4,d5]
-	, [ (w0,world(
+	, [ world( [curt]
+		,
 		[ f(0,mia,d1)
 		, f(0,vincent,d2)
 		, f(0,pumpkin,d3)
@@ -18,11 +20,7 @@ example(1,model([d1,d2,d3,d4,d5]
 		, f(1,customer,[d1,d2])
 		, f(1,robber,[d3,d4])
 		, f(2,love,[(d3,d4)])
-		]))
-	  , (w1,world(
-		[ f(1,snort,[d2])
-		, f(2,love,[(d2,d5)])
-		]))
+		])
 	  ]
 	, [(d1,w2)]
 	))

### 2008-08-25 src/advertentCurt.pl
work on interpret/2 predicate

@@ -133,19 +133,32 @@ curtUpdate(Input,Moves,run) :-
 	, updateHistory(Input)
 	,
 	% TODO: Write Answer Code
-	interpret(Readings,Model,Moves)
+	interpret(Model,Readings,Moves)
 	.
 
 curtUpdate(_,[noparse],run).
 
-%%% interpret/3
-% take some readings and generate appropriate models based on what's already
-% there. Also inform the rest of Curt about what we do next based on the result.
+domainSize(15).
 
-interpret([],[],[noparse]).
+% fails if Old is not a curt world or empty
+interpret(Old,New,World) :-
+	getKnowledge(Old,New,BK)
+	, check(and(BK,New),'consistency',World)
+	, check(and(BK,not(New)),'informativity',_)
+	.
 
-interpret(Readings,Model,Move).
+getKnowledge(world([curt|_Others],Old,_Functions),New,and(BackgroundKnowledge,Old)) :- 
+	backgroundKnowledge(and(Old,New),BackgroundKnowledge).
 
+getKnowledge([],New,BackgroundKnowledge) :- 
+	backgroundKnowledge(New,BackgroundKnowledge).
+
+check(Formula,Reason,Model) :-
+	domainSize(DomainSize)
+	, callTPandMB(not(Formula),Formula,DomainSize,Proof,Model,Engine)
+	, format('~nMessage (~p checking): ~p found a result.',[Reason,Engine])
+	, \+ Proof=proof, Model=model([_|_],_)
+	.
 
 /*========================================================================
    Combine New Utterances with History

### 2008-08-25 src/advertentCurt.pl
finished interpretation (hope so)
Warning: there *may* be a bug with the backgroundKnowledge stuff, I'm not sure. I think not...

@@ -142,21 +142,23 @@ domainSize(15).
 
 % fails if Old is not a curt world or empty
 interpret(Old,New,World) :-
-	getKnowledge(Old,New,BK)
-	, check(and(BK,New),'consistency',World)
+	getKnowledge(Old,New,BK,Others,Reading)
+	, check(and(BK,New),'consistency',BBModel)
 	, check(and(BK,not(New)),'informativity',_)
+	, BBModel = model(D,F)
+	, World = world(D,[curt|Others],Reading,F)
 	.
 
-getKnowledge(world([curt|_Others],Old,_Functions),New,and(BackgroundKnowledge,Old)) :- 
+getKnowledge(world([curt|Others],Old,_Functions),New,and(BackgroundKnowledge,Old),Others,and(Old,New)) :- 
 	backgroundKnowledge(and(Old,New),BackgroundKnowledge).
 
-getKnowledge([],New,BackgroundKnowledge) :- 
+getKnowledge([],New,BackgroundKnowledge,[],New) :- 
 	backgroundKnowledge(New,BackgroundKnowledge).
 
-check(Formula,Reason,Model) :-
+check(Formula,Job,Model) :-
 	domainSize(DomainSize)
 	, callTPandMB(not(Formula),Formula,DomainSize,Proof,Model,Engine)
-	, format('~nMessage (~p checking): ~p found a result.',[Reason,Engine])
+	, format('~nMessage (~p checking): ~p found a result.',[Job,Engine])
 	, \+ Proof=proof, Model=model([_|_],_)
 	.
 

### 2008-08-25 thesis/paper.tex
Cleaned up section 1 and 2. introduced citations

@@
 \newcommand{\stress}{\textbf} % put stress somewhere (as opposed to \emph)
-\newcommand{\term}{\textsf} % introduce a term
+\newcommand{\term}[1]{\textsf{\textbf{#1}}} % introduce a term
 \newcommand{\code}{\texttt} % layout of in-line code snippets
@@
 %theorem environments
-\theoremstyle{remark} \newtheorem*{termin}{Definition} % terminology. Could use a
-%better title fixme
+\theoremstyle{remark} \newtheorem*{termin}{Terminology} % terminology.
 
@@
 
-\section*{Introduction}
+\section{Introduction}
 
@@
 establish relations between objects of its logical representation. In addition,
-providing a means to compute with said representations is will eventually enable
-the machine-processing of meaning.  In doing so, the ultimate goal may be to
-enable a man-made system to \emph{understand} what a humans communicate.
+providing a means to compute with said representations will eventually enable
+the machine-processing of an abstract concept of meaning. In doing so, the
+ultimate goal may be to enable a man-made system to \emph{understand} what a
+humans communicate.
 
@@
 Development of exhaustive systems in this area appears to be a daunting task of
-quite some scope. Therefore most approaches have been centered on providing
-coverage of at least a fragment of natural language. Many such systems exist and
-they prove to be increasingly successful. Some of them will be discussed here
-and presenting an extension to one system in particular is the main matter of
-this Bachelor's thesis.
+quite some scope. Therefore most approaches have been focused on providing
+coverage of a small fragment of natural language often with respect to a
+particular domain. Many such systems exist and they prove to be increasingly
+successful. Some of them will be discussed here and presenting an extension to
+one system in particular is the main matter of this Bachelor's thesis.
 
-In doing so, the present work will aim to provide the \curt system of
+\subsection{Premises}
+
+The present work will aim to provide the \curt system of
 \cite{blackburnbos:cl1} with the mechanics of dealing with more than one
@@
 
-As
+As the scope of a Bachelor's thesis is limited by its nature, not all of the
+theoretical background can be understood without background knowledge in some
+rather divergent domains. In particular, a basic familiarity with traditional
+\term{natural language semantics} and \term{extensional} as well as
+\term{intensional logic} frameworks will be assumed. Furthermore, a basic
+acquaintance with \term{eroteric logic} will be helpful in understanding some of
+the reasoning in section \ref{sec:theory}.
+
+Some knowledge of the \prol programming language and a good overview of the
+architecture of the \curt system is imperative in order to understand the
+implementation itself.
 
@@
 assumptions of this implementation and justifies some choices and restrictions
-that had to be made because of the limited scope of this analysis. While the
-third section will display and clarify the actual implementation on top of the
+that had to be made due to the limited scope of this analysis. While the
+third section will display and document the actual implementation on top of the
 \curt system, the last section will discuss the results of this work and also
@@
 
-\subsection*{Acknowledgements}
-
-Frank... who else?
-
 \section{Concepts}
 
-The basis of
-
-%fixme: to be formalized
-Where we use ``natural language'', usually the English language will be treated
-in practice, though it would be nice to come up with treatments of other
-languages. We abstract away from pragmatic affairs and concern ourselves with
-mostly the purely semantic/logic domain, so this should be feasible to achieve.
-
-A brief overview of the premisses for the theory (which itself lays ground to
-the implementation).
+todo: write something here... :-(
 
@@
 
-\pn{Curt}, a system written in \prol that is able to interpret basic utterances
-in natural language and store them in expressions of first order logic, is
+\curt, a system written in \prol that is able to interpret basic utterances
+of English and store them in expressions of first order logic, is
 presented in \cite[chapter 6]{blackburnbos:cl1}. While relatively limited when
-compared to other systems presented in \ref{sec:comparision} in terms of lexical
+compared to other systems presented in \ref{sec:comparison} in terms of lexical
 and grammatical coverage, it provides a sound basis for extended work because of
@@
 formulae -- within the bounds of the tools it uses.
-
-% FIXME: don't really like this footnote
-\footnotetext{Here, the notion of \emph{consistency} is meant as \emph{validity}
-of the input and \emph{informativity} denotes \emph{validity}. Thus, if
-\Disc, the discourse so far and $\phi$ a new input token (fully parsed
-sentence), then consistency means $\Disc \to \phi$ holds and informativity
-means $\Disc \nvDash \phi$.}
-
 It can thus detect incoherent discourse and would reject a sequence of
 sentences such as \emph{``Mary likes every man. Peter is a man. Mary does not
-like Peter''}, claiming it found a contradiction in its input. For every input
-token, \pn{Curt} will also try to infer whether the current discourse would
-entail the information given in its newest token and is thus irrelevant with
-regard to previous utterances – effectively allowing it to
-reduce the size of its internal representation of the discourse by not adding
-clauses to it which would not contribute to the overall informational value
+like Peter''}, claiming it found a contradiction in its input. The discourse is
+then said to be \term{inconsistent} with the new input.
+
+For every input
+sentence, \pn{Curt} will also try to infer whether the current discourse would
+entail the information it contains and would thus be \term{uninformative} with
+respect to previous utterances – effectively allowing it to
+reduce the size of its internal representation of the discourse by avoiding to
+add clauses to its database that would not contribute to the overall informational value
 already present in the program.
 
-The system is introduced during the final chapter of \cite{blackburnbos:cl1},
-incrementally increasing in features and complexity. Support for external
-inferencing via theorem provers and model builders is added in a step-by-step
-fashion and finally an ontology and a straight-forward way of parsing and
-answering direct questions for noun phrases conclude its development. The theory
-and implementation of this paper is meant as a replacement and enhancement for
-the latter feature. This will also happen in an incremental manner and new
-features will be introduced one at a time.
+The final chapter of \cite{blackburnbos:cl1} introduces the implementation
+incrementally, presenting new versions of \curt that add to the software's
+feature set. Support for external inferencing via theorem provers and model
+builders is added right after the first naïve implementations and
+finally an ontology and a straight-forward way of parsing and answering direct
+questions for noun phrases conclude its development. The theory and
+implementation of this paper is meant as a replacement and enhancement for the
+latter feature. This will also happen in an incremental manner and new features
+will be introduced one at a time.
 
-\subsection{General Comparison of the Curt System}\label{sec:comparision}
+\subsection{Comparison of the Curt System}\label{sec:comparison}
 
-This section will briefly discuss how the features of the Curt system compare to
-other similar systems or systems with the same goal, namely implementing a
-system that is able to model a logic representation of human language.
+Generally, question answering and semantic parsing are not related and seldom
+interdependent in current implementations of similar systems, with a few notable
+exceptions.
 
-\subsubsection{Early Attempts: \pn{Shrdlu} and \pn{Cyc}}
-%TODO: need some reference for this section!
+\cite{chat} are among  the first to attempt creating a question answering system
+with deep semantic coverage. Their effort has been focused on the geographical
+domain, and their implementation, dubbed \pn{CHAT-80}, is written entirely in
+\prol. The lack of hardware resources of that time forced the system to be
+written with efficiency in mind. Moreover, it had to maintain a statical
+database of facts which was hand-crafted. The \curt system is built on  an
+entirely different set of premises and aims to provide a proof-of-concept of a
+more general purpose system that is also able to extend its knowledge
+dynamically.
 
-Interest in deep semantic parsing of natural language has existed since the very
-beginning of the informational age. One of the first systems to achieve public
-recognition, the \pn{Shrdlu} system
+Soon after \pn{CHAT-80}, two entirely different types of systems emerged. The
+first type is probably the most promising to be used on a wider scale, given
+today's technological standards. Implementations are usually based on
+\term{information retrieval} and the generation of question/answer pairs. Such
+systems are generally referred to as a \abbr{QA} (\term{Question Answering})
+systems.  Examples implementations of such systems are discussed in \cite{rama},
+\cite{iesqa} and many others. Most of these seem to provide a good lexical and
+grammatical coverage due to their large input corpora and
+shallow interpretation techniques. The latter is also the \abbr{QA} technique's
+greatest shortcoming, as no deep analysis is  performed and usually
+only factoids of less then a couple of words in length can be supplied as
+answers.\footnote{But see \cite{aqabtf} for an approach that aims to cover a
+wider range of possible answers.}
 
-While the \pn{Curt} system is unique in its approach to deep semantic analysis, its
-architecture has been crafted in a pragmatic manner. The authors chose to use
-external tools where it was possible in order to avoid reimplementation of
-existing functionality and be able to focus on novel parts of the
-implementation\footnote{The name of the \pn{Curt} system reflects this choice.
-It stands for \emph{\textbf{C}lever \textbf{u}se of \textbf{r}easoning
-\textbf{t}ools}.}. This also allows for great flexibility in adopting new
-solutions and improvements made in the field of inferencing tools for first
-order predicate logic.
+In contrast, \pn{CHAT-80} runs all its queries against a pre-built database
+without the ability to acquire new knowledge from the discourse. While a
+\abbr{QA} implementation will typically just acquire knowledge by consuming
+more input and analysing its statistical properties, programs employing deep
+semantic analysis will have to check their input for logical contradictions in
+to maintain their database in a reasonable\footnote{FIXME pun intended, but
+appropriate? %todo
+} state.
+
+\cite{prontoqa} introduce the \pn{Pronto QA System}, which couples traditional
+corpus-based \abbr{QA} techniques with deep semantic parsing based on
+\term{Discourse Representation Theory}.
+
+A very interesting project, called \pn{Godot} is introduced in \cite{godot}.
+\pn{Godot} is a so-called \term{embodied agent}, a robot with a sophisticated
+software driving interaction with the outer world. As such its conception of
+discourse not only relies on spoken dialogue, but also on sensory input.
+\pn{Godot} is focused on executing orders -- after evaluating their feasibility
+in its particular discourse model -- and does so far not support the answering
+of questions about its state.
 
@@
 At the end of \cite{blackburnbos:cl1}, \curt is able to cope with a small
-grammar and provides some features for interacting with humans in natural
-language. It does only understand basic extensional transitive verbs - modal
-verbs or the treatment of questions are not implemented. Anaphora resolution
-still remains an open issue. The grammar is rather limited and \curt only comes
-with a basic vocabulary oriented around the setting of a popular movie.
+grammar and provides some features for interaction in natural
+language. It does only understand basic extensional transitive and intransitive
+verbs and its support for question answering remains limited.
+Anaphora resolution still remains an open issue.
 
-The aim of this paper is to introduce the reader into question semantics as it
-will be relevant for implementing a basic capability of treating intensional
-verbs like `to \stress{know} whether something is the case' and `to
-\stress{believe} that something might be the case' and then present such an
-implementation within the framework of the \curt system. Further and more
-advanced investigations will be dealt with at the end of this discussion or left
-for further research.
+In the course of this thesis, a new version of \curt will be developed, called
+\acurt. This version is not only aware of its own state, but can also gather
+information about the individuals it knows about. Furthermore, \acurt is
+able to account for several possible readings of a certain sentence at once and
+in parallel, which is a side-effect of the introduction of \term{intensionality}
+into the system.
 
-\section{Theory}
+\acurt is able to parse and, according to the theoretical background of section
+\ref{sec:theory}, correctly interpret sentences as the ones in
+\ref{ex:whemba}-\ref{ex:whembc}.
+
+\ex.
+\a.  Mia knows Vincent snorts.\label{ex:whemba}
+\b.  Mia knows who snorts.
+\b.  Vincent believes that Mia likes Jody.\label{ex:whembc}
+
+For this to work, the internal storage of \curt will have to be restructured
+significantly in order to deal with more than one possible world or situation.
+Its grammar will have to be enhanced so that it can handle \wh-embedding verbs
+like `know' and `believe' and its inferencing mechanisms will have to be
+enhanced.
+
+The following section explains the problems at hand and existing solutions. A
+subset thereof will be implemented in section \ref{sec:implementation}, which
+will also raise some pragmatic issues discussed later on in section
+\ref{sec:discussion}.
+
+\section{Theory}\label{sec:theory}
 
@@
 
-Following \cite{gs:q}, we will not use the natural natural language word
-``question'' but distinguish between \term{questions},
+Following \cite{gs:q}, we will not use the natural English word
+``question'' in its traditional sense, but distinguish between \term{questions},
 \term{interrogative sentences} and \term{interrogative acts} in the following
@@
   An \term{interrogative} is a certain type of sentence in natural language
-  distinguished by certain features\footnote{To quote \cite{gs:q}: \quote{[An
+  distinguished by certain characteristics\footnote{To quote \cite{gs:q}: \quote{[An
   interrogative sentence is] characterized by word order, intonation, question
@@
   whereas an \term{interrogative act} refers to the utterance of an
-  \emph{interrogative}. Finally, a 
+  \term{interrogative}. Finally, a
   \term{question} will denote the semantic entity uttered by the
-  \emph{interrogative act}, whichever form it may have.
+  \term{interrogative act}, as it will be introduced in section
+  \ref{sec:protoq}
 \end{termin}
 
-\abbr{Ty2} is two-sorted type logic\footnote{See \cite{gallin:ty2} for a
+The following abbreviations will be used: 
+\abbr{Ty2} stands for two-sorted type logic\footnote{See \cite{gallin:ty2} for a
 discussion and \cite{gs:sqpa} for its usage in describing the semantics of
-questions}. \abbr{IL} is intensional logic\footnote{As used by \cite{ptq}}. 
-An embedded \wh-phrase looks
-like:%fixme
+questions}. \abbr{IL} stands for intensional logic\footnote{As used by \cite{ptq}}.
 
-\ex. John \stress{knows} \emph{whether} good examples are hard to come up with.
-
-After having elaborated on the grounds of the implementation, we will now
-display the implementation's underlying concepts. % FIXME this sentence is too short
-
-%OUTLINE Presenting a conceptual overview, while arguing for and against stuff.
+%todo OUTLINE Presenting a conceptual overview, while arguing for and against stuff.
 
@@
 
-\subsection{Existing Implementations of Similar Systems}\label{sec:othercrap}
-
-Enhancing \curt by an implementation dealing with questions in natural language
-sparks interest in whether there have been similar efforts made in the past. No
-further extensions to the \curt system aiming at comparable goals are known to the
-author at the time of writing. However, there is a lively interest in providing
-so-called question answering or QA-systems since the early stages of computer
-science. % fixme; cite some ancient 60's paper
-This section discusses a few of the more elaborate approaches: first an early
-but already quite impressive implementation is shown, followed by a newer approach
-using embodied agents and finally an overview over the many database-based
-QA-systems is given. Finally, all this is compared what the current extension to
-\curt can do.
-
-\subsubsection{Godot, the Talking Robot}
-
-\subsubsection{CHAT-80}
-
-\subsubsection{Modern Intensional Database Approaches}
-
 \subsubsection{Advertent Curt}
@@
 
-\prol programs can be self-modifying, or dynamic. This means that the program is
+\prol programs can be self-modifying, or dynamic. In this way the program is
 able to change its own state and thus store information.
@@
 presented in \cite{keller:storage}. \cite{blackburnbos:cl1} call this version
-\pn{Keller storage}. That name will be used from here % FIXME sucks. Please change
+\pn{Keller Storage}. That name will be used from here % FIXME sucks. Please change
 to refer to this particular implementation as presented in
@@
 
-\pn{Keller storage} is already quite capable
+\pn{Keller Storage} is already quite capable
 of dealing with more complex constituents such as nested noun phrases, but it
@@
 with some quantifier scope ambiguities\footnote{\cite{blackburnbos:cl1} mention
-sentences such as \ref{ex:kellersucks}, where Keller storage produces six
+sentences such as \ref{ex:kellersucks}, where Keller Storage produces six
 instead of the desired five readings.
@@
 Clearly, \example{a hash bar} cannot take scope over \example{every owner}, but
-Keller storage predicts this reading.}.
+Keller Storage predicts this reading.}.
 
@@
 %Bos' paper
-}. Since, however, none of the cases discussed so far would require the enhanced
-expressive power of \pn{Hole Semantics} it was decided to go on using \pn{Keller
-Storage} instead, since this is what the original implementation of the \curt
-system relies on.
-%todo: fix capitalization in Keller storage all over the document
+}. However none of the cases discussed so far would require the enhanced
+expressive power of \pn{Hole Semantics} it was decided to not change the \curt
+implementation's current architecture which employs \pn{Keller Storage}.
 
@@
 
-\section{Discussion}
+\section{Discussion}\label{sec:discussion}
 
@@
 \cite{kampreyle:drt}) like the one given in \cite{blackburnbos:cl2} might be
-interesting.
+interesting. The Pronto \abbr{QA} system of \cite{prontoqa} already defines a
+means of representing questions and queries against a Discourse Representation
+Structure.
 

### 2008-08-27 thesis/paper.tex
Section 3

@@
-\documentclass[12pt,a4paper]{article}
+\documentclass[11pt,a4paper]{article}
 
@@
 
+\usepackage{tuebingerspruechle}
+
 \newcommand{\abbr}{\textsf} % how to render abbreviations like IL or Ty2
@@
 \newcommand{\Disc}{\ensuremath{\mathcal{D}}} % the discourse
-\newcommand{\wh}{\textsc{wh}} %
+\newcommand{\wh}{\textsc{wh}}
 
 %names
-\newcommand{\curt}{\pn{Curt}\mbox{ }}
-\newcommand{\acurt}{\pn{Advertent Curt}\mbox{ }}
-\newcommand{\prol}{\pn{Prolog}\mbox{ }}
+\newcommand{\curt}{\pn{Curt}}
+\newcommand{\acurt}{\pn{Advertent Curt}}
+\newcommand{\prol}{\pn{Prolog}}
+\pagestyle{headings}
 
 %theorem environments
-\theoremstyle{remark} \newtheorem*{termin}{Terminology} % terminology.
+\newtheorem{thm}{Theorem}[section]
+\newtheorem{cor}[thm]{Corollary}
+\newtheorem{lem}[thm]{Lemma}
+
+\theoremstyle{remark}
+\newtheorem{rem}[thm]{Remark}
+\theoremstyle{remark}
+\newtheorem*{termin}{Terminology}
+
+\theoremstyle{definition}
+\newtheorem{definition}[thm]{Definition}
+
+\renewcommand{\labelenumi}{(\roman{enumi})}
 
@@
 \usepackage{fontspec}
-\usepackage[xetex,colorlinks,citecolor=magenta]{hyperref}
+\usepackage[xetex,colorlinks]{hyperref}
 \begin{document}
+\renewcommand{\thepage}{\roman{page}}
+\maketitle
+\thispagestyle{empty}
+\begin{center}
+  \vskip 4em
+  Thesis submitted for the Degree of Bachelor of Arts\\
 
-\maketitle
+  \paragraph{Supervisor:} Dr. Frank Richter
+  \vskip 6em
+  Seminar für Sprachwissenschaft\\
+  Eberhard-Karls-Universität\\
+  Tübingen
+\end{center}
+\newpage
 \tableofcontents
 \newpage
-
+\SpruechleAufsagen{Aleksandar L. Dimitrov}
+\newpage
+\renewcommand{\thepage}{\arabic{page}}
+\setcounter{page}{1}
 \section{Introduction}
@@
 
-\subsection{Premises}
+\subsection{Overview}
 
-The present work will aim to provide the \curt system of
+The present work will aim to provide the \curt{} system of
 \cite{blackburnbos:cl1} with the mechanics of dealing with more than one
@@
 As the scope of a Bachelor's thesis is limited by its nature, not all of the
-theoretical background can be understood without background knowledge in some
+theory and reasoning can be understood without background knowledge in some
 rather divergent domains. In particular, a basic familiarity with traditional
@@
 the reasoning in section \ref{sec:theory}.
-
-Some knowledge of the \prol programming language and a good overview of the
-architecture of the \curt system is imperative in order to understand the
+Some knowledge of the \prol{} programming language and a good overview of the
+architecture of the \curt{} system is imperative in order to understand the
 implementation itself.
 
-The first section gives a conceptual overview, introducing the \curt
-system and comparing it to other similar systems in existence. It then goes on
-to explain the target domain of natural language the implementation has to
-account for.
+The first section will give a conceptual overview of the technologies to date
+and the particular software, \curt, that will serve as a basis for extension.
 The next section concerns itself with a more detailed analysis of the theoretic
@@
 third section will display and document the actual implementation on top of the
-\curt system, the last section will discuss the results of this work and also
+\curt{} system, the last section will discuss the results of this work and also
 give a prospect on possible further work on the subject.
@@
 
-todo: write something here... :-(
+Computational Semantics is a discipline that has  lately gained a wider
+focus in literature due to recent progress in different areas.
+Current development in computer hardware has made it possible to focus on
+theoretical soundness and important abstract designs and concepts. On the other
+hand, research in both computer science and linguistics has fueled new ideas and
+intriguing hypotheses emerge from the discussion on a very frequent basis.
+
+This section delineates the ground the software presented in this thesis is
+based on. \ref{sec:curt} introduces the base system of the implementation, \curt
+and \ref{sec:comparison} relates it to other similar systems in existence.
+\ref{sec:extension} then goes on to explain where the \curt{} system lacks in
+features and thus exhibits potential for extension.
 
@@
 
-\curt, a system written in \prol that is able to interpret basic utterances
+\curt, a system written in \prol{} that is able to interpret basic utterances
 of English and store them in expressions of first order logic, is
@@
 
-Besides parsing and translating utterances into logic, \curt is able to perform
+Besides parsing and translating utterances into logic, \curt{} is able to perform
 basic checks for \emph{consistency} and \emph{informativity} on the resulting
@@
 For every input
-sentence, \pn{Curt} will also try to infer whether the current discourse would
+sentence, \curt{} will also try to infer whether the current discourse would
 entail the information it contains and would thus be \term{uninformative} with
@@
 The final chapter of \cite{blackburnbos:cl1} introduces the implementation
-incrementally, presenting new versions of \curt that add to the software's
+incrementally, presenting new versions of \curt{} that add to the software's
 feature set. Support for external inferencing via theorem provers and model
@@
 interdependent in current implementations of similar systems, with a few notable
-exceptions.
+exceptions. As question answering becomes a more and more sophisticated task,
+the systems employed gather an increasing capability of \emph{understanding}
+their (input) domain. On the other side of the fence, interest in reasoning
+tools and inference engines operating on natural language have grown
+considerably over the past decade and some have stretched out to encompass more
+advanced domains of logic and discourse representation.
 
@@
 written with efficiency in mind. Moreover, it had to maintain a statical
-database of facts which was hand-crafted. The \curt system is built on  an
+database of facts which was hand-crafted. The \curt{} system is built on  an
 entirely different set of premises and aims to provide a proof-of-concept of a
@@
 corpus-based \abbr{QA} techniques with deep semantic parsing based on
-\term{Discourse Representation Theory}.
+\term{Discourse Representation Theory}. It is one of the few \abbr{QA} systems
+with a capability of reasoning about its input text. Unfortunately, there seems
+to be no working demo to date. The homepage of the ongoing efforts of Johan Bos
+and his team can be found at the following address:
+\href{http://svn.ask.it.usyd.edu.au/trac/candc/}{\url{http://svn.ask.it.usyd.edu.au/trac/candc/}}.
 
-A very interesting project, called \pn{Godot} is introduced in \cite{godot}.
+\pn{Godot}, a very interesting project, is introduced in \cite{godot}.
 \pn{Godot} is a so-called \term{embodied agent}, a robot with a sophisticated
@@
 \pn{Godot} is focused on executing orders -- after evaluating their feasibility
-in its particular discourse model -- and does so far not support the answering
+in its particular discourse model -- and does so far not seem to support the answering
 of questions about its state.
 
-\subsection{Extending the Curt System}
+\subsection{Extending the Curt System}\label{sec:extension}
 
-At the end of \cite{blackburnbos:cl1}, \curt is able to cope with a small
+At the end of \cite{blackburnbos:cl1}, \curt{} is able to cope with a small
 grammar and provides some features for interaction in natural
@@
 
-In the course of this thesis, a new version of \curt will be developed, called
+In the course of this thesis, a new version of \curt{} will be developed, called
 \acurt. This version is not only aware of its own state, but can also gather
-information about the individuals it knows about. Furthermore, \acurt is
-able to account for several possible readings of a certain sentence at once and
-in parallel, which is a side-effect of the introduction of \term{intensionality}
-into the system.
+information about the individuals it knows about. Furthermore, \acurt{} is able
+to account for several possible readings of a certain sentence at once and in
+parallel, which is a side-effect of the introduction of the notion of
+\term{possible worlds} into the system.
 
-\acurt is able to parse and, according to the theoretical background of section
+\acurt{} is able to parse and, according to the theoretical background of section
 \ref{sec:theory}, correctly interpret sentences as the ones in
@@
 \ex.
-\a.  Mia knows Vincent snorts.\label{ex:whemba}
+\a.  Mia knows (whether) Vincent snorts.\label{ex:whemba}
 \b.  Mia knows who snorts.
@@
 
-For this to work, the internal storage of \curt will have to be restructured
+For this to work, the internal storage of \curt{} will have to be restructured
 significantly in order to deal with more than one possible world or situation.
 Its grammar will have to be enhanced so that it can handle \wh-embedding verbs
-like `know' and `believe' and its inferencing mechanisms will have to be
-enhanced.
-
-The following section explains the problems at hand and existing solutions. A
-subset thereof will be implemented in section \ref{sec:implementation}, which
-will also raise some pragmatic issues discussed later on in section
-\ref{sec:discussion}.
+like `know' and `believe' and its inferencing mechanisms are going to be
+employed in a different manner. %todo sucks.
 
@@
 
-This section first gives an overview over the terminology that will be used for
-the discussion of the formal treatment of questions. It then elaborates on the
-differences between different logical formalisms employed in the treatment of
-natural language questions. After a small overview on the types of questions
-(and pointing out the subset of questions that will be treated), %fixme maybe a
-% bit sloppy
-the discussion
-turns to how those questions are best treated in the given formalism.
+Question semantics has traditionally been a very sophisticated field of
+linguistics and philosophy, with many different approaches arguing for and
+against each other. This section will not try to give an exhausting overview of
+the field of question semantics since this would be a task of considerable size
+and by far beyond the scope of this thesis\footnote{For a good summary of some
+of the literature, particularly the issue of whether questions are a concern of
+pragmatics or semantics, consider \cite[section 2 and 3]{gs:q}.}. Instead, it
+was chosen to roughly follow the work by Groenendijk and Stokhof, lending from
+other places where deemed appropriate or practical. As such, the theoretical
+framework of \acurt{} is not going to be an exact realization of any particular
+approach to question semantics, but rather an accumulation of interesting
+concepts from different sources. The main sources, however, are \cite{gs:q} and
+\cite{gs:sqpa}.
 
-The reader should note that all choices presented in this section have been made
-with the implementation of section \label{sec:implementation} in mind.
-% todo talk a bit more
+Following an overview of the terminology that will be used for the discussion of
+the formal treatment of questions, an  elaboration on the differences between
+different logical formalisms employed in the treatment of questions in natural
+language sets the stage for the discussion of some theoretical frameworks and
+issues. Then, the target domain of natural language is introduced by examples
+alongside the mechanisms employed to interpret them.
+
+Note, however, that all choices presented in this section have been made
+with the implementation of section \ref{sec:implementation} in mind.
+Sometimes, theoretical soundness might suffer from these choices as the main
+goal of the present work is to provide a proof-of-concept of a software that is
+able to reason about possible worlds, employing first-order logic inferencing
+tools. % really? fix your title then, would you? … doh. fixme
 
@@
 
-Following \cite{gs:q}, we will not use the natural English word
+Following \cite{gs:q}, we will not use the English word
 ``question'' in its traditional sense, but distinguish between \term{questions},
@@
 
+Furthermore, interrogatives are split into several major categories.
+
+\ex. \a. \label{ex:yesno} Q: Does Mia smoke? A: \emph{Yes.} or \emph{No.}
+\b. \label{ex:const} Q: Who smokes? A: \emph{Mia smokes.} or \emph{Vincent
+somkes.} or \emph{Nobody smokes.}
+
+Example \ref{ex:yesno} is considered an \term{alternative question} or
+\term{yes/no-interrogative} since only \emph{yes} and \emph{no} are possible
+answers to this interrogative. \ref{ex:const} on the other hand is considered a
+\term{\wh-question} or \term{constituent interrogative} since one of the
+sentences' nominal constituents is replaced by a \wh-pronoun, prompting an
+enumeration of possible fillers for the resulting gap as an answer.
+
 The following abbreviations will be used: 
@@
 discussion and \cite{gs:sqpa} for its usage in describing the semantics of
-questions}. \abbr{IL} stands for intensional logic\footnote{As used by \cite{ptq}}.
+questions}. \abbr{IL} stands for intensional logic\footnote{As used by
+\cite{ptq}. See also \cite{gamut:2} for an excellent introduction to \abbr{IL}.},
+\abbr{FOL}  is used as an abbreviation for first order logic.
 
-%todo OUTLINE Presenting a conceptual overview, while arguing for and against stuff.
-
-First we will present a brief overview of the considerations regarding the
-choice of an appropriate formalism for the system in section \ref{sec:formal}
-to then compare it with already existing implementations of question answering
-systems in \ref{sec:othercrap}.
-
-\subsection{Formalism}\label{sec:formal}
-
-% FIXME if we have a `Terminology' section, all those `henceforths' aren't really
-% necessary anymore.
+\subsection{Formalism: Models}\label{sec:formal}
 
@@
 \cite{ptq} and thus also makes use of its underlying formalism,
-\emph{Intensional Logic} (henceforth \abbr{IL}). \cite{gs:sqpa} (and many subsequent
-publications) however also use two-sorted type logic (henceforth \abbr{Ty2}) as
+\abbr{IL}. \cite{gs:sqpa} however use \abbr{Ty2} as it was
 proposed in \cite{gallin:ty2}. \cite{z:ilty2} discusses the differences and
@@
 translation from \abbr{Ty2} to \abbr{IL} seems to sometimes produce very complex
-\abbr{IL} terms and is therefore avoided here. In fact, \abbr{IL} will be only
-marginally relevant for the presentation of \acurt – only the type system
-presented in \cite{ptq} is carried over.
+\abbr{IL} terms.
 
-% OK till now, but the rest is a fixme :-(
+As was already mentioned in \ref{sec:curt}, \curt{} is based on 
+the use of external tools only capable of treating first order logic – an
+extensional framework. \prol{} on the other hand, is a language that can very well
+make use of higher order formalisms for calculation\footnote{The two most
+popular predicates in \prol{} used for implementing higher-order logic are
+\code{call/n} and \code{apply/3}. This paper is based on \pn{SWI-Prolog} which
+provides an implementation of the former, but not of the latter. In addition
+\pn{SWI-Prolog} also defines several higher order aggregate functions, including
+\code{maplist/4}, which will be used later on. See \cite{naish:prolhio} for a
+discussion of higher order functional programming in \prol.}.  While there exist
+inferencing tools for modal logic\footnote{ The interested reader is referred to
+\pn{Molle} (\url{http://molle.sf.net}), a modal theorem prover that comes with a
+graphical user interface which allows interactive investigation of the
+constructed indices.}, a restructuring of the \curt{} system to take advantage
+of them would be beyond the scope of this thesis.  It was therefore decided to
+avoid modal logic and implement a pragmatic approach to representing the indices
+of higher-order logic.
 
-Since this paper is not concerned with possible extensions of the underlying
-theories of treating natural language questions, but rather with their
-implementation in the programming language \pn{Prolog} several additional
-constraints have to be considered. Among them the \emph{feasibility} of any
-possible approach takes highest precedence.
+Some research on the logic of questions (eroteric logic) suggests that the
+treatment of interrogatives demands its own formalism.  \acurt{} employs a
+formalism that departs from both \abbr{IL} and \abbr{Ty2} in some ways. Most
+notably, the notion of a \term{model} is changed as follows, in order to
+accommodate the question operator introduced further on, which demands a set of
+possible worlds to be available.
 
-To quote \cite{blackburnbos:cl1}:
+\begin{definition}\label{Models}
+  \stress{Models\footnote{See also the Appendix, section \ref{sec:examplemodel}
+  for an example of such a model.}}
+  \begin{enumerate}
+    \item A model is a set of pairs of \term{indices} $i$ and \term{possible
+    worlds} $w$: $\pair{i,w} \in M$. 
+    \item A world $w$ at index $i$, $\pair{i,w} \in M$ is a triple of $D$, the set
+    of individuals, $F$, the set of functions, and $\varphi$ a first-order
+    logical formula that holds in this particular model: $w = \pair{D,F,\varphi}$
+    \item A \term{function} $f \in F$ is one of the following: 
+    \begin{itemize}
+      \item $f(a,\sigma,L)$, where $\sigma$ is the function symbol, $a$ denotes
+      the function symbol's arity and $L$, the set of arguments, for which the
+      following holds: $|L| = a$,
+      $L = \set{\pair{x_1,\ldots{},x_a}|x_1,\ldots{},x_a \in D}$
+      \item The special case $f(0,\sigma,c)$, where $c \in D$, which denotes the
+      assignment of a proper name $\sigma$ to $c$.
+    \end{itemize}
+    \item $\pair{i,w}, \pair{j,w'} \in M, w = w'$ iff $w = \pair{D,F,\varphi}, w' =
+    \pair{D',F',\varphi'}$ and $\varphi = \varphi'$.
+    \item If $W \in M$, then we say $\forall W$ quantifies over all members of $M$.
+    \item If $W \in M$, then we say $\exists W$ quantifies over the existence of
+    one member of $M$, namely $|M| \geq 1$ and $W \in M$.
+  \end{enumerate}
+\end{definition}
 
-\begin{quote}
-  Questions, unlike assertions, don't have truth-values, so it would be
-  misleading to represent them as ordinary formulas.
-\end{quote}
+One of the more important differences to the `traditional' notion of a model, is
+that it was decided to assign every world a unique domain of discourse referents
+or \term{individuals}. Note that there is no requirement $\pair{i,\pair{D,F}},
+\pair{j,\pair{D',F'}} \in M \wedge D\neq D'$, and most of the time, a model's
+possible worlds will indeed share many of their individuals or have entirely
+equal domains, but diversion in the world's domains will sometimes be useful and
+unavoidable.
 
-It is important to note that this claim is not undisputed and
-\cite{karttunen:1977}, argues against it. The present approach, however will
-base itself on this assumption and extend it by more formal approaches than the
-one given by Blackburn \& Bos. Specifically, the work by J. Groenendijk and M. Stokhof
-will lay ground to the implementation. But in order to realise their formal
-discussions computationally, modifications will have to be made, and even
-compromises.
+Moreover, every possible world carries a formula $\varphi$, to which $\pair{D,F}$
+is the minimal extensional model. This justifies the choice to define equality
+of possible worlds in terms of equality between the logic expressions $\varphi$
+they carry. This is also necessitated by the fact that \acurt{}  only reasons,
+i.e. runs its reasoning tools on these formulae and can therefore not make a
+difference between possible worlds in any other terms. $\pair{D,F}$ will serve
+only as a ground for answering questions. This entails, that the traditional
+difference between \term{models} and \term{readings} present in \curt, is not
+carried over, but instead the readings are now constituents of a model.
 
-As was already mentioned in \ref{sec:curt}, Curt is based on an
-implementation and the use of external tools only capable of treating first
-order logic – an extensional framework. \prol on the other hand, is a language
-that can very well make use of higher order formalisms for
-calculation\footnote{The two most popular predicates in \prol used for
-implementing higher-order logic are
-\code{call/n} and \code{apply/3}. This paper is based on \pn{SWI-Prolog} which
-provides an implementation of the former, but not of the latter. See
-\cite{naish:prolhio} for a discussion of higher order functional programming in
-\prol.}.
+Defining quantification over world-index pairs enables \acurt{} to make a
+distinction between \term{exhaustive} and \term{partial answers}. See section
+\ref{sec:answerhood}.
 
-While there exist inferencing tools for modal logic\footnote{See %todo todo and todo 
-for discussions. The interested reader is also referred to \pn{Molle}
-(\url{http://molle.sf.net}), a modal theorem prover that comes with a graphical user
-interface which allows interactive investigation of the constructed indices.},
-a restructuring of the \pn{Curt} system to take advantage of them would be
-beyond the scope of this thesis.
-
-It was therefore decided to avoid modal logic and implement a pragmatic approach
-to representing the indices of higher-order logic in \prol in section
-\ref{sec:indices}. This logic however has to be augmented with question
-operators as already used by \cite{karttunen:1977}. % todo: writeme!
-
-\cite{g:is} presents an approach to implementing inquisitive semantics in
-predicate logic.
+Finally, logic expressions in \acurt{} do not typically carry any kind of type,
+since \prol{} does not employ a type system. The grammar rules of \curt{} (which
+\emph{do} have \stress{syntactic} types) ensure a correct application of
+unsaturated formulae. They belong in the domain of first-order predicate logic,
+with two enhancements, introduced in section \ref{sec:protoq} and \ref{sec:epistemic}
 
@@
 
+This section is going to elaborate on some of the approaches in greater detail
+and also explain \acurt's notion of a question in particular. Note that the
+latter differs from the literature, but most differences are superficial.
+
+\subsubsection{The Question of Domain}
+
 Often question semantics has been reduced to a problem of
@@
 
-This approach has been challenged and proven unfeasible by \cite{gs:q},
-who have also proposed interesting extensions
-to the existing logic frameworks used in natural language semantics.
-More recent publications, such as \cite{gal:tmbi}
-address this topic and aim to introduce notions of \term{information change
-potential} as opposed to the traditional focus of natural language semantics on
-\term{truth values}.
+The issue is ultimately whether the answering of questions belongs in the realm
+of pragmatics or semantics. Tichy's approach has been challenged and proven
+unfeasible by \cite{gs:q}, who argue for a semantic interpretation of questions.
+They also propose certain adequacy criteria for theories of the semantics of
+questions in order to be able to compare different approaches made in the
+literature and also to defend their own solution. Their analysis is based on the
+introduction of another type of logical expression, an \term{interrogative}.
+In contrast to indicatives, interrogatives do not yield a truth value, but
+instead pose a \emph{question}. They will be formally introduced in section
+\ref{sec:protoq}.
 
-The present work will rely on \cite{gs:q} for providing the main guidelines for
-the theoretic aspects and also lend from earlier works, most importantly
-\cite{gs:sqpa}\footnotemark. Later on, in section \ref{sec:altq} we will also
-discuss the very newest (at the time of writing) developments in question
-semantics, depicted in \cite{g:is}.
-
-\footnotetext{\cite{gs:sqpa} is actually a series of three articles
-that have appeared before. Among them, we will mostly consider \cite{gs:sawhq}.
-Both, the earlier paper and the later dissertation (and many other papers
-originating in Amsterdam) have been kindly made
-available electronically, free of charge, by the Universiteit van Amsterdam at
-the following address: \url{http://dare.uva.nl}}
-
-\subsubsection{Pragmatics and Semantics of Questions}
-
-In choosing to treat only embedded questions, we can escape most pragmatic
-considerations and focus on the (implementation of) pure semantics of
-interrogatives.
-
-\subsection{Classification of Interrogatives}
-
-When treating questions one will be confronted with many different
-categorizations of questions. One of the first distinctions to make is the one
-between \term{embedded questions} as in \ref{ex:embed} and \term{direct
-questions} shown in \ref{ex:direct}.
-
-\ex. \a. John knows \emph{whether this sentence contains an embedded
-question}.\label{ex:embed}
-\b. Does this sentence pose a direct question?\label{ex:direct}
-
-As indicated already by the title, this paper will concern itself with the
-treatment of the former and only briefly comment on the latter. The reasons for
-this are numerous  %fixme I need a synonym for that
-and this section is going to name the most important ones.
-
-Early literature, such as \cite{karttunen:1977} only regards embedded
-\wh-phrases, treating direct questions as semantically equivalent. To quote
-Karttunen:
+It is important to note that \cite{karttunen:1977} attempts a treatment of
+questions in a similar manner, but also \emph{denies} a difference in logical
+type for questions as such. According to his hypothesis, direct questions are a
+syntactic variant of indirect (embedded) questions and yes/no interrogatives are
+a ``syntactically degenerate variant'' of constituent interrogatives.
 
@@
 thus gives an approach for indirect questions only.
+In this respect, his analysis is in certain aspects similar to the one presented in
+\cite{gs:sawhq}, but the latter goes on to challenge certain aspects of his
+hypothesis.
 
-\subsubsection{An Ontology of Embedding Verbs}
+\cite{blackburnbos:cl1} treat interrogatives in a similar manner as \cite{gs:q},
+using a special operator. From chapter six of \cite{blackburnbos:cl1}:
 
-\cite{karttunen:1977} introduces means of classification for \wh-embedding
-verbs, the nature of which \cite{lahiri:diss} covers in a much broader scope.
-The different verb classes ask for different strategies of treating the
-\wh-clauses they embed and also pose different restrictions on the type of
-clause they expect.
+\begin{quote}
+  Questions, unlike assertions, don't have truth-values, so it would be
+  misleading to represent them as ordinary formulas.
+\end{quote}
 
-The present work will not consider a full analysis of \wh-embedding verbs, but
-confine itself on a few ones, namely \example{know}, \example{think} and
-\example{believe}. The grammar implementation of \acurt allows for \wh-embedding
-verbs to have a \code{type} field, determining their semantic denotation. The two
-types used so far are \code{assert} for assertive verbs such as \example{know}
-and \code{stipul} for stipulating verbs, such as \example{believe} and
-\example{think}.
+Their notion of questions will be extended for \acurt in order to account for
+yes/no interrogatives and larger sets of answers made possible by using model
+which allow for possible worlds.
 
-\subsubsection{Grammatic Constraints}
+Another important distinction in the definition of questions, is the definition
+of \term{answerhood}, or, what constitutes a valid answer to a given
+interrogative. The ground-laying work by \cite{hamblin:q} states that in knowing
+all possible answers to a question, one also knows the question itself. The
+notion of answerhood is going to be elaborated on in section \ref{sec:answerhood}
+
+One thing remains clear though, no matter what the notion of a question or
+finding an answer to it is; upon
+finding an answer, said answer has to be \emph{realized} in natural language
+using text or speech synthesis. When realizing an answer, many pragmatic
+considerations have to be taken into account.
+However, as the uttering of an embedded question does not prompt an
+answer\footnote{Unless, of course, the question is uttered as an
+\term{imperative} -- but this would pose an entirely different class of problem
+that is not discussed here.}, the present
+work will not directly concern itself with a solution to this problem.
+
+Constituent interrogatives also differ in what they ask for. One can assume the existence of at
+least three different types:
+
+\begin{enumerate}
+  \item NP-questions ask for a noun phrase, e.g. \example{Who shot Vincent?
+  Butch} or even more then one noun phrase at the same time \example{Who had
+  what in the hash bar?}
+  \item PP-questions ask for prepositional constituent, e.g. \example{Where did
+  Jules stay? In Amsterdam.}
+  \item VP-questions ask for an action performed by an individual, e.g.
+  \example{What did Butch do? He shot Vincent.}
+  \item Temporal questions ask for the date of a certain event, e.g.
+  \example{When did Butch shoot Vincent? When he entered the toilet.}
+  \item etc\ldots
+\end{enumerate}
+
+This classification of interrogatives can go on for quite some more entries,
+including temporal, modal, causal and all sorts of other questions. Out of all
+those categories, only NP-questions will be supported, with an option to
+integrate other sorts of questions at a later stage.
+
+The remainder of this section will mostly compare the first (purely extensional)
+approaches by \cite{blackburnbos:cl1} with the more general framework of
+\cite{gs:q} and come up with a representation of the latter within the lines of
+the implementation of the former. It will assume a modal logic as was discussed
+in section \ref{sec:formal}, but an effort is made to isolate the reasoning on
+possible worlds in just a few rules, so the rest of the framework can stay
+extensional and thus benefit from \curt's external reasoning tools and present
+internal mechanisms.
+
+\subsubsection{The Semantics of Interrogatives: The ? Operator}\label{sec:protoq}
+
+\pn{Helpful Curt}, the most advanced version of \curt{} \cite{blackburnbos:cl1}
+present, only addresses the problem of direct constituent interrogatives and
+does neither treat yes/no interrogatives nor embedded questions. \acurt treats
+embedded questions by embedding a special construct, a \term{proto-question},
+under verbs of certain types (which are going to be discussed in
+\ref{sec:ontology}). Furthermore, the yes/no interrogative is considered a
+special case of the constituent interrogative.
+
+\begin{definition}\label{def:qop}
+  \stress{The Question Operator}
+  \begin{enumerate}
+    \item If $\varphi$ is a proposition in which the free variables
+    $x_1,\ldots,x_n$ for arbitrary values of $n \geq 1$ occur,
+    $?\pair{x_1,\ldots,x_n}.\varphi$ denotes a \term{question}, associated with
+    a constituent interrogative.
+    \item If $\varphi$ does not contain any free variables, $?.\varphi$ is
+    considered a question associated wit a yes/no interrogative.
+    \item The question operator does \emph{not combine} with any other operator.
+  \end{enumerate}
+\end{definition}
+
+To give two examples of the use of the question operator:
+
+\ex. \a. Who shot Vincent?
+\b. $?x.shot(x,vincent)$
+
+\ex. \a. (Does) Vincent snort(s) or smoke(s)?
+\b. $?.snort(vincent) \vee smoke(vincent)$
+
+In the latter example, English grammar in written form will demand the phrase to
+contain the auxiliary verb \emph{does}, although spoken form would permit
+\example{Vincent \stress{snorts}?} as a question with a certain intonation
+pattern on the verb. This is not of relevance to the treatment of embedded
+questions, where the auxiliary is not needed.
+
+Note that the last clause of definition \ref{def:qop} enforces the question operator to
+always be the last operator of a formula of propositional logic. Embedded
+questions will thus not have to operate on the questions directly, but rather on
+their value (answers), or the set of propositions they denote.
+
+As was already mentioned, \pn{Helpful Curt} already has a certain albeit limited
+support for evaluating NP constituent interrogatives. Blackburn and Bos adopt a
+sort of pseudo-notation of the question operator in what they call an
+extra-logical form which does not itself a yield a truth value, but is
+internally evaluated in a special case.
+
+They treat a question as a sort of \emph{quantification
+over a certain individual $X$, for which $R$ and $S$ must hold.} Thus a parse of
+a sentence like \ref{ex:bbqueexa} and \ref{ex:bbqueexb} would yield the representation given in
+\ref{ex:bbqueparsea} and \ref{ex:bbqueparseb}.
+
+\ex. \a. \label{ex:bbqueexa} Which boxer likes Jody?
+\b. \label{ex:bbqueparsea} \code{que(X, boxer(X), like(X, jody))}
+
+\ex. \a. \label{ex:bbqueexb} Who shot Vincent?
+\b. \label{ex:bbqueparseb} \code{que(X, person(X), shoot(X, vincent))}
+
+The aforementioned special case in \pn{Helpful Curt} only evaluates a question
+against the extensional model which is used as a representation of \curt's state
+and knowledge. In this it differs substantially from the approaches made by
+Groenendijk and Stokhof and also later approaches by \cite{g:is}. \acurt will
+bridge the gap between the two.
+
+\subsubsection{What Answers the Question?}\label{sec:answerhood}
+
+from questions...
+                   Hamblin’s postulates read as follows:
+     (i) An answer to a question is a sentence, or statement.
+    (ii) The possible answers to a question form an exhaustive set of mutually
+exclusive possibilities.
+   (iii) To know the meaning of a question is to know what counts as an answer to
+that question.
+
+\subsubsection{The Syntax of Questions}
 
@@
 
-\subsubsection{Extensional WH-embedding Verbs}
+\subsection{Embedded Questions}
 
-Like \emph{know}.
+When treating questions one will be confronted with many different
+categorizations of questions. One of the first distinctions to make is the one
+between \term{embedded questions} as in \ref{ex:embed} and \term{direct
+questions} shown in \ref{ex:direct}.
+
+\ex. \a. John knows \emph{whether this sentence contains an embedded
+question}.\label{ex:embed}
+\b. Does this sentence pose a direct question?\label{ex:direct}
+
+As indicated already by the title, this paper will concern itself with the
+treatment of the former and only briefly comment on the latter. The reasons for
+this are numerous  %fixme I need a synonym for that
+and this section is going to name the most important ones.
+
+Early literature, such as \cite{karttunen:1977} only regards embedded
+\wh-phrases, treating direct questions as semantically equivalent. To quote
+Karttunen:
+
+\subsubsection{Types of Facts, Assertions and Interrogatives}
+
+\subsubsection{Towards an Ontology of Embedding Verbs}\label{sec:ontology}
+
+\cite{karttunen:1977} introduces means of classification for \wh-embedding
+verbs, the nature of which \cite{lahiri:diss} covers in a much broader scope.
+The different verb classes ask for different strategies of treating the
+\wh-clauses they embed and also pose different restrictions on the type of
+clause they expect.
+
+The present work will not consider a full analysis of \wh-embedding verbs, but
+confine itself on a few ones, namely \example{know}, \example{think} and
+\example{believe}. The grammar implementation of \acurt{} allows for \wh-embedding
+verbs to have a \code{type} field, determining their semantic denotation. The two
+types used so far are \code{assert} for assertive verbs such as \example{know}
+and \code{stipul} for stipulating verbs, such as \example{believe} and
+\example{think}.
 
@@
 
-\subsubsection{Intensional WH-embedding Verbs}
+\subsubsection{Modeling Knowledge}\label{sec:epistemic}
 
-Like \emph{believe}.
+\subsection{Treating Disjunction}
 
-\section{Implementation}
+\subsection{Inquisitive Semantics}\label{sec:altq}
+
+The very latest research now focusses on providing an adequate logical semantics
+for treating questions and even discourse within a dialogue. Groenendijk, in his
+theory he calls ``Inquisitive Semantics'', argues for a completely new of of
+treating dialogues (``Dialogue Management'') -- which naturally includes a
+treatment of questions. His work on this topic has been prepared in \cite{g:is}
+and is further developed in the notes to his course held at ESSLLI in Hamburg
+(\cite{g:isdm}), just two weeks before work on this paper was to be finished.
+Because of this time constraint, and because the latter paper is to date an
+unfinished document, his theory could not be considered as a ground for the
+implementation of \acurt. His findings, however, may very well serve as a ground
+for further research on dialogue systems.
+
+\subsubsection{The Rest}
+
+The present work will rely on \cite{gs:q} for providing the main guidelines for
+the theoretic aspects and also lend from earlier works, most importantly
+\cite{gs:sqpa}\footnotemark. Later on, in section \ref{sec:altq} we will also
+discuss the very newest (at the time of writing) developments in question
+semantics, depicted in \cite{g:is}.
+
+\footnotetext{\cite{gs:sqpa} is actually a series of three articles
+that have appeared before. Among them, we will mostly consider \cite{gs:sawhq}.
+Both, the earlier paper and the later dissertation (and many other papers
+originating in Amsterdam) have been kindly made
+available electronically, free of charge, by the Universiteit van Amsterdam at
+the following address: \url{http://dare.uva.nl}}
+
+\section{Implementation} \label{sec:implementation}
+
+Originally, \prol{} was conceptualized as a programming language for a subset of
+first-order logic. This entails the predominant usage of \abbr{FOL} in section
+\ref{sec:implementation}, but some higher order predicates introduced to the
+language in recent years will also be put to use.
 
@@
 
-\acurt lends its name from the fact that it is not only aware of its own
+\acurt{} lends its name from the fact that it is not only aware of its own
 knowledge and presuppositions, but `knows' about what individuals in its
@@
 
-\subsubsection{Representing Intensionality in \prol}
+\subsubsection{Representing Intensionality in \prol{}}
 
-\prol programs can be self-modifying, or dynamic. In this way the program is
+\prol{} programs can be self-modifying, or dynamic. In this way the program is
 able to change its own state and thus store information.
@@
 
-Another example where \curt has to sacrifice exhaustiveness over its one-world
+Another example where \curt{} has to sacrifice exhaustiveness over its one-world
 representation of the world is given in \ref{ex:ambig}
@@
 stick with the minimal model constructed by the model builder. This is not
-really satisfactory and extending the dimensions of knowledge within \curt seems
+really satisfactory and extending the dimensions of knowledge within \curt{} seems
 necessary in order to treat such examples.
 
-The previous work on \acurt enables it to generate a \term{choice point} -- and
+The previous work on \acurt{} enables it to generate a \term{choice point} -- and
 moreover, not pursue only one path of it, but both at the same time, by
@@
 
-\curt parses sentences of natural language using an enhanced version of \pn{Cooper
+\curt{} parses sentences of natural language using an enhanced version of \pn{Cooper
 storage}\footnote{See \cite{cooper:storage2} for the original implementation.} as
@@
 }. However none of the cases discussed so far would require the enhanced
-expressive power of \pn{Hole Semantics} it was decided to not change the \curt
+expressive power of \pn{Hole Semantics} it was decided to not change the \curt{}
 implementation's current architecture which employs \pn{Keller Storage}.
 
-Enhancing the grammatic rules of the \curt system to cope with
+Enhancing the grammatic rules of the \curt{} system to cope with
 proposition-embedding verbs is made trivial by its elaborate grammatical back
@@
 
-\subsection{Treatment of Baker Ambiguities}
-
-\subsection{Alternative Questions – Inquisitive Semantics}\label{sec:altq}
-
-\cite{g:is} introduces a novel approach to treating questions in natural
-semantics: \pn{Inquisitive Semantics}. He presents a new logic
-formalism aimed at providing a wider coverage of the intentions of question in
-human language. The main source of the ``inquisitiveness'' of statements is
-given by the interpretation of \emph{disjunction}. We will now show how to
-incorporate this novel approach into the \pn{Curt} system.
-
-We will, however depart from the path we have been following throughout this
-paper and first implement a theory for direct questions. Only then will we extend
-it to also cover embedded alternative questions.
-
 \section{Discussion}\label{sec:discussion}
 
-We have seen, the Curt system is quite capable of treating questions this way.
+We have seen, the \curt{} system is quite capable of treating questions this way.
 Having implemented modality and possible worlds in such a fashion, let's go on
@@
 
-\subsection{Whatnot}
+\subsection{Examples}
+
+\subsubsection{An Example Model} \label{sec:examplemodel}
+
+\begin{eqnarray*}
+  M &=& \set{\pair{1,\pair{D_1,F_1,\varphi_1}},\pair{2,D_2,F_2,\varphi_2}} \\
+  D_1 &=& \set{d_1,d_2,d_3} \\
+  F_1 &=& \{f(0,mia,d_1), f(0,vincent,d_2), f(0,jody,d_3),\\
+  && f(1,woman,\set{ d_1,d_3 }), f(1,man,\set{ d_2 }),
+  f(2,like,\set{\pair{d_2,d_1},\pair{d_3,d_2}})\} \\
+  \phi &=& \exists x,y,z [mia(x)\wedge woman(x)\wedge \wedge vincent(y)\wedge
+  man(y)\wedge jody(z)\wedge woman(z) \\&&\wedge x \neq y \wedge x\neq z \wedge z
+  \neq y\wedge like(x)(y)\wedge like(y)(z)]\\
+  D_2 &=& 
+\end{eqnarray*}
+
+\newpage
 

### 2008-08-29 thesis/paper.tex
A lot of work

@@
-\documentclass[11pt,a4paper]{article}
+\documentclass[a4paper]{article}
 
@@
 
-\usepackage{amssymb,qtree,natbib,aleks,linguex}
-
-\usepackage{amsthm}
+\usepackage{amssymb,qtree,natbib,bbm,aleks,linguex}
+
+\usepackage{stmaryrd,amsthm,amsmath}
 
@@
 \setsansfont [Ligatures={Common}, BoldFont={Fontin Sans Bold}, ItalicFont={Fontin Sans Italic}]{Fontin Sans}
+\setmonofont [Ligatures={NoCommon,NoRequired,NoRare}] {Monaco}
 
@@
 \newcommand{\term}[1]{\textsf{\textbf{#1}}} % introduce a term
-\newcommand{\code}{\texttt} % layout of in-line code snippets
+\newcommand{\code}[1]{\footnotesize\texttt{#1}\normalsize} % layout of in-line code snippets
 \newcommand{\pn}{\textsf} % how to render proper names like Shrdlu
@@
 \newcommand{\wh}{\textsc{wh}}
+
+%quotations
+\newenvironment{quotes}{\begin{quote}\sf}{\rm\end{quote}}
+\newcommand{\quotesf}[1]{\quote{\textsf{#1}}}
 
@@
 \newtheorem{definition}[thm]{Definition}
+
+\theoremstyle{definition}
+\newtheorem{postulate}[thm]{Postulate}
 
@@
 be) able to extract logical formulae from sentences of ordinary language and
-establish relations between objects of its logical representation. In addition,
+establish relations between objects of their logical representation. In addition,
 providing a means to compute with said representations will eventually enable
 the machine-processing of an abstract concept of meaning. In doing so, the
-ultimate goal may be to enable a man-made system to \emph{understand} what a
+ultimate goal may be to enable a man-made system to \emph{understand} what
 humans communicate.
@@
 Exploration of this  interesting field of linguistics has shown that
-the notion of \emph{understanding} is far more than a linguistic concern, but
+the notion of \emph{understanding} is far more than a linguistic concern, and
 involves at least findings in the domain of logics, (cognitive) psychology and
@@
 \term{natural language semantics} and \term{extensional} as well as
-\term{intensional logic} frameworks will be assumed. Furthermore, a basic
-acquaintance with \term{eroteric logic} will be helpful in understanding some of
-the reasoning in section \ref{sec:theory}.
-Some knowledge of the \prol{} programming language and a good overview of the
-architecture of the \curt{} system is imperative in order to understand the
-implementation itself.
+\term{intensional logic} or \term{modal} frameworks will be assumed.
+Furthermore, a basic acquaintance with \term{eroteric logic} will be helpful,
+but not strictly necessary in understanding some of the reasoning in section
+\ref{sec:theory}.  Some knowledge of the \prol{} programming language and a good
+overview of the architecture of the \curt{} system is imperative in order to
+understand the implementation itself.
 
@@
 
+\subsection{Typographic Conventions}
+
+The following typographic conventions are used throughout the document to
+increase readability:
+
+\begin{itemize}
+  \item Text in a bold face sans serif font marks new terminology and important
+  concepts: \term{embedded interrogatives}, \term{backtracking}
+  \item Proper names, names of technologies, and abbreviations are typeset in a
+  regular sans serif font: \acurt, \prol, \pn{Ty2}
+  \item Code is written in a monospace font, resembling the look of a font class most
+  programmers are used to:\\
+  \code{lam(P, lam(Q, all(X, imp(app(P,X),app(Q,X)))))}.
+  \item Logical formulae are typeset in a different serif font from the rest of
+  the text:\\
+    Assume $k = (\lambda x.\neg xx)$. Then $ kk = (\lambda x .\neg xx)k = \neg kk$
+  \item Quotations  are typeset on a separate paragraph, using a sans serif font:
+  \quotesf{This answers the question!}
+\end{itemize}
+
 \section{Concepts}
 
-Computational Semantics is a discipline that has  lately gained a wider
-focus in literature due to recent progress in different areas.
-Current development in computer hardware has made it possible to focus on
-theoretical soundness and important abstract designs and concepts. On the other
-hand, research in both computer science and linguistics has fueled new ideas and
-intriguing hypotheses emerge from the discussion on a very frequent basis.
+Computational Semantics is a discipline that has  lately gained a wider focus in
+literature due to recent progress in different areas.  Current development in
+computer hardware has made it possible to focus on theoretical soundness and
+important abstract designs and concepts of logic and the theory of meaning
+rather than being able to simply implement a system that is able to do basic
+computations in a reasonable amount of time. On the other hand, research in both
+computer science and linguistics has fueled new ideas and intriguing hypotheses
+emerge from lively discussion in their respective fields on a very frequent
+basis.
 
 This section delineates the ground the software presented in this thesis is
-based on. \ref{sec:curt} introduces the base system of the implementation, \curt
+based on. \ref{sec:curt} introduces the base system of the implementation,
+\curt{}
 and \ref{sec:comparison} relates it to other similar systems in existence.
@@
 compared to other systems presented in \ref{sec:comparison} in terms of lexical
-and grammatical coverage, it provides a sound basis for extended work because of
+and grammatical coverage, it provides a sound basis for extended work due to
 its coherent design and relative simplicity.
@@
 incrementally, presenting new versions of \curt{} that add to the software's
-feature set. Support for external inferencing via theorem provers and model
-builders is added right after the first naïve implementations and
-finally an ontology and a straight-forward way of parsing and answering direct
+feature set one at a time. Support for inferencing via external theorem provers
+and model builders is added right after the first, rather limited implementations.
+Finally an ontology and a straight-forward way of parsing and answering direct
 questions for noun phrases conclude its development. The theory and
@@
 latter feature. This will also happen in an incremental manner and new features
-will be introduced one at a time.
+are introduced one at a time during section \ref{sec:implementation}.
 
@@
 
-Generally, question answering and semantic parsing are not related and seldom
-interdependent in current implementations of similar systems, with a few notable
-exceptions. As question answering becomes a more and more sophisticated task,
-the systems employed gather an increasing capability of \emph{understanding}
-their (input) domain. On the other side of the fence, interest in reasoning
-tools and inference engines operating on natural language have grown
-considerably over the past decade and some have stretched out to encompass more
-advanced domains of logic and discourse representation.
-
+Generally, software that concerns itself with answering questions formulated in
+ordinary human language falls in two main categories: \term{information
+retrieval systems}, also known as \abbr{QA} (question answering) systems, and
+\term{dialogue systems}.
+
+The purpose of the former is in most cases twofold, although present
+\abbr{QA} systems vary in the functionality they offer
+and results they actually deliver. One goal is to enable their users to
+query for information without the need for a special query language or at least
+using a query language that would be easy and natural to use even for the
+uninitiated user. This way the user can pose direct questions to a system
+instead of having to search for a keyword to occur in the plain text of a
+document, like they would have to do with traditional information
+retrieval systems. The second feature provides \abbr{QA} systems with another important
+distinction, viz. the ability to utilize ontologies in order to relate query
+terms akin, but not equal to words in target documents and thus deliver
+satisfactory results even for fuzzy queries.
+
+These systems typically do not try to \emph{reason} about
+their input, but gain `knowledge' from applying statistical methods on large
+input corpora. The methods used and results achieved vary throughout the
+literature. One technique employed is feeding a database with large sets of
+question/answer pairs, as described in \cite{rama} and then running basic
+natural language processing tasks like part-of-speech tagging,
+named-entity-recognition, syntactic and semantic parsing on the input.
+
+Many systems are only able to deliver small bits of information, also known as
+\term{factoids}, or just display fragments of their input text as it occurred in
+their corpus. Their knowledge base is typically not interactively modifiable,
+and acquiring knowledge and answering questions are two entirely separated
+tasks.
+
+\cite{prontoqa} introduce the \pn{Pronto QA System}, which couples traditional
+corpus-based \abbr{QA} techniques with deep semantic parsing based on
+\term{Discourse Representation Theory}. Unfortunately, there seems
+to be no working demo to date. The homepage of the ongoing efforts of Johan Bos
+and his team can be found at the following address:
+\href{http://svn.ask.it.usyd.edu.au/trac/candc/}{\url{http://svn.ask.it.usyd.edu.au/trac/candc/}}.
+
+Interestingly, earlier attempts at providing a software capable of answering
+questions in natural language where also mostly coupled with the usage of
+semantic analysis instead of statistical methods.
 \cite{chat} are among  the first to attempt creating a question answering system
-with deep semantic coverage. Their effort has been focused on the geographical
+with reasonable semantic coverage. Their effort has been focused on the geographical
 domain, and their implementation, dubbed \pn{CHAT-80}, is written entirely in
@@
 written with efficiency in mind. Moreover, it had to maintain a statical
-database of facts which was hand-crafted. The \curt{} system is built on  an
-entirely different set of premises and aims to provide a proof-of-concept of a
-more general purpose system that is also able to extend its knowledge
-dynamically.
-
-Soon after \pn{CHAT-80}, two entirely different types of systems emerged. The
-first type is probably the most promising to be used on a wider scale, given
-today's technological standards. Implementations are usually based on
-\term{information retrieval} and the generation of question/answer pairs. Such
-systems are generally referred to as a \abbr{QA} (\term{Question Answering})
-systems.  Examples implementations of such systems are discussed in \cite{rama},
-\cite{iesqa} and many others. Most of these seem to provide a good lexical and
-grammatical coverage due to their large input corpora and
-shallow interpretation techniques. The latter is also the \abbr{QA} technique's
-greatest shortcoming, as no deep analysis is  performed and usually
-only factoids of less then a couple of words in length can be supplied as
-answers.\footnote{But see \cite{aqabtf} for an approach that aims to cover a
-wider range of possible answers.}
-
-In contrast, \pn{CHAT-80} runs all its queries against a pre-built database
-without the ability to acquire new knowledge from the discourse. While a
-\abbr{QA} implementation will typically just acquire knowledge by consuming
-more input and analysing its statistical properties, programs employing deep
-semantic analysis will have to check their input for logical contradictions in
-to maintain their database in a reasonable\footnote{FIXME pun intended, but
-appropriate? %todo
-} state.
-
-\cite{prontoqa} introduce the \pn{Pronto QA System}, which couples traditional
-corpus-based \abbr{QA} techniques with deep semantic parsing based on
-\term{Discourse Representation Theory}. It is one of the few \abbr{QA} systems
-with a capability of reasoning about its input text. Unfortunately, there seems
-to be no working demo to date. The homepage of the ongoing efforts of Johan Bos
-and his team can be found at the following address:
-\href{http://svn.ask.it.usyd.edu.au/trac/candc/}{\url{http://svn.ask.it.usyd.edu.au/trac/candc/}}.
+database of facts which was hand-crafted. As such, it falls more into the
+category of question answering systems than dialogue systems.
+
+Today it seems that question answering systems make inroads to the search market,
+and some argue that they are bound to eventually supplant current document/term-based
+search engines. Even though \abbr{QA} systems also operate on the domain of
+answering questions, their approach to resolving them differs in most aspects
+from that of dialogue systems. The latter have a much higher obligation to
+correctness, since their results are typically not subject to peer
+review and server an entirely different purpose.
 
@@
 software driving interaction with the outer world. As such its conception of
-discourse not only relies on spoken dialogue, but also on sensory input.
+discourse not only relies on spoken dialogue but also also on sensory input, rather
+than on an input corpus. Its knowledge is also \emph{dynamic}, i.e. it can be
+extended through the aforementioned means, making it
+dialogue system.
 \pn{Godot} is focused on executing orders -- after evaluating their feasibility
@@
 of questions about its state.
+
+\curt{} is a dialogue system and thus based on an entirely different set of
+premises than \abbr{QA} systems. Moreover, it differs from \pn{Godot} in some
+major points, the most distinctive of which is that \curt{} only exists inside a
+regular computer and has no access to sensory input. Also, \curt's
+representation of knowledge differs from that of \pn{Godot} or the \pn{Pronto QA
+System} in that it does not use discourse representation theory (yet). \acurt{}
+addresses some of its shortcomings and lays ground to the implementation of a
+broader feature set for it.
 
@@
   An \term{interrogative} is a certain type of sentence in natural language
-  distinguished by certain characteristics\footnote{To quote \cite{gs:q}: \quote{[An
+  distinguished by certain characteristics
+  \footnote{To quote \cite{gs:q}: \quotesf{[An
   interrogative sentence is] characterized by word order, intonation, question
@@
 
-\begin{quote}A direct question can be treated as semantically equivalent to a
+\begin{quotes}A direct question can be treated as semantically equivalent to a
 certain kind of declarative sentence containing the corresponding indirect
-question embedded under  a suitable `performative' verb.\end{quote}
+question embedded under  a suitable `performative' verb.\end{quotes}
 
 Using this strategy, Karttunen escapes the treatment of direct  questions and
-thus gives an approach for indirect questions only.
-In this respect, his analysis is in certain aspects similar to the one presented in
+thus gives an approach for indirect (i.e. \emph{embedded}) questions  only.
+In this respect, his analysis is somewhat similar to the one presented in
 \cite{gs:sawhq}, but the latter goes on to challenge certain aspects of his
@@
 
-\begin{quote}
+\begin{quotes}
   Questions, unlike assertions, don't have truth-values, so it would be
   misleading to represent them as ordinary formulas.
-\end{quote}
-
-Their notion of questions will be extended for \acurt in order to account for
+\end{quotes}
+
+Their notion of questions will be extended for \acurt{} in order to account for
 yes/no interrogatives and larger sets of answers made possible by using model
@@
 
-Another important distinction in the definition of questions, is the definition
-of \term{answerhood}, or, what constitutes a valid answer to a given
-interrogative. The ground-laying work by \cite{hamblin:q} states that in knowing
-all possible answers to a question, one also knows the question itself. The
-notion of answerhood is going to be elaborated on in section \ref{sec:answerhood}
-
-One thing remains clear though, no matter what the notion of a question or
-finding an answer to it is; upon
-finding an answer, said answer has to be \emph{realized} in natural language
-using text or speech synthesis. When realizing an answer, many pragmatic
-considerations have to be taken into account.
-However, as the uttering of an embedded question does not prompt an
-answer\footnote{Unless, of course, the question is uttered as an
-\term{imperative} -- but this would pose an entirely different class of problem
-that is not discussed here.}, the present
-work will not directly concern itself with a solution to this problem.
-
-Constituent interrogatives also differ in what they ask for. One can assume the existence of at
-least three different types:
+Constituent interrogatives also differ in what they ask for. One can assume the
+existence of at least three different types:
 
@@
   \item NP-questions ask for a noun phrase, e.g. \example{Who shot Vincent?
-  Butch} or even more then one noun phrase at the same time \example{Who had
-  what in the hash bar?}
+  Butch (shot Vincent.)} or even more then one noun phrase at the same time, \example{Who had
+  what in the hash bar?\footnote{Multiple constituent interrogatives pose a
+  somewhat greater challenge, since they come with an inherent ambiguity, the
+  so-called \term{Baker Ambiguity}, as presented in \cite{baker:1970}.
+  Such ambiguities will not be concerned here.}}
   \item PP-questions ask for prepositional constituent, e.g. \example{Where did
@@
 This classification of interrogatives can go on for quite some more entries,
-including temporal, modal, causal and all sorts of other questions. Out of all
+including local, modal, causal and all sorts of other questions. Out of all
 those categories, only NP-questions will be supported, with an option to
-integrate other sorts of questions at a later stage.
+integrate other types of interrogatives at a later stage.
+
+Another important distinction in the definition of questions, is the definition
+of \term{answerhood}, i.e. what constitutes a valid answer to a given
+interrogative. The ground-laying work by \cite{hamblin:q} states that in knowing
+all possible answers to a question, one also knows the question itself. The
+notion of answerhood is going to be elaborated on in section \ref{sec:answerhood}
+
+One thing remains clear though, no matter what the notion of a question or
+finding an answer to it is; upon
+finding an answer, said answer has to be \emph{realized} in natural language
+using text or speech synthesis. When formulating an answer, many pragmatic
+considerations have to be taken into account.
+However, as the uttering of an embedded question does not prompt an
+answer\footnote{Unless, of course, the question is uttered as an
+\term{imperative} or \term{performative}-- but this is an entirely different
+class of problems that is not discussed here.}, the present
+work will not directly concern itself with a solution to this problem.
 
@@
 present, only addresses the problem of direct constituent interrogatives and
-does neither treat yes/no interrogatives nor embedded questions. \acurt treats
-embedded questions by embedding a special construct, a \term{proto-question},
-under verbs of certain types (which are going to be discussed in
-\ref{sec:ontology}). Furthermore, the yes/no interrogative is considered a
-special case of the constituent interrogative.
+does neither treat yes/no interrogatives nor embedded questions. The
+interrogative operator, as it will be introduced shortly, is designed to cover
+the semantics of both types of questions simultaneously. It takes a proposition
+(a formula yielding a truth value as a result) but gives back a set of
+propositions. Computation on more than one proposition in parallel will be
+introduced into the \curt{} systems in section \ref{sec:possworlds}.  \acurt{}
+represents embedded questions by embedding a special construct, a
+\term{proto-question}, under verbs of certain types (which will be discussed
+later on). Since the yes/no interrogative is considered a special case of the
+constituent interrogative the semantic formulae produced by the grammatical
+back end of \acurt{} do not differ for the two cases.
 
@@
     considered a question associated wit a yes/no interrogative.
-    \item The question operator does \emph{not combine} with any other operator.
+    \item \label{def:outermost}The question operator does \emph{not combine} with any other operator.
   \end{enumerate}
@@
 Note that the last clause of definition \ref{def:qop} enforces the question operator to
-always be the last operator of a formula of propositional logic. Embedded
-questions will thus not have to operate on the questions directly, but rather on
-their value (answers), or the set of propositions they denote.
-
-As was already mentioned, \pn{Helpful Curt} already has a certain albeit limited
+always be the last operator of a formula. This constraint will be overridden
+during parsing of questions to produce forms that do have nested questions.
+When calculating the models and readings of a formula, however, these operators
+will be resolved first. This is done in order to increase the efficiency during
+parsing, since the syntactic parser will sometimes have to backtrack.
+
+As was already mentioned, \pn{Helpful Curt} has a certain albeit limited
 support for evaluating NP constituent interrogatives. Blackburn and Bos adopt a
 sort of pseudo-notation of the question operator in what they call an
-extra-logical form which does not itself a yield a truth value, but is
+extra-logical form which does not itself yield a truth value, but is
 internally evaluated in a special case.
 
-They treat a question as a sort of \emph{quantification
+They treat a question as a \emph{quantification
 over a certain individual $X$, for which $R$ and $S$ must hold.} Thus a parse of
@@
 and knowledge. In this it differs substantially from the approaches made by
-Groenendijk and Stokhof and also later approaches by \cite{g:is}. \acurt will
-bridge the gap between the two.
+Groenendijk and Stokhof and also later approaches by \cite{g:is}. \acurt{}
+bridges the gap between both worlds.
 
@@
 
-from questions...
-                   Hamblin’s postulates read as follows:
-     (i) An answer to a question is a sentence, or statement.
-    (ii) The possible answers to a question form an exhaustive set of mutually
+Knowing the structure of an interrogative, the next step is to evaluate the
+question's \stress{value}, namely the propositions it generates. This will
+invariably bring up the question of the definition of \term{answerhood}, viz.
+the relation of the question to its answer or answers with respect to the
+current discourse. It is important to note that an answer other than `no'
+`nobody' cannot occur without a certain discourse or background knowledge already being present
+for the answer to be evaluated against.
+
+\cite{hamblin:q} defines a basic notion of answerhood, relating the value of
+a question with the set of propositions that might serve as partial answers to
+it. \cite{gs:q} phrase Hamblin’s postulates as follows:
+
+\begin{quotes}
+  \begin{enumerate}
+    \item An answer to a question is a sentence, or statement.
+    \item The possible answers to a question form an exhaustive set of mutually
     exclusive possibilities.
-   (iii) To know the meaning of a question is to know what counts as an answer to
+    \item To know the meaning of a question is to know what counts as an answer to
     that question.
-
-\subsubsection{The Syntax of Questions}
-
-With this ontology in mind, the next step is to analyse and enumerate the sorts
-of grammatic constructions that are to be treated. The following examples will
-illustrate them.
-
-\begin{itemize}
-  \item Alternative Questions
-  \ex. \a. Mia \emph{knows} whether Vincent snorts.
-  \b. * Mia \emph{believes} whether Vincent snorts.
-  \b. * Mia \emph{thinks} whether Vincent snorts.
-
-  \item Enumerative Questions
-  \ex. \a.  Mia \emph{knows} who snorts.
-  \b. * Mia \emph{believes} who snorts.
-  \b. * Mia \emph{thinks} who snorts.
-
-  \item Embedding Propositions
-  \ex. \label{ex:stipul}
-  \a. Mia \emph{believes} Vincent snorts.
-  \b. * Mia \emph{believes} who snorts.
-
-\end{itemize}
-
-An important observation to be made here is that only \emph{assertive} verbs can
-embed real questions. \emph{Stipulating} verbs only embed \emph{propositions} as
-in \ref{ex:stipul}.
-Nevertheless, both are treated in the following analysis, but with a focus on
-the treatment of assertive \wh-embedding verbs.
+  \end{enumerate}
+\end{quotes}
+
+For yes/no interrogatives this means, that if the question is $?.\varphi$, the
+answer two it is either $\varphi$ or $\neg\varphi$. For constituent
+interrogatives, this question is not so easily answered.  There is one central
+issue to be addressed first: the issue of the definition of answerhood: should
+the set of answers be an \emph{exhaustive} set of answers, including the
+negative ones, or just the set of possible \stress{true} answers?
+
+The following situation might serve as an example: Suppose Mia and Vincent both
+play air guitar. Butch and Marcellus do not play air guitar. When answering the
+question \example{Who plays air guitar?}, one can observe three degrees of
+exhaustiveness, according to \cite[section 1.5]{gs:sawhq}. Clearly, it does not
+suffice to just say \example{Vincent.} or \example{Mia.}, since that would still
+not account for all possible answers to the question, and that both
+\example{Butch.} and \example{Marcellus.} would be outright wrong. In the
+analysis of \cite{karttunen:1977}, every set of propositions that is a superset
+of the set of propositions containing \example{Mia plays air guitar.} and
+\example{Vincent plays air guitar.} would be a valid answer. This is the lowest
+degree of exhaustiveness, and accounts for all \emph{true} answers. So, if
+Yolanda believed to not only know of Vincent and Mia that they played air
+guitar, but also of Butch, she would still know all possible true answers, and,
+according to this analysis, a correct answer.
+
+Groenendijk and Stokhof argue for two more degrees of exhaustiveness. The first
+one would require the answer not only to contain all possible \emph{true}
+propositions, but also to \emph{not contain any \stress{false} propositions}. So
+in this case, if Yolanda believed Butch played air guitar, she would not know
+the true answer.
+
+The second and strongest degree of exhaustiveness would
+require any set of answers not only to contain the set of propositions that
+hold, but also the set of propositions that \emph{do not hold}. In this case, for
+the sentence \example{Yolanda knows who plays air guitar.} to be true, the
+sentence \example{Yolanda knows who does not play air guitar.} would
+automatically follow.
+
+There even more ways of categorizing the number of answers to a question. So far
+the so called \term{mention-all} case has been discussed, but when considering
+a sentence like \example{John knows where to buy a Chinese newspaper.} (example
+due to \cite{rooy:qar}), it is clear that only one or two out of all possible
+propositions for \example{Where to buy a Chinese newspaper?} will be sufficient
+for John to know in order for him to know where to actually buy one. This has
+been regarded as the \term{mention-some} reading.
+
+Even after having settled on the \emph{number} and \emph{value} of propositions
+that have to serve as an answer to a question, one still might be concerned with
+their kind. Taken the previously given example, if Butch and Marcellus were
+all boxers in this situation, would not an answer such as \example{Every boxer.}
+suffice? \cite{rooy:qar} elaborates on the many aspects of different analyses
+regarding the relevance of answers. Suffice to say, \acurt{} is only content
+with \emph{maximally specific} resolutions to answers.
+
+Moreover, as \acurt{} will only have to deal with answers to questions
+internally, it will also only regard maximally exhaustive interpretations of
+answers.
+Correctly answering a direct question with respect to relevance and other
+pragmatic factors is a task left as a task for turning \curt{} into a complete
+dialogue system more potent of dealing with queries to its knowledge.
 
@@
 
-When treating questions one will be confronted with many different
-categorizations of questions. One of the first distinctions to make is the one
-between \term{embedded questions} as in \ref{ex:embed} and \term{direct
-questions} shown in \ref{ex:direct}.
-
-\ex. \a. John knows \emph{whether this sentence contains an embedded
-question}.\label{ex:embed}
-\b. Does this sentence pose a direct question?\label{ex:direct}
-
-As indicated already by the title, this paper will concern itself with the
-treatment of the former and only briefly comment on the latter. The reasons for
-this are numerous  %fixme I need a synonym for that
-and this section is going to name the most important ones.
-
-Early literature, such as \cite{karttunen:1977} only regards embedded
-\wh-phrases, treating direct questions as semantically equivalent. To quote
-Karttunen:
+After having agreed on a treatment of interrogatives, the next step is to
+construct a framework for embedding the result under certain verbs. The main
+issues of this section will thus be focused on providing a classification of
+embedding verbs and elaborating on the details of the relationship between
+embedded and embedding constituents of an utterance.
+
+As was already mentioned, clause \ref{def:outermost} of definition \ref{def:qop}
+will be overridden when treating embedded interrogatives. The following example
+illustrates a syntactic parse of an embedded interrogative:
+
+\ex. \label{ex:whembed}
+\a. Mia knows who snorts.\footnotesize
+\b.  \Tree
+[.$t$\\$know(?X_1.snort(X_1))(Mia)\wedge ?X_1.snort(X_1)$ [.$T$ { $\lambda X.X(Mia)$\\Mia } ]
+[.$IV$\\$\lambda X.know(?X_1.snort(X_1))(X)\wedge ?X_1.snort(X_1)$ [.{$IV/\bar{t}$} 
+{ $\lambda S.\lambda X.[  \lambda Y.[ know(Y)(X) \wedge Y ](S) ]$\\knows } ]
+[.{$\bar{t}$\\$?X_1.snort(X_1)$ } [.$\bar{t}/t$ ? ] [.{ $t$\\$snort(X_1)$ }
+[.$T$ { $X_1$\\who } ]
+[.$IV$ { $\lambda X.snort(X)$\\snorts } ] ] ] ]!\qsetw{4cm} ]\normalsize
+\b. $know(?X_1.snort(X_1))(Mia)\wedge ?X_1.snort(X_1)$\label{ex:formula}
+
+Note that this example is using  the (syntactic) type system of \cite{gs:sawhq}
+but different meaning postulates. Grammar rules, later explained in
+\ref{sec:curtgrammar}, are
+responsible for determining the correct order of functional application in the
+untyped logical system \acurt{} is using.
+
+One can see \ref{ex:formula} as being an \emph{intensional} denotation of the
+meaning of \ref{ex:whembed} rather than an extensional one. In fact, this
+sentence will evaluate to more than one proposition in case there is more than one
+person snorting.
+
+The reason for $?$ to be embeddable within more complex formulae is merely one
+of convenience, since one could transform $?X.\varphi(X)$ to mean
+
+\[\forall w \in W_M [ \forall x \in D_w [ \varphi(w)(x) = 1 ] ]\]
+
+assuming $W_M$ is the set of possible worlds in a model $M$, but this would
+require modifications to the current way of evaluating predicates with \curt{}
+as well as the definition of a quantification over possible worlds, which is
+currently not supported and would require a special treatment. Furthermore, in
+the case that the explicit formula does not yield any results, the question
+should return its correct answer
+
+\[ \neg\exists x [\varphi(x)] \]
+
+Instead, before calling its inferencing tools \acurt{} will look out only for
+the question operator itself and take appropriate measures.
+
+Another advantage of this approach is that the treatment for yes/no
+interrogatives and constituent interrogatives will not require special
+intervention at the level of grammar.
+
+\ex. \label{ex:simpleembed}
+\a. Mia knows whether Vincent snorts.  \footnotesize
+\b. \Tree
+[.$t$\\$know(?.snort(Vincent))(Mia)\wedge ?.snort(Vincent)$ [.$T$ { $\lambda X.X(Mia)$\\Mia } ]
+[.$IV$\\$\lambda X.know(?.snort(Vincent))(X)\wedge ?.snort(Vincent)$ [.{$IV/\bar{t}$} 
+{ $\lambda S.\lambda X.[  \lambda Y.[ know(Y)(X) \wedge Y ](S) ]$\\knows } ]
+[.{$\bar{t}$\\$?.snort(Vincent)$} [.$\bar{t}/t$ whether ] [.{ $t$\\$snort(Vincent)$ } 
+[.$T$ { $\lambda X.X(Vincent)$\\Vincent } ]
+[.$IV$ { $\lambda X.snort(X)$\\snorts } ] ] ] ]!\qsetw{5cm} ]
+\normalsize
+\b. $know(?.snort(Vincent))(Mia)\wedge ?.snort(Vincent)$
+
+The question operator in front of saturated formulae ($?.\varphi$) could also be
+rephrased as
+
+\[ \forall w\in W_M [ \varphi(w) \wedge \neg\varphi(w) \]
+
+This would again demand explicit quantification over worlds and a different
+treatment of predicates. Instead, it can be treated with the same procedure as
+before when using an embedded question operator.
 
 \subsubsection{Types of Facts, Assertions and Interrogatives}
+
+Embedded clauses are always embedded under a certain verb, which usually
+takes a certain type of phrase and returns an intransitive verb. Examples of
+this were given in the previous section was in \ref{ex:simpleembed} and
+\ref{ex:whembed}. The verb used in both example was \example{know}. The meaning
+postulate for \example{know} is the following:
+
+\begin{postulate}\label{def:know}
+  Knowledge necessitates the belief of a proposition $p$ by the knowing
+  individual as well as the verity of $p$.
+
+  \sem{know} = $\lambda p\lambda y.[know(y)(p)\wedge p]$
+\end{postulate}
+
+While the argumentation about exhaustiveness in section \ref{sec:answerhood}
+already drew upon the this statement, postulate \ref{def:know} puts this into a
+more formal perspective. The postulate is based on the fact that, clearly, one
+can not \emph{know} something that is not true. In that case one's knowledge
+would be wrong, and one would only \emph{believe to know}.
+
+The proposition $snorts(Vincent)$, which would evaluate to a truth value, is
+turned into a \wh-construct by \example{whether} -- the exact circumstances of
+this will be explained further on.
 
@@
 
+With this ontology in mind, the next step is to analyse and enumerate the sorts
+of grammatic constructions that are to be treated. The following examples will
+illustrate them.
+
+\begin{itemize}
+  \item Alternative Questions
+  \ex. \a. Mia \emph{knows} whether Vincent snorts.
+  \b. * Mia \emph{believes} whether Vincent snorts.
+  \b. * Mia \emph{thinks} whether Vincent snorts.
+
+  \item Enumerative Questions
+  \ex. \a.  Mia \emph{knows} who snorts.
+  \b. * Mia \emph{believes} who snorts.
+  \b. * Mia \emph{thinks} who snorts.
+
+  \item Embedding Propositions
+  \ex. \label{ex:stipul}
+  \a. Mia \emph{believes} Vincent snorts.
+  \b. * Mia \emph{believes} who snorts.
+
+\end{itemize}
+
+An important observation to be made here is that only \emph{assertive} verbs can
+embed real questions. \emph{Stipulating} verbs only embed \emph{propositions} as
+in \ref{ex:stipul}.
+Nevertheless, both are treated in the following analysis, but with a focus on
+the treatment of assertive \wh-embedding verbs.
 \subsubsection{Modeling Knowledge}\label{sec:epistemic}
@@
 
+When treating questions one will be confronted with many different
+categorizations of questions. One of the first distinctions to make is the one
+between \term{embedded questions} as in \ref{ex:embed} and \term{direct
+questions} shown in \ref{ex:direct}.
+
+As indicated already by the title, this paper will concern itself with the
+treatment of the former and only briefly comment on the latter. The reasons for
+this are numerous  %fixme I need a synonym for that
+and this section is going to name the most important ones.
+
+Early literature, such as \cite{karttunen:1977} only regards embedded
+\wh-phrases, treating direct questions as semantically equivalent. To quote
+Karttunen:
+
+\ex. \a. John knows \emph{whether this sentence contains an embedded
+question}.\label{ex:embed}
+\b. Does this sentence pose a direct question?\label{ex:direct}
+
 The present work will rely on \cite{gs:q} for providing the main guidelines for
@@
 \pn{CHAT-80} is.
+
+\subsubsection{The Syntax of Questions}
+
+The internal structure of the subset of interrogative sentences that will be
+considered here is not a very complicated one. Since yes/no interrogatives  embed
+simple propositions under a question operator only minimal modifications will
+need to be made to the grammar in order to account for the latter. Constituent
+interrogatives were already covered by \pn{Helpful Curt} and their syntactic
+treatment does not differ in \acurt{}.
+
+However, the syntax of embedded interrogatives is a more delicate matter.
 
@@
 use.
-
-\ex. \label{ex:kel1} Mia knows whether Vincent snorts.\\
-\Tree
-[.$t$ [.$T$ { $\lambda X.X(Mia)$\\Mia } ]
-[.$IV$ [.{$IV/\bar{t}$} { $\lambda X.know(X)$\\knows } ]
-[.{$\bar{t}$} [.$\bar{t}/t$ whether ] [.{ $t$\\$snort(Vincent)$ } 
-[.$T$ { $\lambda X.X(Vincent)$\\Vincent } ]
-[.$IV$ { $\lambda X.snort(X)$\\snorts } ] ] ] ] ]
 

### 2008-08-29 thesis/paper.tex
Some of the final work on 3

@@
 \setsansfont [Ligatures={Common}, BoldFont={Fontin Sans Bold}, ItalicFont={Fontin Sans Italic}]{Fontin Sans}
-\setmonofont [Ligatures={NoCommon,NoRequired,NoRare}] {Monaco}
+%\setmonofont [Ligatures={NoCommon,NoRequired,NoRare}] {Monaco}
 
@@
 \newcommand{\term}[1]{\textsf{\textbf{#1}}} % introduce a term
-\newcommand{\code}[1]{\footnotesize\texttt{#1}\normalsize} % layout of in-line code snippets
+\newcommand{\code}[1]{\texttt{#1}} % layout of in-line code snippets
 \newcommand{\pn}{\textsf} % how to render proper names like Shrdlu
 \newcommand{\url}[1]{\code{http://#1}} % how urls appear
-\newcommand{\example}{\textit} % in-text examples
+\newcommand{\example}[1]{`\textit{#1}'} % in-text examples
 
@@
 
-\subsubsection{Types of Facts, Assertions and Interrogatives}
+\subsubsection{Types of Facts, Assertions and Interrogatives, Towards an
+Ontology of Embedding}\label{sec:ontology}
 
@@
 takes a certain type of phrase and returns an intransitive verb. Examples of
-this were given in the previous section was in \ref{ex:simpleembed} and
-\ref{ex:whembed}. The verb used in both example was \example{know}. The meaning
-postulate for \example{know} is the following:
+this were given in the previous section in \ref{ex:simpleembed} and
+\ref{ex:whembed}. The verb used in both examples was \example{know}.
+
+Examining some examples of other verbs, one will quickly establish that not
+every verb can embed any kind of sentence, but also that there seems to be  a
+pattern one might be able to exploit. Leaving the question operator in place
+again helps distinguishing between the different types of embedding in such
+cases, since it is of major importance to distinguish between indicatives and
+interrogatives, as we can see from the next example.
+
+\begin{itemize}
+  \item Embedded yes/no interrogative
+  \ex. \a. Mia \emph{knows} whether Vincent snorts.
+  \b. * Mia \emph{believes} whether Vincent snorts.
+  \b. Mia \emph{asks} whether Vincent snorts.
+
+  \item Embedded constituent interrogative
+  \ex. \a.  Mia \emph{knows} who snorts.
+  \b. * Mia \emph{believes} who snorts.
+  \b. Mia \emph{asks} who snorts.
+
+  \item Embedded proposition
+  \ex. \a. Mia \emph{knows} Vincent snorts.
+  \b. Mia \emph{believes} Vincent snorts.
+  \b. * Mia \emph{asks} Vincent snorts.
+
+\end{itemize}
+
+From these examples it seems that, again, there is no difference between
+constituent interrogatives and yes/no interrogatives concerning the position
+they can occupy. Furthermore, it seems that the three different verbs behave
+differently: `know' can take both questions and propositions, `ask' can only
+take questions and `believe' can only take propositions.
+
+\cite{karttunen:1977} introduces means of classification for \wh-embedding
+verbs, the nature of which \cite{lahiri:diss} covers in a much broader scope.
+Both come to the conclusion that different verb classes ask for different
+strategies of treating the \wh-clauses they embed and also pose different
+restrictions on the type of clause they expect. The present work will mostly
+draw from the results of \cite{ginzburg}, who summarizes previous efforts,
+in order to provide an account for an ontology of embedding verbs.
+
+According to \cite{ginzburg} there are (at least) three different types of
+embedding verbs\footnote{The nomenclature of the different types departs from
+what Ginzburg presents. He calls Questioning predicates QI, propositional
+predicates TF. Resolutive and factive predicates are the same.}:
+
+\begin{definition}\label{def:embed1}
+  Types of Embedding Predicates
+\begin{itemize}
+  \item \term{Questioning predicates}, take only questions as counterparts.
+  Examples: \example{ask}, \example{wonder}, \example{investigate},
+  \example{discuss}
+  \item \term{Propositional predicates}, need propositions as counterparts.
+  Examples: \example{claim}, \example{believe}, \example{think},
+  \example{assume}, \example{deny}
+  \item \term{Factive predicates} take both, questions and propositions.
+  Examples:
+  \example{know}, \example{forget}, \example{reveal}, \example{discover},
+  \example{report}, \example{predict}
+\end{itemize}
+\end{definition}
+
+Both, \term{questioning predicates} and \term{propositional predicates} do not
+effect any change or demand any particular state in the current situation of the
+world in order to be used. One can ask \example{Who dances?} without anybody
+present being in the state of dancing and one can believe someone to be in the
+state of dancing, without this actually being the case. They are clear in what
+they take as a counterpart, too: one cannot \emph{claim who dances} and one
+cannot \emph{ask every boxer dances}.
+
+The matter is more complicated, however, with \term{factive predicates}, where
+one might wish to establish another distinction. The example
+\ref{ex:whembed} already used a denotation for \example{know}, which will now be
+introduced formally:
 
@@
 While the argumentation about exhaustiveness in section \ref{sec:answerhood}
-already drew upon the this statement, postulate \ref{def:know} puts this into a
+already drew upon this statement, postulate \ref{def:know} puts this into a
 more formal perspective. The postulate is based on the fact that, clearly, one
 can not \emph{know} something that is not true. In that case one's knowledge
-would be wrong, and one would only \emph{believe to know}.
-
-The proposition $snorts(Vincent)$, which would evaluate to a truth value, is
-turned into a \wh-construct by \example{whether} -- the exact circumstances of
-this will be explained further on.
-
-\subsubsection{Towards an Ontology of Embedding Verbs}\label{sec:ontology}
-
-\cite{karttunen:1977} introduces means of classification for \wh-embedding
-verbs, the nature of which \cite{lahiri:diss} covers in a much broader scope.
-The different verb classes ask for different strategies of treating the
-\wh-clauses they embed and also pose different restrictions on the type of
-clause they expect.
-
-The present work will not consider a full analysis of \wh-embedding verbs, but
-confine itself on a few ones, namely \example{know}, \example{think} and
-\example{believe}. The grammar implementation of \acurt{} allows for \wh-embedding
-verbs to have a \code{type} field, determining their semantic denotation. The two
-types used so far are \code{assert} for assertive verbs such as \example{know}
-and \code{stipul} for stipulating verbs, such as \example{believe} and
-\example{think}.
-
-We first derive know(john,all(X,imp(woman(X),love(X,john)))) from the syntax of
-john knows whether every woman loves john, then we see if
-all(X,imp(woman(X),love(X,john))) is valid (uninformative) and ridicule John, if
-it's not valid and not consistent, assert his knowledge of the fact that it's
-not true. If it's true, we assert he knows that.
-
-For whether we can ridicule him if it's not true (inconsistent).
-The MB/TP do a comparison \Disc to all(X,imp(woman(X),love(X,john)))
-
-With this ontology in mind, the next step is to analyse and enumerate the sorts
-of grammatic constructions that are to be treated. The following examples will
-illustrate them.
-
+would be wrong, and one would only \emph{believe to know}. This also holds for
+verbs like \example{discover}, \example{reveal} and \example{forget}. But does
+it also hold for \example{report} and \example{predict}? Although both verbs are
+in the same \emph{syntactic} category, since one can \example{predict/report who will
+win/wins the elections}, they seem to demand different \emph{syntactic}
+treatment, since both, predictions on, and reports of the results of elections are
+in no way guaranteed to be true.
+
+So, another distinction has to be made. Restating the last clause of definition
+\ref{def:embed1}:
+
+\begin{definition}
+  Resolutive and factive predicates:
   \begin{itemize}
-  \item Alternative Questions
-  \ex. \a. Mia \emph{knows} whether Vincent snorts.
-  \b. * Mia \emph{believes} whether Vincent snorts.
-  \b. * Mia \emph{thinks} whether Vincent snorts.
-
-  \item Enumerative Questions
-  \ex. \a.  Mia \emph{knows} who snorts.
-  \b. * Mia \emph{believes} who snorts.
-  \b. * Mia \emph{thinks} who snorts.
-
-  \item Embedding Propositions
-  \ex. \label{ex:stipul}
-  \a. Mia \emph{believes} Vincent snorts.
-  \b. * Mia \emph{believes} who snorts.
-
+  \item \term{Factive predicates} take both, questions and propositions, and
+  demand the extension of (the set of) propositions they embed to evaluate to true (exhaustively).
+  Examples:
+  \example{know}, \example{forget}, \example{reveal}, \example{discover},
+  \item \term{Resolutive predicates} take both, questions and propositions, and
+  do not demand the extension of (set of propositions) they embed to have any particular truth value.
+  Examples:
+  \example{report}, \example{predict}
   \end{itemize}
-
-An important observation to be made here is that only \emph{assertive} verbs can
-embed real questions. \emph{Stipulating} verbs only embed \emph{propositions} as
-in \ref{ex:stipul}.
-Nevertheless, both are treated in the following analysis, but with a focus on
-the treatment of assertive \wh-embedding verbs.
+\end{definition}
+
+Yet still, all is not solved. The following example, taken from \cite[example
+(27)]{ginzburg} illustrates a flaw in the definition of factive predicates:
+
+\ex. \label{ex:ginzburg}
+\a. Jill believed a certain hypothesis. Hence, Jill believed that hypothesis
+to be true.
+\b. Jill discovered a certain hypothesis. It \emph{does not follow} that Jill
+discovered that hypothesis to be true. \cite{ginzburg} defends an analysis that
+introduces another distinction in the nature of indicatives: the distinction
+between \term{facts} and \term{propositions} -- where in \ref{ex:ginzburg} the
+embedded clause is treated as a proposition and not as a fact.
+
+However, since \acurt{} is only concerned with embedding \wh-sentences, and
+because it would exceed the scope of this thesis to introduce another type of
+indicative, only the four already explained types of embedding verbs are 
+considered here and a modification of the notion of indicatives will be left for
+further research.
+
 \subsubsection{Modeling Knowledge}\label{sec:epistemic}
 
+TODO Discussion of Lewis' paper.
+
 \subsection{Treating Disjunction}
+
+Consider the following example dialogue with the \pn{Helpful Curt}
+system\footnote{In a real conversation with \curt, it will usually inform the
+user about the activity of external inferencing tools. This has been omitted and
+marked by \code{[…]}.}:
+
+\begin{verbatim}
+  > Mia or Jody dances.
+  […]
+  Curt: OK.
+
+  > Who dances?
+  […]
+  Curt: This question makes sense!
+
+  Curt: maybe mia
+\end{verbatim}
+
+Clearly, \curt{} should give a full answer in this situation, but instead it
+just guesses, and does not reply with something along the lines of \example{Either
+Mia or Jody}. Unfortunately, this result does not suffice for building a system
+that can properly treat embedded questions, since it does not reflect all
+available information about the discourse properly.
+
+A solution to this will again require the presence of the concept of
+\term{possible worlds}. When uttering the phrase \example{Mia or Jody dances.}
+then this can be understood as creating \emph{two possible scenarios}, one in
+which there is an individual named \example{Mia}, and this individual dances,
+and one in which there is an individual named \example{Jody} and the same holds
+for this individual. In effect, this can be translated to two possible worlds
+being present in the current discourse.
+
+\begin{definition}
+  Disjunction
+
+  For any formula $\rho$ holding a disjunction, this formula can be expressed as a set
+  of the constituents of the disjunction
+
+  \[ \rho = \varphi \vee \psi, \textrm{ then } \rho' = \set{\varphi, \psi}\]
+
+  Every other operator or function can take a set of propositions of the type of
+  $\rho'$ to produce a set of propositions containing \emph{all possible
+  permutations} of $\rho'$. For every possible operator $\sem{$\cdot$}$ with arity
+  $n$ and arguments $\sem{$\cdot$}(\set{\varphi,\psi})$
+
+  todo: more generic: just go from 1 to j, which is rho, then from j+1 to n
+
+  \begin{eqnarray*}
+    \sem{$\cdot$}(\rho') \mapsto \set{\varphi,\psi}\\
+    \sem{$\cdot$}(\sigma_1,\dots,\sigma_j,\rho )
+
+    \sigma \wedge \set{\varphi, \psi} \mapsto \set{\sigma\wedge\varphi,\sigma\wedge\psi}\\
+    \sigma \to \set{\varphi, \psi} \mapsto \set{\sigma\to\varphi,\sigma\to\psi}\\
+  \end{eqnarray*}
+\end{definition}
 
@@
 
+We first derive know(john,all(X,imp(woman(X),love(X,john)))) from the syntax of
+john knows whether every woman loves john, then we see if
+all(X,imp(woman(X),love(X,john))) is valid (uninformative) and ridicule John, if
+it's not valid and not consistent, assert his knowledge of the fact that it's
+not true. If it's true, we assert he knows that.
+
+For whether we can ridicule him if it's not true (inconsistent).
+The MB/TP do a comparison \Disc to all(X,imp(woman(X),love(X,john)))
+
+An important observation to be made here is that only \emph{assertive} verbs can
+embed real questions. \emph{Stipulating} verbs only embed \emph{propositions} as
+in \ref{ex:stipul}.
+Nevertheless, both are treated in the following analysis, but with a focus on
+the treatment of assertive \wh-embedding verbs.
+
 When treating questions one will be confronted with many different

### 2008-08-29 thesis/paper.tex
Some corrections to the the first two sections.

@@
 quite some scope. Therefore most approaches have been focused on providing
-coverage of a small fragment of natural language often with respect to a
+coverage of a small fragment of natural language, often with respect to a
 particular domain. Many such systems exist and they prove to be increasingly
-successful. Some of them will be discussed here and presenting an extension to
+successful. Some of them will be discussed here, and presenting an extension to
 one system in particular is the main matter of this Bachelor's thesis.
@@
 partition of its logical domain, viz. the ability to handle more than one
-\term{possible} or \term{accessible world}. As a ground for development of this
+\term{possible} or \term{accessible world}. As a ground for the development of this
 work, the treatment of \term{embedded questions} was chosen as one that
@@
 literature due to recent progress in different areas.  Current development in
-computer hardware has made it possible to focus on theoretical soundness and
-important abstract designs and concepts of logic and the theory of meaning
+computer hardware has made it possible to focus on theoretical soundness,
+important abstract designs, and concepts of logic and the theory of meaning,
 rather than being able to simply implement a system that is able to do basic
@@
 computer science and linguistics has fueled new ideas and intriguing hypotheses
-emerge from lively discussion in their respective fields on a very frequent
+emerge from lively discussions in their respective fields on a very frequent
 basis.
@@
 based on. \ref{sec:curt} introduces the base system of the implementation,
-\curt{}
+\curt{},
 and \ref{sec:comparison} relates it to other similar systems in existence.
@@
 questions for noun phrases conclude its development. The theory and
-implementation of this paper is meant as a replacement and enhancement for the
-latter feature. This will also happen in an incremental manner and new features
+implementation of this paper are meant as a replacement and enhancement for the
+latter feature. Their introduction will also happen in an incremental manner and new features
 are introduced one at a time during section \ref{sec:implementation}.
@@
 Interestingly, earlier attempts at providing a software capable of answering
-questions in natural language where also mostly coupled with the usage of
+questions in natural language were mostly coupled with the usage of
 semantic analysis instead of statistical methods.
 \cite{chat} are among  the first to attempt creating a question answering system
-with reasonable semantic coverage. Their effort has been focused on the geographical
+with reasonable coverage. Their effort has been focused on the geographical
 domain, and their implementation, dubbed \pn{CHAT-80}, is written entirely in
@@
 
-Today it seems that question answering systems make inroads to the search market,
+Today it seems that statistical question answering systems make inroads to the search market,
 and some argue that they are bound to eventually supplant current document/term-based
@@
 correctness, since their results are typically not subject to peer
-review and server an entirely different purpose.
+review and serve an entirely different purpose.
 
@@
 
-\curt{} is a dialogue system and thus based on an entirely different set of
+\curt{} is a prototype dialogue system and as such it is based on an entirely different set of
 premises than \abbr{QA} systems. Moreover, it differs from \pn{Godot} in some
@@
 
-In the course of this thesis, a new version of \curt{} will be developed, called
-\acurt. This version is not only aware of its own state, but can also gather
-information about the individuals it knows about. Furthermore, \acurt{} is able
-to account for several possible readings of a certain sentence at once and in
-parallel, which is a side-effect of the introduction of the notion of
+In the course of this thesis, a new version of \curt{} is introduced, called
+\acurt. This version is not only aware of its own state of knowledge, as was the
+original \curt{} implementation, but can also gather information about the
+knowledge and beliefs of individuals it knows about. Furthermore, \acurt{} is
+able to account for several possible readings of a certain sentence at once and
+in parallel, which is a side-effect of the introduction of the notion of
 \term{possible worlds} into the system.
@@
 like `know' and `believe' and its inferencing mechanisms are going to be
-employed in a different manner. %todo sucks.
+employed in a different manner, to encompass a wider range of functionality.
 
@@
 
-Question semantics has traditionally been a very sophisticated field of
-linguistics and philosophy, with many different approaches arguing for and
-against each other. This section will not try to give an exhausting overview of
-the field of question semantics since this would be a task of considerable size
-and by far beyond the scope of this thesis\footnote{For a good summary of some
-of the literature, particularly the issue of whether questions are a concern of
-pragmatics or semantics, consider \cite[section 2 and 3]{gs:q}.}. Instead, it
-was chosen to roughly follow the work by Groenendijk and Stokhof, lending from
-other places where deemed appropriate or practical. As such, the theoretical
-framework of \acurt{} is not going to be an exact realization of any particular
-approach to question semantics, but rather an accumulation of interesting
-concepts from different sources. The main sources, however, are \cite{gs:q} and
-\cite{gs:sqpa}.
+The semantics of questions has traditionally been a very sophisticated field of
+linguistics,  philosophy and logics, with many different approaches arguing for
+and against each other. This section will not try to give an exhausting overview
+of the field of question semantics since this would be a task of considerable
+size and by far beyond the scope of this thesis\footnote{For a good summary of
+some of the literature, particularly the issue of whether questions are a
+concern of pragmatics or semantics, consider \cite[section 2 and 3]{gs:q}.}.
+Instead, it was chosen to roughly follow the work by Groenendijk and Stokhof,
+who present one of the best-established and recognized theories in the field,
+and also lending from other places where deemed appropriate or practical. As
+such, the theoretical framework of \acurt{} is not going to be an exact
+realization of any particular approach to question semantics, but rather an
+accumulation of interesting concepts from different sources. The main sources,
+however, are \cite{gs:q} and \cite{gs:sqpa}.
 
@@
 able to reason about possible worlds, employing first-order logic inferencing
-tools. % really? fix your title then, would you? … doh. fixme
+tools in order to realize an approach to the treatment of embedded questions and
+questions in general.
 
@@
 
-Following \cite{gs:q}, we will not use the English word
-``question'' in its traditional sense, but distinguish between \term{questions},
+Following \cite{gs:q}, the English word ``question'' is not used in its
+traditional sense, but a distinction is made between \term{questions},
 \term{interrogative sentences} and \term{interrogative acts} in the following
@@
 \begin{termin}
-  An \term{interrogative} is a certain type of sentence in natural language
+  An \term{interrogative sentence}, or simply \term{interrogative}, is a
+  type of sentence in natural language
   distinguished by certain characteristics
@@
   interrogative sentence is] characterized by word order, intonation, question
-  mark, the occurrence of interrogative pronouns}.}
+  mark, the occurrence of interrogative pronouns}.},
   whereas an \term{interrogative act} refers to the utterance of an
@@
 discussion and \cite{gs:sqpa} for its usage in describing the semantics of
-questions}. \abbr{IL} stands for intensional logic\footnote{As used by
+questions}, \abbr{IL} stands for intensional logic\footnote{As used by
 \cite{ptq}. See also \cite{gamut:2} for an excellent introduction to \abbr{IL}.},
-\abbr{FOL}  is used as an abbreviation for first order logic.
+\abbr{FOL}  is used as an abbreviation for first order logic. As an alternative
+term for \term{possible world}, sometimes the expression \term{index} is used.
 
@@
 \cite{ptq} and thus also makes use of its underlying formalism,
-\abbr{IL}. \cite{gs:sqpa} however use \abbr{Ty2} as it was
+\abbr{IL}. \cite{gs:sqpa}, however, use \abbr{Ty2} as it was
 proposed in \cite{gallin:ty2}. \cite{z:ilty2} discusses the differences and
@@
 As was already mentioned in \ref{sec:curt}, \curt{} is based on 
-the use of external tools only capable of treating first order logic – an
+the use of external tools. These tools are only capable of treating first order logic – an
 extensional framework. \prol{} on the other hand, is a language that can very well
@@
 of them would be beyond the scope of this thesis.  It was therefore decided to
-avoid modal logic and implement a pragmatic approach to representing the indices
-of higher-order logic.
-
-Some research on the logic of questions (eroteric logic) suggests that the
-treatment of interrogatives demands its own formalism.  \acurt{} employs a
-formalism that departs from both \abbr{IL} and \abbr{Ty2} in some ways. Most
-notably, the notion of a \term{model} is changed as follows, in order to
-accommodate the question operator introduced further on, which demands a set of
-possible worlds to be available.
+avoid modal logic and implement a pragmatic approach to representing indices in
+order to increase the expressive power of the system.
+
+Some research on the logic of questions (sometimes also referred to as eroteric
+logic) suggests that the treatment of interrogatives demands its own formalism.
+\acurt{} employs a formalism that departs from both \abbr{IL} and \abbr{Ty2} in
+some ways. Most notably, the notion of a \term{model} is changed as follows, in
+order to accommodate the question operator introduced further on, which demands
+a set of possible worlds to be available.
 
@@
   \begin{enumerate}
-    \item A model is a set of pairs of \term{indices} $i$ and \term{possible
+    \item A model $M$ is a set of pairs of \term{indices} $i$ and \term{possible
     worlds} $w$: $\pair{i,w} \in M$. 
-    \item A world $w$ at index $i$, $\pair{i,w} \in M$ is a triple of $D$, the set
+    \item A world $w$ at index $i$, $\pair{i,w} \in M$, is a triple of $D$, the set
     of individuals, $F$, the set of functions, and $\varphi$ a first-order
@@
     \item If $W \in M$, then we say $\exists W$ quantifies over the existence of
-    one member of $M$, namely $|M| \geq 1$ and $W \in M$.
+    one member of $M$, so $|M| \geq 1$ and $W \in M$.
   \end{enumerate}
@@
 
-One of the more important differences to the `traditional' notion of a model, is
+One of the more important differences to the `traditional', \abbr{FOL}, notion of a model, is
 that it was decided to assign every world a unique domain of discourse referents
@@
 \b. Jill discovered a certain hypothesis. It \emph{does not follow} that Jill
-discovered that hypothesis to be true. \cite{ginzburg} defends an analysis that
+discovered that hypothesis to be true.
+
+\cite{ginzburg} defends an analysis that
 introduces another distinction in the nature of indicatives: the distinction
@@
 
-  Every other operator or function can take a set of propositions of the type of
-  $\rho'$ to produce a set of propositions containing \emph{all possible
-  permutations} of $\rho'$. For every possible operator $\sem{$\cdot$}$ with arity
-  $n$ and arguments $\sem{$\cdot$}(\set{\varphi,\psi})$
+  %Every other operator or function can take a set of propositions of the type of
+  %$\rho'$ to produce a set of propositions containing \emph{all possible
+  %permutations} of $\rho'$. For every possible operator $\sem{$\cdot$}$ with arity
+  %$n$ and arguments $\sem{$\cdot$}(\set{\varphi,\psi})$
 
@@
   \begin{eqnarray*}
-    \sem{$\cdot$}(\rho') \mapsto \set{\varphi,\psi}\\
-    \sem{$\cdot$}(\sigma_1,\dots,\sigma_j,\rho )
-
-    \sigma \wedge \set{\varphi, \psi} \mapsto \set{\sigma\wedge\varphi,\sigma\wedge\psi}\\
-    \sigma \to \set{\varphi, \psi} \mapsto \set{\sigma\to\varphi,\sigma\to\psi}\\
+    %\sem{$\cdot$}(\rho') \mapsto \set{\varphi,\psi}\\
+    %\sem{$\cdot$}(\sigma_1,\dots,\sigma_j,\rho )
+    \varphi
+    %\sigma \wedge \set{\varphi, \psi} \mapsto \set{\sigma\wedge\varphi,\sigma\wedge\psi}\\
+    %\sigma \to \set{\varphi, \psi} \mapsto \set{\sigma\to\varphi,\sigma\to\psi}\\
   \end{eqnarray*}

### 2008-08-30 src/advertentCurt.pl
disjunction and possible worlds now working

@@ -131,29 +131,64 @@ curtUpdate(Input,Moves,run) :-
 	, maplist(disjunk,Rs,Nested)
 	, flatten(Nested,Readings)
 	, updateHistory(Input)
-	,
-	% TODO: Write Answer Code
-	interpret(Model,Readings,Moves)
+	, interpretReadings(Readings,Model)
 	.
 
 curtUpdate(_,[noparse],run).
 
 domainSize(15).
 
-% fails if Old is not a curt world or empty
-interpret(Old,New,World) :-
-	getKnowledge(Old,New,BK,Others,Reading)
-	, check(and(BK,New),'consistency',BBModel)
-	, check(and(BK,not(New)),'informativity',_)
-	, BBModel = model(D,F)
-	, World = world(D,[curt|Others],Reading,F)
+interpretReadings(Readings,Model) :-
+	model(Old)
+	, interpretReadings(Old,Readings,Model)
+	, assert(curt:model(Model))
 	.
 
-getKnowledge(world([curt|Others],Old,_Functions),New,and(BackgroundKnowledge,Old),Others,and(Old,New)) :- 
-	backgroundKnowledge(and(Old,New),BackgroundKnowledge).
+interpretReadings([],Readings,Ws) :-
+	maplist(curt:interpret([]),Readings,Ws)
+	.
 
-getKnowledge([],New,BackgroundKnowledge,[],New) :- 
+
+interpretReadings([(_Index,World)|Worlds],Readings,NewWorlds) :-
+	maplist(curt:interpret(World),Readings,W1)
+	,
+	(
+		\+ Worlds = []
+		, interpretReadings(Worlds,Readings,Ws)
+		, append(W1,Ws,NewWorlds)
+	;
+		Worlds = []
+		, NewWorlds = W1
+	)
+	.
+
+% Interpret one old reading (may be the empty list) wrt to one new reading
+% and gives back an index/world pair
+interpret(Old,New,World) :-
+	getKnowledge(Old,New,BK,Reading)
+	,
+	(
+		check(and(BK,New),'consistency',BBModel), !
+		,
+		(
+			check(and(BK,not(New)),'informativity',_), !
+			, BBModel = model(D,F)
+			, World = (_,world(D,F,Reading))
+		;
+			format('~nFound uninformative reading. Dropping world.',[])
+			, World = []
+		)
+	;
+		format('~nFound inconsistency. Dropping world.',[])
[…]

### 2008-08-30 thesis/paper.tex
wrote intro

@@
 \section{Introduction}
+
+As access to vast amounts of information becomes more and more ubiquitous, and
+the information accessible to humans grows more diverse every day, users
+struggle to find what they need among the vast amounts of data. Current
+search technologies aim to provide easy access to a constantly changing pool of
+knowledge, yet they are hardly beyond the stage of looking for exact text
+matches. Machine processing still fails to \emph{understand} the information it
+has to deal with and sees data as nothing more than a set of strings. But often
+a user's demands do not ask for a piece of text, but for an abstract concept, or
+a relation. Knowing that natural language is most likely infinite, such
+relations are virtually endless in number, and  one can only
+conclude that trying to enumerate all human knowledge in gargantuan databases is
+bound to be a never-ending task.
 

### 2008-08-30 thesis/paper.tex
Put two new papers in discussion, correct mistakes

--- 2008-08-29 thesis/paper.tex
@@
 
+As access to vast amounts of information becomes more and more ubiquitous, and
+the information accessible to humans grows more diverse every day, users
+struggle to find what they need among the vast amounts of data available to
+them. Current search technologies aim to provide easy access to a constantly
+changing pool of knowledge, yet they are hardly beyond the stage of looking for
+exact text matches. Machine processing still fails to \emph{understand} the
+information it has to deal with and sees data as nothing more than a set of
+strings. But often a user's demands do not ask for a piece of text, but for an
+abstract concept, or a relation. Knowing that natural language is most likely
+infinite, such relations are virtually endless in number, and  one can only
+conclude that trying to enumerate all relations between the bits of  human
+knowledge in gargantuan databases is bound to be a never-ending task.
+
 Growing efforts have thus been made to provide a technology that is (going to
@@
 ultimate goal may be to enable a man-made system to \emph{understand} what
-humans communicate.
+humans communicate. Then, these representations of knowledge could be queried, and
+infer from their knowledge, what an optimal answer to a given query might be --
+maybe even inventing an answer to a question never posed before along the way.
 
@@
 
-\subsection{Comparison of the Curt System}\label{sec:comparison}
+\subsection{Inferencing, Question Answering, and Dialogue Systems Today}\label{sec:comparison}
 
@@
 
+\cite{tenCate} present a tableaux method for resolving questions and also
+propose an algorithm in \prol{} as an implementation. Their approach is a more
+theoretical one, and they do not provide any particular implementation of their
+work. Although their methods seems promising, to say the least, it was chosen to
+avoid the implementation of a specialized inferencing system. \cite{bosgabdsil}
+mention the combinatorial explosion that lurks behind the treatment of questions
+and this is a considerable threat to any implementation of deep semantic
+\abbr{QA} systems. By using external inferencing tools, an implementation of
+the semantics of questions can focus on more essential issues than the
+resolution of formulae. % too offending. fixme
+
+Discourse Representation Theory is also used in \cite{bosgabdsil}, who present
+an approach to question answering quite similar to what is presented in this
+paper. In contrast to \cite{tenCate}, who use their own internal inferencing
+method to resolve questions, Bos and Gabdsil also use external inferencing very
+similar to the \curt{} system. Their results will be discussed later on, and
+explicitly compared to the present results.
+
+An attempt to bring inferencing to the world of \abbr{QA} systems can be found
+in \cite{webber}.
+
 Interestingly, earlier attempts at providing a software capable of answering
@@
 
-Today it seems that statistical question answering systems make inroads to the search market,
-and some argue that they are bound to eventually supplant current document/term-based
-search engines. Even though \abbr{QA} systems also operate on the domain of
-answering questions, their approach to resolving them differs in most aspects
-from that of dialogue systems. The latter have a much higher obligation to
-correctness, since their results are typically not subject to peer
-review and serve an entirely different purpose.
+Today it is clear that statistical question answering systems make inroads to
+the search market, and not many doubt that they are bound to eventually replace
+current document/term-based search engines. Even though \abbr{QA} systems also
+operate on the domain of answering questions, their approach to resolving them
+differs in most aspects from that of dialogue systems. The latter have a much
+higher obligation to correctness, since their results are typically not subject
+to peer review and serve an entirely different purpose.
 

### 2008-08-30 src/englishGrammar.pl
Grammar should be perfect now. YEAH

@@ -107,23 +107,23 @@ q([sem:Sem])-->
  */
 
 %todo: really coord:no for s? Bill knows that either Jane loves Mary or …
-tbar([sem:TBar])
--->	whemb([sem:WHemb])
+tbar([sem:TBar]) -->
+	whemb([sem:WHemb])
 	, s([coord:no,sem:S])
 	, { combine(tbar:TBar,[whemb:WHemb,s:S]) }
-.
+	.
 
-% for empty complementizers
-tbar([sem:TBar])
--->	s([coord:no,sem:S])
+% for empty complementizers 'Mia knows Vincent snorts'
+tbar([sem:TBar]) -->
+	s([coord:no,sem:S])
 	, { combine(tbar:TBar,[s:S]) }
-.
+	.
 
 % for 'mia knows who likes vincent'
-tbar([sem:TBar])
--->	q([sem:Q])
+tbar([sem:TBar]) -->
+	q([sem:Q])
 	, { combine(tbar:TBar,[q:Q]) }
-.
+	.
 
 /*========================================================================
    Noun Phrases
@@ -213,11 +213,37 @@ nmod([sem:Sem])-->
    Verb Phrases
 ========================================================================*/
 
-vp([coord:no,inf:Inf,num:Num,gap:[],sem:VP])
--->	ivtbar([inf:Inf,num:Num,sem:IVTBAR]) % verb
+vp([coord:no,inf:Inf,num:Num,gap:[],sem:VP]) -->
+	ivtbar([inf:Inf,num:Num,sem:IVTBAR,type:question]) % verb
 	, tbar([sem:TBAR]) % embedded sentence
-	, {combine(vp:VP,[ivtbar:IVTBAR,tbar:TBAR])}
-.
+	,
+	{
+		TBAR = [app(lam(X,que(_,_,X)),X)]
+		, combine(vp:VP,[ivtbar:IVTBAR,tbar:TBAR])
+	}
+	.
+
+vp([coord:no,inf:Inf,num:Num,gap:[],sem:VP]) -->
+	ivtbar([inf:Inf,num:Num,sem:IVTBAR,type:propos]) % verb
+	, tbar([sem:TBAR]) % embedded sentence
+	,
+	{
+		\+ TBAR = [app(lam(X,que(_,_,X)),X)]
[…]

### 2008-08-30 src/englishLexicon.pl

@@ -25,17 +25,21 @@
  * wh-embedding verbs
  * cf karttunen 77 for the complete classification according to 'type'
  */
-lexEntry(ivtbar,[symbol:knowi,syntax:[know],inf:inf,num:sg,type:assert]).
-lexEntry(ivtbar,[symbol:knowi,syntax:[knows],inf:fin,num:sg,type:assert]).
-lexEntry(ivtbar,[symbol:knowi,syntax:[know],inf:fin,num:pl,type:assert]).
+lexEntry(ivtbar,[symbol:knowledge,syntax:[know],inf:inf,num:sg,type:factive]).
+lexEntry(ivtbar,[symbol:knowledge,syntax:[knows],inf:fin,num:sg,type:factive]).
+lexEntry(ivtbar,[symbol:knowledge,syntax:[know],inf:fin,num:pl,type:factive]).
 
-lexEntry(ivtbar,[symbol:believei,syntax:[believe],inf:inf,num:sg,type:stipul]).
-lexEntry(ivtbar,[symbol:believei,syntax:[believes],inf:fin,num:sg,type:stipul]).
-lexEntry(ivtbar,[symbol:believei,syntax:[believe],inf:fin,num:pl,type:stipul]).
+lexEntry(ivtbar,[symbol:report,syntax:[report],inf:inf,num:sg,type:resolutive]).
+lexEntry(ivtbar,[symbol:report,syntax:[reports],inf:fin,num:sg,type:resolutive]).
+lexEntry(ivtbar,[symbol:report,syntax:[report],inf:fin,num:pl,type:resolutive]).
 
-lexEntry(ivtbar,[symbol:thinki,syntax:[think],inf:inf,num:sg,type:stipul]).
-lexEntry(ivtbar,[symbol:thinki,syntax:[thinks],inf:fin,num:sg,type:stipul]).
-lexEntry(ivtbar,[symbol:thinki,syntax:[think],inf:fin,num:pl,type:stipul]).
+lexEntry(ivtbar,[symbol:believe,syntax:[believe],inf:inf,num:sg,type:propos]).
+lexEntry(ivtbar,[symbol:believe,syntax:[believes],inf:fin,num:sg,type:propos]).
+lexEntry(ivtbar,[symbol:believe,syntax:[believe],inf:fin,num:pl,type:propos]).
+
+lexEntry(ivtbar,[symbol:wonder,syntax:[wonder],inf:inf,num:sg,type:question]).
[…]

### 2008-08-30 src/semLexStorage.pl

@@ -67,14 +67,13 @@ semLex(ivtbar,M)
 .
 
 % a simple unit rule for assertive embedders
-semLex(whemb,M)
-:-	M = [sem:[lam(S,S),type:assert]]
+semLex(whemb,M) :-
+	M = [sem:[lam(S,S)],type:propos]
 .
 
-% a more complex rule for yes/no embedders
-semLex(whemb,M)
-:-	M = [sem:[lam(S,or(S,not(S))),type:altern]]
-.
+semLex(whemb,M) :-
+	M = [sem:[lam(Q, que([],[],Q))],type:question]
+	.
 
 semLex(qnp,M):-
    M = [type:wh,

### 2008-08-30 src/semRulesKeller.pl

@@ -55,16 +55,16 @@ combine(q:Q,[sinv:Q]):-
    Q=que(_,_,_).
 
 
-combine(tbar:[app(WH,SRed)|S1],[whemb:[WH],s:[S|S1]])
-:-	betaConvert(S,SRed)
+combine(tbar:[app(WH,SRed)],[whemb:[WH],s:S]) :-
+	betaConvert(S,SRed)
 .
 
-combine(tbar:[SRed],[s:S])
-:-	betaConvert(S,SRed)
+combine(tbar:[SRed],[s:S]) :-
+	betaConvert(S,SRed)
 .
 
-combine(tbar:[QRed],[q:Q])
-:-	betaConvert(Q,QRed)
+combine(tbar:[QRed],[q:Q]) :-
+	betaConvert(Q,QRed)
 .
 
 combine(np:A,[pn:A]).

### 2008-08-30 thesis/paper.tex
Cleaned up sec 4, added grammar stuff.

@@
 \usepackage{xltxtra,xunicode}
+
+\usepackage{listings}
 
@@
 \begin{document}
+\lstset{language=Prolog,numbers=left,numberstyle=\tiny,numbersep=5pt,stepnumber=2,basicstyle=\ttfamily,identifierstyle=\textbf,keywordstyle=\underbar}
 \renewcommand{\thepage}{\roman{page}}
@@
 
-\subsection{Embedded Questions}
+\subsection{Embedded Questions}\label{sec:embed}
 
@@
 
-\begin{definition}
+\begin{definition}\label{def:embed2}
   Resolutive and factive predicates:
@@
 
+We have seen that some embedding verbs may not be interpreted without
+implementing a higher order formalism that is able to model the notion of
+\emph{possible} and \emph{accessible} worlds. While foobar %fixme! Nelken &
+% Francez 2001
+argue for an extensional treatment of questions, they already % n & f 2006
+retract their claims and admit that a ``minimal account of intensionality'' has
+to be assumed. While they propose a two-world system later adopted in
+\cite{g:is}, this is not going to suffice if we want to model each individual's
+knowledge in order to be able to discriminate people's knowledge and beliefs.
+%fixme bla bla
+
+The underlying theory for 
 TODO Discussion of Lewis' paper.
@@
 
+\prol{} programs can be self-modifying, or dynamic. In this way the program is
+able to change its own state and thus store information.
+
+\subsubsection{Disjunction and Ambiguity}
+Consider the following sentences:
+
+\ex. \label{ex:disjunction} \a. Mia or Jody dances
+\b. Either Butch snores or Vincent is using the chainsaw
+
+\pn{Helpful Curt}, with its unpartitioned logical space and its lack of
+the awareness of more than one (possible) world, would generate only one model
+to represent the facts of \ref{ex:disjunction}. It is therefore only able to
+generate \emph{one} possible answer to both questions, effectively accounting only
+for one possible constituent of the disjunction, prefixing it with a `maybe' to
+indicate its uncertainty.
+
+Another example where \curt{} has to sacrifice exhaustiveness over its one-world
+representation of the world is given in \ref{ex:ambig}
+
+\ex.\label{ex:ambig} Every boxer likes a woman.
+
+This will generally generate two readings, but only one model -- with only one
+world. And further discourse will have to choose between one of the readings and
+stick with the minimal model constructed by the model builder. This is not
+really satisfactory and extending the dimensions of knowledge within \curt{} seems
+necessary in order to treat such examples.
+
+The previous work on \acurt{} enables it to generate a \term{choice point} -- and
+moreover, not pursue only one path of it, but both at the same time, by
+generating more than one world in its model.
+Unfortunately, this will build up very large models over time and computations
+with it might prove to be very complex and resource demanding. Adding such a
+functionality to the system, however, is not a hard task and hopefully the
+technique employed here can be further refined. %todo: maybe add a footnote
+%about future prospects or so
+
 We first derive know(john,all(X,imp(woman(X),love(X,john)))) from the syntax of
@@
 
-\subsubsection{Advertent Curt}
+\subsection{Advertent Curt}
 
@@
 
-\subsubsection{The Syntax of Questions}
+\subsection{The Syntax of Embedded Questions}
 
@@
 
-However, the syntax of embedded interrogatives is a more delicate matter.
-
-\subsection{Modality and Possible Worlds}
-\label{sec:indices}
-
-We have seen that some embedding verbs may not be interpreted without
-implementing a higher order formalism that is able to model the notion of
-\emph{possible} and \emph{accessible} worlds. While foobar %fixme! Nelken &
-% Francez 2001
-argue for an extensional treatment of questions, they already % n & f 2006
-retract their claims and admit that a ``minimal account of intensionality'' has
-to be assumed. While they propose a two-world system later adopted in
-\cite{g:is}, this is not going to suffice if we want to model each individual's
-knowledge in order to be able to discriminate people's knowledge and beliefs.
-%fixme bla bla
-
-The underlying theory for
-
-\subsubsection{Representing Intensionality in \prol{}}
-
-\prol{} programs can be self-modifying, or dynamic. In this way the program is
-able to change its own state and thus store information.
-
-\subsubsection{Disjunction and Ambiguity}
-Consider the following sentences:
-
-\ex. \label{ex:disjunction} \a. Mia or Jody dances
-\b. Either Butch snores or Vincent is using the chainsaw
-
-\pn{Helpful Curt}, with its unpartitioned logical space and its lack of
-the awareness of more than one (possible) world, would generate only one model
-to represent the facts of \ref{ex:disjunction}. It is therefore only able to
-generate \emph{one} possible answer to both questions, effectively accounting only
-for one possible constituent of the disjunction, prefixing it with a `maybe' to
-indicate its uncertainty.
-
-Another example where \curt{} has to sacrifice exhaustiveness over its one-world
-representation of the world is given in \ref{ex:ambig}
-
-\ex.\label{ex:ambig} Every boxer likes a woman.
-
-This will generally generate two readings, but only one model -- with only one
-world. And further discourse will have to choose between one of the readings and
-stick with the minimal model constructed by the model builder. This is not
-really satisfactory and extending the dimensions of knowledge within \curt{} seems
-necessary in order to treat such examples.
-
-The previous work on \acurt{} enables it to generate a \term{choice point} -- and
-moreover, not pursue only one path of it, but both at the same time, by
-generating more than one world in its model.
-Unfortunately, this will build up very large models over time and computations
-with it might prove to be very complex and resource demanding. Adding such a
-functionality to the system, however, is not a hard task and hopefully the
-technique employed here can be further refined. %todo: maybe add a footnote
-%about future prospects or so
-
-\subsubsection{Defining the Grammar Rules}
+However, the syntax of embedded interrogatives is a more delicate matter. Since,
+as was discussed in section \ref{sec:embed}, not every embedding verb can embed
+every type of sentence (recall definition \ref{def:embed1} and definition
+\ref{def:embed2}), a type system has to be established in order to account for
+the different behaviour of the verbs, and said type system has to be integrated
+into the process of syntactic and semantic parsing of the system.
 
@@
 presented in \cite{keller:storage}. \cite{blackburnbos:cl1} call this version
-\pn{Keller Storage}. That name will be used from here % FIXME sucks. Please change
-to refer to this particular implementation as presented in
-\cite{blackburnbos:cl1}.
+\pn{Keller Storage}. The remainder of this section will follow the nomenclature
+of \cite{blackburnbos:cl1}.
 
@@
 %Bos' paper
-}. However none of the cases discussed so far would require the enhanced
+}. 
+However none of the cases discussed so far would require the enhanced
 expressive power of \pn{Hole Semantics} it was decided to not change the \curt{}
@@
 Enhancing the grammatic rules of the \curt{} system to cope with
-proposition-embedding verbs is made trivial by its elaborate grammatical back
-end. \ref{ex:kel1} is an example of the logic behind the grammar that is put to
-use.
-
-\footnotetext{Note that \ref{ex:kel1} is shown using the same syntactic type system used in
-\cite{gs:sawhq}, which extends the system given in \cite{ptq} by $\bar{t}$ to
-denote embedded \wh-constituents.}
+proposition-embedding verbs is made easy by its elaborate grammatical back
+end. In particular, four files are enhanced with rules to enable interpretation
+of embedded questions: \code{englishGrammar.pl}, which holds generic \abbr{DCG}
+rules for syntactic parsing, \code{englishLexicon.pl}, a dictionary that uses
+feature structures to keep track of agreement, and two \pn{Keller Storage}
+specific files, \code{semLexStorage.pl} and \code{semRulesKeller.pl}.
+
+The grammatic rules are extended by the introduction of three new categories,
+\code{whemb}, \code{tbar} and \code{ivtbar}.
+
+The \code{whemb} and \code{ivtbar} categories have a \code{type} parameter, that dictates
+their behaviour, although their respective type systems never cooperate with
+each other. The type of the \code{whemb} category can hold the values
+\code{propos}, for a proposition embedding complementizer, like \example{that},
+and \code{question}, for a complementizer that turns a proposition into a yes/no
+interrogative, like \example{if} and \example{whether} (listing \ref{whemb}).
+
+\begin{lstlisting}[label=whemb,caption={Syntactic and Lexical Rules for Complementizers},float]
+whemb([sem:Sem]) -->
+    {lexEntry(whemb,[syntax:Word,type:T])}
+    , Word
+    , {semLex(whemb,[sem:Sem,type:T])}
+    .
+lexEntry(whemb,[syntax:[that],type:propos]).
+lexEntry(whemb,[syntax:[whether],type:question]).
+\end{lstlisting}
+
+\begin{lstlisting}[label=whembsemLex,caption={Semantic Rules for Complementizers},float]
+semLex(whemb,M) :-
+    M = [sem:[lam(S,S)],type:propos]
+    .
+semLex(whemb,M) :-
+    M = [sem:[lam(Q, que([],[],Q))],type:question]
+    .
+\end{lstlisting}
+
+The types now determine whether to produce an embedded yes/no
+interrogative or just pass through as a simple proposition (in which case a
+simple $\lambda S. S$ serves as an identity function), and the
+\code{semLex/2} predicate takes care of this (listing \ref{whembsemLex}). In the
+latter case, a new sort of question is introduced, namely one that does not
+quantify over a free variable. It has the form \lstinline!que([],[],Q)! and is
+representative of $?.\varphi$.
+
+The \code{tbar} rule has three different right hand sides. 
+In one of them, it first parses a \code{whemb} and combines it with a sentential
+category \code{s} (listing \ref{tbar}, line 1). The second case omits the
+complementizer and just passes the sentence as a normal proposition (line 7).
+Next, the final rule takes care of embedding questions (line 12).
+
+In every case, the sentence embedded is first fully reduced by the predicates
+shown in listing \ref{tbarCombine}. This has to be done in order to avoid
+unsaturated formulae in the end result. Thus the $\bar{t}$ category can   produce
+one of the following: $\varphi$, $?.\varphi$ or $?x.\varphi$.
+
+\begin{lstlisting}[label=tbar,caption={Semantic Rules for $\bar{t}$},float]
+tbar([sem:TBar]) -->
+    whemb([sem:WHemb])
+    , s([coord:no,sem:S])
+    , { combine(tbar:TBar,[whemb:WHemb,s:S]) } .
+
+tbar([sem:TBar]) -->
+    s([coord:no,sem:S])
+    , { combine(tbar:TBar,[s:S]) } .
+
+tbar([sem:TBar]) -->
+    q([sem:Q])
+    , { combine(tbar:TBar,[s:Q]) } .
+\end{lstlisting}
+
+\begin{lstlisting}[label=tbarCombine,caption={$\bar{t}$ is responsible for beta-converting the embedded sentence.},float]
+combine(tbar:[app(WH,SRed)],[whemb:[WH],s:S]) :-
+    betaConvert(S,SRed).
+
+combine(tbar:[SRed],[s:S]) :-
+    betaConvert(S,SRed).
+\end{lstlisting}
+
+As a final step, \code{ivtbar} (or $IV/\bar{t})$ in a formal type system), takes
+the result of \code{tbar} and embeds it under a suitable verb, taking the verb
+type (one of \code{factive}, \code{resolutive}, \code{question} and
+\code{propos}) and the nature of the embedded sentence into account. Listing
+\ref{ivtbar} shows only one case, for \code{question} embedding verbs, and the
+(semantic) lexical rules have been omitted. See the
+appendix %todo: section number
+for the full definition.
+
+\begin{lstlisting}[label=ivtbar,caption={A question embedding VP requires its antecedent to be a question.},float]
+vp([coord:no,inf:Inf,num:Num,gap:[],sem:VP]) -->
+    ivtbar([inf:Inf,num:Num,sem:IVTBAR,type:question])
+   , tbar([sem:TBAR])
+   , { TBAR = [app(lam(X,que(_,_,X)),X)]
+       , combine(vp:VP,[ivtbar:IVTBAR,tbar:TBAR])
+     }.
+\end{lstlisting}
+
+\subsection{Modality and Possible Worlds}
+\label{sec:indices}
+
+\subsubsection{Representing Intensionality in \prol{}}
 

### 2008-08-30 src/englishGrammar.pl
fixed bug in grammar
we now reduce the term fully also for alternative questions

@@ -218,7 +218,7 @@ vp([coord:no,inf:Inf,num:Num,gap:[],sem:VP]) -->
 	, tbar([sem:TBAR]) % embedded sentence
 	,
 	{
-		TBAR = [app(lam(X,que(_,_,X)),X)]
+		TBAR = [que(_,_,_)]
 		, combine(vp:VP,[ivtbar:IVTBAR,tbar:TBAR])
 	}
 	.
@@ -228,7 +228,7 @@ vp([coord:no,inf:Inf,num:Num,gap:[],sem:VP]) -->
 	, tbar([sem:TBAR]) % embedded sentence
 	,
 	{
-		\+ TBAR = [app(lam(X,que(_,_,X)),X)]
+		\+ TBAR = [que(_,_,_)]
 		, combine(vp:VP,[ivtbar:IVTBAR,tbar:TBAR])
 	}
 	.

### 2008-08-30 src/semRulesKeller.pl

@@ -54,18 +54,17 @@ combine(q:Q,[whnp:[A|S1],vp:[B|S2]]):-
 combine(q:Q,[sinv:Q]):-
    Q=que(_,_,_).
 
-
-combine(tbar:[app(WH,SRed)],[whemb:[WH],s:S]) :-
-	betaConvert(S,SRed)
-.
+combine(tbar:[SRed],[whemb:[WH],s:S]) :-
+	betaConvert(app(WH,S),SRed)
+	.
 
 combine(tbar:[SRed],[s:S]) :-
 	betaConvert(S,SRed)
-.
+	.
 
 combine(tbar:[QRed],[q:Q]) :-
 	betaConvert(Q,QRed)
-.
+	.
 
 combine(np:A,[pn:A]).
 combine(np:A,[qnp:A]).

### 2008-08-30 src/advertentCurt.pl
Temporary commit. include's still not working.

@@ -46,6 +46,8 @@ history([]).
 readings([]).
 models([]).
 
+% domain size for model builders
+domainSize(15).
 
 /*========================================================================
    Start Curt
@@ -75,8 +77,8 @@ curtTalk(run):-
 curtUpdate([],[clarify],run):- !.
 
 curtUpdate([bye],[bye],quit):- !,
-   updateReadings([]),
-   updateModels([]),
+   %updateReadings([]),
+   %updateModels([]),
    clearHistory.
 
 curtUpdate([new],[],run):- !,
@@ -132,25 +134,30 @@ curtUpdate(Input,Moves,run) :-
 	, flatten(Nested,Readings)
 	, updateHistory(Input)
 	, interpretReadings(Readings,Model)
+	,
+	(
+		\+ Model = []
+		, updateModels(Model), !
+		, Moves = [accept]
+	;
+		Moves = [reject]
+	)
 	.
 
 curtUpdate(_,[noparse],run).
 
-domainSize(15).
-
 interpretReadings(Readings,Model) :-
-	model(Old)
+	models(Old)
 	, interpretReadings(Old,Readings,Model)
-	, assert(curt:model(Model))
 	.
 
-interpretReadings([],Readings,Ws) :-
-	maplist(curt:interpret([]),Readings,Ws)
+interpretReadings([],Readings,Model) :-
+	include(curt:interpret([]),Readings,Model)
 	.
 
 
 interpretReadings([(_Index,World)|Worlds],Readings,NewWorlds) :-
-	maplist(curt:interpret(World),Readings,W1)
+	include(curt:interpret(World),Readings,W1)
 	,
 	(
 		\+ Worlds = []
@@ -175,12 +182,12 @@ interpret(Old,New,World) :-
 			, BBModel = model(D,F)
 			, World = (_,world(D,F,Reading))
 		;
-			format('~nFound uninformative reading. Dropping world.',[])
-			, World = []
+			format('~nFound uninformative reading. Dropping reading.',[])
+			, fail
 		)
 	;
 		format('~nFound inconsistency. Dropping world.',[])
-		, World = []
+		, fail
 	)
 	.
 

### 2008-08-30 src/curtPredicates.pl

@@ -117,8 +117,9 @@ clearHistory:-
 ========================================================================*/
 
 updateReadings(R):-
-   retract(curt:readings(_)),
+   retractall(curt:readings(_)),
    assert(curt:readings(R)).
+
 
 
 /*========================================================================
@@ -126,7 +127,7 @@ updateReadings(R):-
 ========================================================================*/
 
 updateModels(R):-
-   retract(curt:models(_)),
+   retractall(curt:models(_)),
    assert(curt:models(R)).
 
 

### 2008-08-30 src/advertentCurt.pl
fixed bugs. Now everything should be fine.

@@ -146,18 +146,24 @@ curtUpdate(Input,Moves,run) :-
 
 curtUpdate(_,[noparse],run).
 
+noEmpties([],[]).
+noEmpties([[]|Xs],L) :- !, noEmpties(Xs,L).
+noEmpties([X|Xs],[X|L]) :- noEmpties(Xs,L).
+
 interpretReadings(Readings,Model) :-
 	models(Old)
-	, interpretReadings(Old,Readings,Model)
+	, interpretReadings(Old,Readings,M)
+	, noEmpties(M,Model)
 	.
 
 interpretReadings([],Readings,Model) :-
-	include(curt:interpret([]),Readings,Model)
+	maplist(curt:interpret([]),Readings,M)
+	, noEmpties(M,Model)
 	.
 
 
 interpretReadings([(_Index,World)|Worlds],Readings,NewWorlds) :-
-	include(curt:interpret(World),Readings,W1)
+	maplist(curt:interpret(World),Readings,W1)
 	,
 	(
 		\+ Worlds = []
@@ -183,11 +189,11 @@ interpret(Old,New,World) :-
 			, World = (_,world(D,F,Reading))
 		;
 			format('~nFound uninformative reading. Dropping reading.',[])
-			, fail
+			, World = (_,Old)
 		)
 	;
 		format('~nFound inconsistency. Dropping world.',[])
-		, fail
+		, World = []
 	)
 	.
 

### 2008-08-30 src/functionalPredicates.pl

@@ -6,7 +6,7 @@
 %
 % Aleksandar Dimitrov
 
-module(functionalPredicates,
+:- module(functionalPredicates,
 	[filter/3
 	, converse/4
 	, compose/4

### 2008-08-31 src/advertentCurt.pl
Curt is advertent. I can't believe it.

@@ -20,7 +20,7 @@
 
 :- use_module(kellerStorage,[kellerStorage/2]).
 
-:- use_module(intensionalModels,[satisfy/4]).
+:- use_module(modelChecker2,[satisfy/4]).
 
 :- use_module(backgroundKnowledge,[backgroundKnowledge/2]).
 
@@ -40,11 +40,13 @@
    Dynamic Predicates
 ========================================================================*/
 
-:- dynamic history/1, readings/1, models/1.
+:- dynamic history/1, readings/1, models/1, epistemic/1, doxastic/1.
 
 history([]).
 readings([]).
 models([]).
+epistemic([]).
+doxastic([]).
 
 % domain size for model builders
 domainSize(15).
@@ -178,22 +180,27 @@ interpretReadings([(_Index,World)|Worlds],Readings,NewWorlds) :-
 % Interpret one old reading (may be the empty list) wrt to one new reading
 % and gives back an index/world pair
 interpret(Old,New,World) :-
-	getKnowledge(Old,New,BK,Reading)
-	,
 	(
-		check(and(BK,New),'consistency',BBModel), !
+		beAdvertent(Old,New), !
+		, World = (_,Old)
+	;
+		getKnowledge(Old,New,BK,Reading)
 		,
 		(
-			check(and(BK,not(New)),'informativity',_), !
-			, BBModel = model(D,F)
-			, World = (_,world(D,F,Reading))
+			check(and(BK,New),'consistency',BBModel), !
+			,
+			(
+				check(and(BK,not(New)),'informativity',_), !
+				, BBModel = model(D,F)
+				, World = (_,world(D,F,Reading))
+			;
+				format('~nFound uninformative reading. Dropping reading.',[])
+				, World = (_,Old)
+			)
 		;
-			format('~nFound uninformative reading. Dropping reading.',[])
-			, World = (_,Old)
+			format('~nFound inconsistency. Dropping world.',[])
+			, World = []
 		)
-	;
-		format('~nFound inconsistency. Dropping world.',[])
-		, World = []
 	)
 	.
 
@@ -203,11 +210,55 @@ getKnowledge([],New,BackgroundKnowledge,New) :-
 getKnowledge(world(_D,_F,Old),New,and(BackgroundKnowledge,Old),and(Old,New)) :- 
 	backgroundKnowledge(and(Old,New),BackgroundKnowledge).
 
[…]

### 2008-08-31 src/advertentCurt.pl
fixed model checker polarity

@@ -102,7 +102,7 @@ curtUpdate([select,X],[],run):-
    readings(R1),
    selectReadings(X,R1,R2), !,
    updateReadings(R2),
-   models(M1),
+   models(M1), !,
    selectReadings(X,M1,M2),
    updateModels(M2).
 
@@ -153,7 +153,7 @@ noEmpties([[]|Xs],L) :- !, noEmpties(Xs,L).
 noEmpties([X|Xs],[X|L]) :- noEmpties(Xs,L).
 
 interpretReadings(Readings,Model) :-
-	models(Old)
+	models(Old) , !
 	, interpretReadings(Old,Readings,M)
 	, noEmpties(M,Model)
 	.
@@ -212,17 +212,17 @@ getKnowledge(world(_D,_F,Old),New,and(BackgroundKnowledge,Old),and(Old,New)) :-
 
 beAdvertent(world(_D,_F,Background),knowledge(X,P)) :-
 	(
-		P = que([],[],Q)
+		P = que(_,[],Q)
 		, !
 		, backgroundKnowledge(and(Q,Background),BK2)
 		,
 		(
-			check(and(and(Background,BK2),not(Q)),'yes/ interrogative: informativity',_)
+			\+ check(and(and(Background,BK2),not(Q)),'yes/ interrogative: informativity',_)
 			, !
 			, backgroundKnowledge(Q,BK)
 			, check(and(Q,BK),'preparing world: consistency', model(D,F))
 		;
-			check(and(and(Background,BK2),Q),'/no interrogative: informativity',_)
+			\+ check(and(and(Background,BK2),Q),'/no interrogative: informativity',_)
 			, backgroundKnowledge(Q,BK)
 			, check(and(not(Q),BK),'preparing world: consistency', model(D,F))
 		)
@@ -234,7 +234,7 @@ beAdvertent(world(_D,_F,Background),knowledge(X,P)) :-
 		, fail
 	;
 		backgroundKnowledge(and(P,Background),BK2)
-		, check(and(and(Background,BK2),not(P)),'embedded proposition: informativity',_)
+		, \+ check(and(and(Background,BK2),not(P)),'embedded proposition: informativity',_)
 		, backgroundKnowledge(P,BK)
 		, check(and(P,BK),'preparing world: consistency', model(D,F))
 		, World = (X,world(D,F,Q))

### 2008-08-31 src/advertentCurt.pl
worked around a disjunk bug

@@ -212,7 +212,7 @@ getKnowledge(world(_D,_F,Old),New,and(BackgroundKnowledge,Old),and(Old,New)) :-
 
 beAdvertent(world(_D,_F,Background),knowledge(X,P)) :-
 	(
-		P = que(_,[],Q)
+		P = que(_,alt,Q)
 		, !
 		, backgroundKnowledge(and(Q,Background),BK2)
 		,

### 2008-08-31 src/semLexStorage.pl

@@ -72,7 +72,7 @@ semLex(whemb,M) :-
 .
 
 semLex(whemb,M) :-
-	M = [sem:[lam(Q, que([],[],Q))],type:question]
+	M = [sem:[lam(Q, que([],alt,Q))],type:question]
 	.
 
 semLex(qnp,M):-

### 2008-08-31 src/advertentCurt.pl
Added epistemic history.

@@ -159,12 +159,12 @@ interpretReadings(Readings,Model) :-
 	.
 
 interpretReadings([],Readings,Model) :-
-	maplist(curt:interpret([]),Readings,M)
+	maplist(curt:interpret((_,[])),Readings,M)
 	, noEmpties(M,Model)
 	.
 
 
-interpretReadings([(_Index,World)|Worlds],Readings,NewWorlds) :-
+interpretReadings([World|Worlds],Readings,NewWorlds) :-
 	maplist(curt:interpret(World),Readings,W1)
 	,
 	(
@@ -179,10 +179,10 @@ interpretReadings([(_Index,World)|Worlds],Readings,NewWorlds) :-
 
 % Interpret one old reading (may be the empty list) wrt to one new reading
 % and gives back an index/world pair
-interpret(Old,New,World) :-
+interpret((Index,Old),New,World) :-
 	(
-		beAdvertent(Old,New), !
-		, World = (_,Old)
+		beAdvertent((Index,Old),New), !
+		, World = (Index,Old)
 	;
 		getKnowledge(Old,New,BK,Reading)
 		,
@@ -192,10 +192,10 @@ interpret(Old,New,World) :-
 			(
 				check(and(BK,not(New)),'informativity',_), !
 				, BBModel = model(D,F)
-				, World = (_,world(D,F,Reading))
+				, World = (Index,world(D,F,Reading))
 			;
 				format('~nFound uninformative reading. Dropping reading.',[])
-				, World = (_,Old)
+				, World = (Index,Old)
 			)
 		;
 			format('~nFound inconsistency. Dropping world.',[])
@@ -210,34 +210,70 @@ getKnowledge([],New,BackgroundKnowledge,New) :-
 getKnowledge(world(_D,_F,Old),New,and(BackgroundKnowledge,Old),and(Old,New)) :- 
 	backgroundKnowledge(and(Old,New),BackgroundKnowledge).
 
-beAdvertent(world(_D,_F,Background),knowledge(X,P)) :-
+getEpistemicBG(X,I,world(D,F,R)) :-
+	epistemic(E), !
+	, delete(E,(X,I,world(D,F,R)),Rest)
+	, \+ Rest = E
+	, retract(epistemic(_)), !
+	, assert(epistemic(Rest))
+	.
+
+getDoxasticBG(X,I,world(D,F,R)) :-
+	doxastic(E), !
[…]

### 2008-08-31 src/advertentCurt.pl
some basic treatment of wh-predicates.

@@ -163,7 +163,6 @@ interpretReadings([],Readings,Model) :-
 	, noEmpties(M,Model)
 	.
 
-
 interpretReadings([World|Worlds],Readings,NewWorlds) :-
 	maplist(curt:interpret(World),Readings,W1)
 	,
@@ -226,7 +225,7 @@ getDoxasticBG(X,I,world(D,F,R)) :-
 	, assert(doxastic(Rest))
 	.
 
-beAdvertent((Index,world(_D,_F,Background)),knowledge(X,P)) :-
+beAdvertent((Index,world(D,F,Background)),knowledge(X,P)) :-
 	(
 		P = que(_,alt,Q)
 		, !
@@ -247,19 +246,23 @@ beAdvertent((Index,world(_D,_F,Background)),knowledge(X,P)) :-
 			\+ check(and(and(Background,BK2),not(BG)),'yes/ interrogative: informativity',_)
 			, !
 			, backgroundKnowledge(BG,BK)
-			, check(and(BG,BK),'preparing world: consistency', model(D,F))
-			, World = (X,Index,world(D,F,BG))
+			, check(and(BG,BK),'preparing world: consistency', model(D2,F2))
+			, World = (X,Index,world(D2,F2,BG))
 		;
 			\+ check(and(and(Background,BK2),BG),'/no interrogative: informativity',_)
 			, backgroundKnowledge(NBG,BK)
-			, check(and(NBG,BK),'preparing world: consistency', model(D,F))
-			, World = (X,Index,world(D,F,NBG))
+			, check(and(NBG,BK),'preparing world: consistency', model(D2,F2))
+			, World = (X,Index,world(D2,F2,NBG))
 		)
 		, addEpistemic(World)
 	;
-		P = que(_Y,_Domain,_Body)
+		P = que(Y,Domain,Body)
 		, !
-		, fail
+		, satisfy(some(Y,and(Domain,Body)),model(D,F),[],Result)
+		, \+ Result = undef
+		, !
+		, findall(A,satisfy(and(Domain,Body),model(D,F),[g(Y,A)],pos),Answers)
+		, realiseAnswer(Answers,que(Y,Domain,Body),model(D,F),String)
 	;
 		(
 			getEpistemicBG(X,I,world(_,_,EBG))
@@ -272,8 +275,8 @@ beAdvertent((Index,world(_D,_F,Background)),knowledge(X,P)) :-
 		backgroundKnowledge(and(Q,Background),BK2)
[…]

### 2008-08-31 src/advertentCurt.pl
OK, basic treatment of constituent interrogatives.
Now is not the time to continue with the implementation...

@@ -229,8 +229,7 @@ beAdvertent((Index,world(D,F,Background)),knowledge(X,P)) :-
 	(
 		P = que(_,alt,Q)
 		, !
-		,
-		(
+		, (
 			getEpistemicBG(X,I,world(_,_,EBG))
 			, !
 			, BG = and(Q,EBG)
@@ -238,11 +237,9 @@ beAdvertent((Index,world(D,F,Background)),knowledge(X,P)) :-
 		;
 			BG = Q
 			, NBG = not(Q)
-
 		)
 		, backgroundKnowledge(and(BG,Background),BK2)
-		,
-		(
+		, (
 			\+ check(and(and(Background,BK2),not(BG)),'yes/ interrogative: informativity',_)
 			, !
 			, backgroundKnowledge(BG,BK)
@@ -262,7 +259,27 @@ beAdvertent((Index,world(D,F,Background)),knowledge(X,P)) :-
 		, \+ Result = undef
 		, !
 		, findall(A,satisfy(and(Domain,Body),model(D,F),[g(Y,A)],pos),Answers)
-		, realiseAnswer(Answers,que(Y,Domain,Body),model(D,F),String)
+		, realiseAnswer(Answers,que(Y,Domain,Body),world(D,F,Background),String)
+		,
+		(
+			String = none, !
+			, beAdvertent((Index,world(D,F,Background)),knowledge(X,que([],alt,some(Y,Body))))
+		;
+			Y = String
+			, (
+				getEpistemicBG(X,I,world(_,_,EBG))
+				, !
+				, BG = and(Body,EBG)
+				, NBG = and(not(Body),EBG)
+			;
+				BG = Body
+				, NBG = not(Body)
+			) 
+			, backgroundKnowledge(BG,BK)
+			, check(and(BG,BK),'preparing world: consistency', model(D2,F2))
+			, World = (X, Index,world(D2,F2,BG))
+			, addEpistemic(World)
+		)
 	;
 		(
 			getEpistemicBG(X,I,world(_,_,EBG))
@@ -271,8 +288,7 @@ beAdvertent((Index,world(D,F,Background)),knowledge(X,P)) :-
 		;
 			Q = P
 		)
-		,
-		backgroundKnowledge(and(Q,Background),BK2)
+		, backgroundKnowledge(and(Q,Background),BK2)
 		, \+ check(and(and(Background,BK2),not(Q)),'embedded proposition: informativity',_)
[…]

### 2008-08-31 thesis/paper.tex
Final versions of sec 1 and 2

@@
 
-As access to vast amounts of information becomes more and more ubiquitous, and
+As access to vast amounts of information becomes increasingly ubiquitous, and
 the information accessible to humans grows more diverse every day, users
-struggle to find what they need among the vast amounts of data available to
-them. Current search technologies aim to provide easy access to a constantly
-changing pool of knowledge, yet they are hardly beyond the stage of looking for
-exact text matches. Machine processing still fails to \emph{understand} the
-information it has to deal with and sees data as nothing more than a set of
-strings. But often a user's demands do not ask for a piece of text, but for an
-abstract concept, or a relation. Knowing that natural language is most likely
-infinite, such relations are virtually endless in number, and  one can only
-conclude that trying to enumerate all relations between the bits of  human
-knowledge in gargantuan databases is bound to be a never-ending task.
+struggle to find what they need among the vast amounts of data available.
+Current search technologies aim to address this problem by providing quick and
+easy access to a constantly changing pool of knowledge, yet they are hardly
+beyond the stage of looking for exact text matches. Machine processing still
+fails to \emph{understand} the information it has to deal with and often sees data as
+nothing more than a set of strings. But what if a user's demands do not ask for a
+piece of text, but for an abstract concept, or a relation? Knowing that natural
+language is most likely infinite, such relations are virtually endless in
+number, and  one can only conclude that trying to enumerate all relations
+between the bits of  human knowledge in gargantuan databases is bound to be a
+never-ending task.
 
@@
 be) able to extract logical formulae from sentences of ordinary language and
-establish relations between objects of their logical representation. In addition,
-providing a means to compute with said representations will eventually enable
-the machine-processing of an abstract concept of meaning. In doing so, the
-ultimate goal may be to enable a man-made system to \emph{understand} what
-humans communicate. Then, these representations of knowledge could be queried, and
-infer from their knowledge, what an optimal answer to a given query might be --
-maybe even inventing an answer to a question never posed before along the way.
+establish relations between objects of their logical representation. In
+addition, providing a means to compute with said representations will eventually
+enable the machine-processing of an abstract concept of meaning. In doing so,
+the ultimate goal may be to enable a man-made system to \emph{understand} what
+humans communicate.  These systems could then be queried and, given a capability
+of performing logical inference on their domains of knowledge, they could
+compute what an optimal answer to any given query might be -- perhaps even
+by inventing an answer to a question never posed before.
 
@@
 
-The present work will aim to provide the \curt{} system of
-\cite{blackburnbos:cl1} with the mechanics of dealing with more than one
-partition of its logical domain, viz. the ability to handle more than one
-\term{possible} or \term{accessible world}. As a ground for the development of this
-work, the treatment of \term{embedded questions} was chosen as one that
-naturally demands a possibility to account for more than one division in a
-system's \emph{understanding} or representation of a universe of objects and
-relations.
+The present work aims to provide the \curt{} system of
+\cite{blackburnbos:cl1} with the mechanics of dealing with \term{embedded
+questions}, which enables it to reason not only about its own knowledge, but
+also about the knowledge of individuals in its target domain.
+For this, the system will have to enhanced with the capability of handling more
+than one partition of its logical domain, viz. the ability to handle several
+\term{possible} or \term{accessible worlds}.
 
 As the scope of a Bachelor's thesis is limited by its nature, not all of the
-theory and reasoning can be understood without background knowledge in some
-rather divergent domains. In particular, a basic familiarity with traditional
-\term{natural language semantics} and \term{extensional} as well as
+theory and reasoning can be explained and background knowledge in some rather
+divergent domains is assumed. In particular, a basic familiarity with
+traditional \term{natural language semantics} and \term{extensional} as well as
 \term{intensional logic} or \term{modal} frameworks will be assumed.
-Furthermore, a basic acquaintance with \term{eroteric logic} will be helpful,
+Furthermore, a basic acquaintance with theories of the logic of questions will be helpful,
 but not strictly necessary in understanding some of the reasoning in section
@@
 
-The first section will give a conceptual overview of the technologies to date
-and the particular software, \curt, that will serve as a basis for extension.
-The next section concerns itself with a more detailed analysis of the theoretic
-assumptions of this implementation and justifies some choices and restrictions
-that had to be made due to the limited scope of this analysis. While the
-third section will display and document the actual implementation on top of the
-\curt{} system, the last section will discuss the results of this work and also
-give a prospect on possible further work on the subject.
+Section \ref{sec:concepts}  gives a conceptual overview of the technologies to
+date and the particular software, \curt, that will serve as a basis for
+extension.  The next section concerns itself with a more detailed analysis of
+the theoretic assumptions of this implementation and justifies some choices and
+restrictions that had to be made due to the limited scope of this analysis.
+While the section \ref{sec:implementation} displays and documents the actual
+implementation on top of the \curt{} system, the last section discusses the
+results of the previous sections and also give a prospect on possible further work on the
+subject. Finally, the appendix holds the sources code and some example
+computations and models that show the basic functionality of \curt.
 
@@
   \item Code is written in a monospace font, resembling the look of a font class most
-  programmers are used to:\\
-  \code{lam(P, lam(Q, all(X, imp(app(P,X),app(Q,X)))))}.
+  programmers are used to:
+  \begin{lstlisting}
+    lam(P, lam(Q, all(X, imp(app(P,X),app(Q,X)))))
+  \end{lstlisting}
   \item Logical formulae are typeset in a different serif font from the rest of
@@
 
-\section{Concepts}
+\section{Concepts}\label{sec:concepts}
 
@@
 This section delineates the ground the software presented in this thesis is
-based on. \ref{sec:curt} introduces the base system of the implementation,
-\curt{},
-and \ref{sec:comparison} relates it to other similar systems in existence.
+based on while putting it into this rapidly evolving context. \ref{sec:curt}
+introduces the base system of the implementation, \curt{}, and
+\ref{sec:comparison} relates it to other similar systems in existence.
 \ref{sec:extension} then goes on to explain where the \curt{} system lacks in
@@
 basic checks for \emph{consistency} and \emph{informativity} on the resulting
-formulae -- within the bounds of the tools it uses.
+formulae -- within the bounds of the tools it uses\footnote{\curt{} makes use
+of (among others) the \pn{Otter} theorem prover and the \pn{mace2} model
+builder. During the process of the implementation of section
+\ref{sec:implementation}, it was ported to also make use of the \pn{Prover9}
+theorem prover.}.
 It can thus detect incoherent discourse and would reject a sequence of
@@
 implementation of this paper are meant as a replacement and enhancement for the
-latter feature. Their introduction will also happen in an incremental manner and new features
-are introduced one at a time during section \ref{sec:implementation}.
+latter feature. 
+%Their introduction will also happen in an incremental manner and new features
+%are introduced one at a time during section \ref{sec:implementation}.
 
@@
 
-The purpose of the former is in most cases twofold, although present
-\abbr{QA} systems vary in the functionality they offer
-and results they actually deliver. One goal is to enable their users to
-query for information without the need for a special query language or at least
-using a query language that would be easy and natural to use even for the
-uninitiated user. This way the user can pose direct questions to a system
-instead of having to search for a keyword to occur in the plain text of a
-document, like they would have to do with traditional information
-retrieval systems. The second feature provides \abbr{QA} systems with another important
-distinction, viz. the ability to utilize ontologies in order to relate query
-terms akin, but not equal to words in target documents and thus deliver
-satisfactory results even for fuzzy queries.
+The purpose of the former is in most cases twofold, although present \abbr{QA}
+systems vary in the functionality they offer and results they actually deliver.
+One goal is to enable their users to query for more complex information without
+the need for a special query language or at least using a query language that
+would be easy and natural to use even for the uninitiated user. The second
+feature provides \abbr{QA} systems with another important distinction, viz. the
+ability to utilize ontologies in order to relate query terms akin, but not equal
+to words in target documents and thus deliver satisfactory results even for
+fuzzy queries. This way the user can pose direct questions to a system instead
+of having to search for a keyword to occur in the plain text of a document, like
+they would have to do with traditional information retrieval systems.
 
@@
 
+\cite{webber}, argue that \abbr{QA} systems would undoubtedly benefit from
+introducing inferencing to their capability set. They analyse some of the
+typical \abbr{QA} tasks and give examples to show that \emph{reasoning},
+i.e. logical inferencing on the unstructured text that underlies modern
+\abbr{QA} systems would enhance their functionality.
+
+Interestingly, early attempts at providing a software capable of answering
+questions in natural language were mostly coupled with the usage of semantic
+analysis rather than statistical methods.  \cite{chat} are among  the first to
+attempt creating a question answering system with reasonable coverage. Their
+effort has been focused on the geographical domain, and their implementation,
+dubbed \pn{CHAT-80}, is written entirely in \prol. The lack of hardware
+resources of that time forced the system to be written with efficiency in mind.
+Moreover, it had to maintain a statical database of facts which was
+hand-crafted. As such, it falls more into the category of question answering
+systems than dialogue systems. Within its domain, \pn{CHAT-80} can deliver quite
+impressive results for different types of questions, including questions about
+the nature of relations between objects in its database.  This way, the system
+can answer a user's question as \example{Which country is borders at least two
+oceans?} without having a list of countries that border two oceans at its
+disposal.
+
 \cite{prontoqa} introduce the \pn{Pronto QA System}, which couples traditional
-corpus-based \abbr{QA} techniques with deep semantic parsing based on
-\term{Discourse Representation Theory}. Unfortunately, there seems
-to be no working demo to date. The homepage of the ongoing efforts of Johan Bos
-and his team can be found at the following address:
+corpus-based \abbr{QA} techniques with deep semantic parsing, utilizing
+\term{Discourse Representation Theory} as a data structure. Unfortunately, there
+seems to be no working demo to date. The homepage of the ongoing efforts of
+Johan Bos and his team can be found at the following address:
 \href{http://svn.ask.it.usyd.edu.au/trac/candc/}{\url{http://svn.ask.it.usyd.edu.au/trac/candc/}}.
 
-\cite{tenCate} present a tableaux method for resolving questions and also
-propose an algorithm in \prol{} as an implementation. Their approach is a more
-theoretical one, and they do not provide any particular implementation of their
-work. Although their methods seems promising, to say the least, it was chosen to
-avoid the implementation of a specialized inferencing system. \cite{bosgabdsil}
-mention the combinatorial explosion that lurks behind the treatment of questions
-and this is a considerable threat to any implementation of deep semantic
-\abbr{QA} systems. By using external inferencing tools, an implementation of
-the semantics of questions can focus on more essential issues than the
-resolution of formulae. % too offending. fixme
-
 Discourse Representation Theory is also used in \cite{bosgabdsil}, who present
-an approach to question answering quite similar to what is presented in this
-paper. In contrast to \cite{tenCate}, who use their own internal inferencing
-method to resolve questions, Bos and Gabdsil also use external inferencing very
-similar to the \curt{} system. Their results will be discussed later on, and
-explicitly compared to the present results.
-
-An attempt to bring inferencing to the world of \abbr{QA} systems can be found
-in \cite{webber}.
-
-Interestingly, earlier attempts at providing a software capable of answering
-questions in natural language were mostly coupled with the usage of
-semantic analysis instead of statistical methods.
-\cite{chat} are among  the first to attempt creating a question answering system
-with reasonable coverage. Their effort has been focused on the geographical
-domain, and their implementation, dubbed \pn{CHAT-80}, is written entirely in
-\prol. The lack of hardware resources of that time forced the system to be
-written with efficiency in mind. Moreover, it had to maintain a statical
-database of facts which was hand-crafted. As such, it falls more into the
-category of question answering systems than dialogue systems.
-
-Today it is clear that statistical question answering systems make inroads to
-the search market, and not many doubt that they are bound to eventually replace
-current document/term-based search engines. Even though \abbr{QA} systems also
-operate on the domain of answering questions, their approach to resolving them
-differs in most aspects from that of dialogue systems. The latter have a much
-higher obligation to correctness, since their results are typically not subject
-to peer review and serve an entirely different purpose.
+an approach to question answering to a certain extend similar to what is pursued
+in this paper. They choose to pursue the theoretical background compiled by
+\cite{gs:sqpa} and several later publications, including \cite{gs:q}. Moreover,
+they also use external first-order inferencing tools to perform computations on
+the consistency and informativity of the system's DRS models, which for this
+purpose are first translated to first order logic.  Their results will be
+discussed later on, and explicitly compared to the present results.
+
+\cite{tenCate} present a tableaux method for resolving questions via inferenceand also
+propose an algorithm in \prol{} as an implementation. Although their methods
+seems promising, it was chosen to
+the present approach will avoid the implementation of a specialized inferencing
+system and rely on external inferencing tools.
+
+Today it is clear that question answering systems make inroads to the search
+market, and not many doubt that they are bound to eventually replace current
+document/term-based search engines. Dialogue systems, are still not capable of
+convincing a larger user base. Among other things, their representation of
+knowledge and capability to interact with stored information is still considered
+to be lacking.  Even though \abbr{QA} systems also operate on the domain of
+answering questions, their approach to resolving them differs from that of
+dialogue systems. The latter have a much higher obligation to correctness, since
+their results are typically not subject to peer review and serve an entirely
+different purpose.
 
@@
 
-\curt{} is a prototype dialogue system and as such it is based on an entirely different set of
-premises than \abbr{QA} systems. Moreover, it differs from \pn{Godot} in some
-major points, the most distinctive of which is that \curt{} only exists inside a
+\curt{} is a prototype teaching system, designed to showcase the possibilities
+of a full-fledged dialogue system. It differs from \pn{Godot} in some major
+points, the most distinctive of which is that \curt{} only exists inside a
 regular computer and has no access to sensory input. Also, \curt's
@@
 language. It does only understand basic extensional transitive and intransitive
-verbs and its support for question answering remains limited.
+verbs and its support for question answering is only rudimentary.
 Anaphora resolution still remains an open issue.
@@
 
-For this to work, the internal storage of \curt{} will have to be restructured
+For this to work, the internal storage of \curt{} had to be restructured
 significantly in order to deal with more than one possible world or situation.
-Its grammar will have to be enhanced so that it can handle \wh-embedding verbs
-like `know' and `believe' and its inferencing mechanisms are going to be
-employed in a different manner, to encompass a wider range of functionality.
+Its grammar had to be enhanced so that it can handle \wh-embedding verbs
+like \example{know} and \example{believe} and its inferencing functionality had
+to be ported to the new system.
+
+The next two sections explain, in order, the theoretical background and
+practical operation of \acurt{}.
 
@@
 
-The semantics of questions has traditionally been a very sophisticated field of
+The semantics of questions have traditionally been a very sophisticated field of
 linguistics,  philosophy and logics, with many different approaches arguing for
@@
 Instead, it was chosen to roughly follow the work by Groenendijk and Stokhof,
-who present one of the best-established and recognized theories in the field,
-and also lending from other places where deemed appropriate or practical. As
-such, the theoretical framework of \acurt{} is not going to be an exact
-realization of any particular approach to question semantics, but rather an
-accumulation of interesting concepts from different sources. The main sources,
-however, are \cite{gs:q} and \cite{gs:sqpa}.
+who present one of the best-established and recognized theories in the field.
+In parts, the framework of the next sections is lending from other places, where
+deemed appropriate or practical. As such, the theoretical framework of \acurt{}
+is not going to be an exact realization of any particular approach to question
+semantics, but rather an accumulation of interesting, and, most of all,
+practicable concepts from different sources. The main sources, however, are
+\cite{gs:q} and \cite{gs:sqpa}.
 
@@
   \term{question} will denote the semantic entity uttered by the
-  \term{interrogative act}, as it will be introduced in section
+  \term{interrogative act}, and is introduced in section
   \ref{sec:protoq}
@@
 \term{yes/no-interrogative} since only \emph{yes} and \emph{no} are possible
-answers to this interrogative. \ref{ex:const} on the other hand is considered a
-\term{\wh-question} or \term{constituent interrogative} since one of the
+answers to this interrogative. \ref{ex:const}, on the other hand, is considered a
+\term{\wh-question} or \term{constituent interrogative}, since one of the
 sentences' nominal constituents is replaced by a \wh-pronoun, prompting an
@@
 
-The following abbreviations will be used: 
-\abbr{Ty2} stands for two-sorted type logic\footnote{See \cite{gallin:ty2} for a
+The following abbreviations are used: 
+\abbr{Ty2}, for two-sorted type logic\footnote{See \cite{gallin:ty2} for a
 discussion and \cite{gs:sqpa} for its usage in describing the semantics of
-questions}, \abbr{IL} stands for intensional logic\footnote{As used by
+questions}, \abbr{IL}, for intensional logic\footnote{As used by
 \cite{ptq}. See also \cite{gamut:2} for an excellent introduction to \abbr{IL}.},
@@
 \acurt{} employs a formalism that departs from both \abbr{IL} and \abbr{Ty2} in
-some ways. Most notably, the notion of a \term{model} is changed as follows, in
-order to accommodate the question operator introduced further on, which demands
-a set of possible worlds to be available.
+some ways. Most notably, the notion of a \term{model} is changed as follows.
 
@@
 
+Burhans and Shapiro
+
 \section{Appendix}

### 2008-09-01 thesis/paper.tex
Preliminary final version of theory

@@
-\documentclass[a4paper]{article}
-
-%\usepackage[english]{babel}
+\documentclass[notitlepage,twoside,a4paper]{scrreprt}
 
@@
 \renewcommand{\thepage}{\roman{page}}
+\begin{titlepage}
   \maketitle
@@
   \end{center}
+\end{titlepage}
+\mbox{}
 \newpage
@@
 \SpruechleAufsagen{Aleksandar L. Dimitrov}
+\thispagestyle{plain}
 \newpage
@@
 \setcounter{page}{1}
-\section{Introduction}
+\chapter{Introduction}
 
@@
 
-\section{Theory}\label{sec:theory}
+\chapter{Theory}\label{sec:theory}
 
@@
 
-\subsection{Formalism: Models}\label{sec:formal}
+\section{Formalism: Models}\label{sec:formal}
 
@@
 
-As was already mentioned in \ref{sec:curt}, \curt{} is based on 
-the use of external tools. These tools are only capable of treating first order logic – an
-extensional framework. \prol{} on the other hand, is a language that can very well
-make use of higher order formalisms for calculation\footnote{The two most
-popular predicates in \prol{} used for implementing higher-order logic are
-\code{call/n} and \code{apply/3}. This paper is based on \pn{SWI-Prolog} which
-provides an implementation of the former, but not of the latter. In addition
-\pn{SWI-Prolog} also defines several higher order aggregate functions, including
-\code{maplist/4}, which will be used later on. See \cite{naish:prolhio} for a
-discussion of higher order functional programming in \prol.}.  While there exist
+As was already mentioned in \ref{sec:curt}, \curt{} is based on the use of
+external inferencing tools. These tools are only capable of treating first order
+logic – an extensional framework. \prol{} on the other hand, is a language that
+can very well make use of higher order formalisms for calculation\footnote{The
+two most popular predicates in \prol{} used for implementing higher-order logic
+are \code{call/n} and \code{apply/3}. This paper is based on \pn{SWI-Prolog}
+which provides an implementation of the former, but not of the latter. In
+addition \pn{SWI-Prolog} also defines several higher order aggregate functions,
+including \code{maplist/4}, which will be used later on. See
+\cite{naish:prolhio} for a discussion of higher order functional programming in
+\prol.}, but as was already mentioned, the implementation of separate
+inferencing mechanisms for \acurt{} will be avoided. While there exist
 inferencing tools for modal logic\footnote{ The interested reader is referred to
@@
 of them would be beyond the scope of this thesis.  It was therefore decided to
-avoid modal logic and implement a pragmatic approach to representing indices in
-order to increase the expressive power of the system.
-
-Some research on the logic of questions (sometimes also referred to as eroteric
+avoid modal logic as much as possible and implement a pragmatic approach to
+representing indices in order to increase the expressive power of the system.
+
+Most research on the logic of questions (sometimes also referred to as eroteric
 logic) suggests that the treatment of interrogatives demands its own formalism.
@@
 
-One of the more important differences to the `traditional', \abbr{FOL}, notion of a model, is
-that it was decided to assign every world a unique domain of discourse referents
-or \term{individuals}. Note that there is no requirement $\pair{i,\pair{D,F}},
-\pair{j,\pair{D',F'}} \in M \wedge D\neq D'$, and most of the time, a model's
-possible worlds will indeed share many of their individuals or have entirely
-equal domains, but diversion in the world's domains will sometimes be useful and
-unavoidable.
+One of the more important differences to the `traditional' \abbr{Ty2} notion
+of a model, is that it was decided to assign every world a unique domain of
+discourse referents or \term{individuals}, instead of having one global set of
+individuals. This follows the definition of
+\cite{gs:q}, and is generally referred to as \term{partitioning of the logical
+space}. \cite{tenCate} propose a
+quantification over the assignment function $g$ in order to generalize a
+partitioned logical space into a more standard \term{Ty2} model, however, this
+would also demand the introduction of an set of an interpretation functions $I$
+to assign elements of $W$ an extension, which is not necessary here because
+every possible world is, in fact, an extensional model.
+
+Note that there is no requirement $\pair{i,\pair{D,F}}, \pair{j,\pair{D',F'}}
+\in M \wedge D\neq D'$, and most of the time, a model's possible worlds will
+indeed share many of their individuals or have entirely equal domains, but
+diversion in the world's domains will sometimes be useful and unavoidable.
+%fixme: really?
 
@@
 \emph{do} have \stress{syntactic} types) ensure a correct application of
-unsaturated formulae. They belong in the domain of first-order predicate logic,
+unsaturated formulae. These formulae are for the most part expressions of \pn{FOL}
 with two enhancements, introduced in section \ref{sec:protoq} and \ref{sec:epistemic}
 
-\subsection{On the Proper Treatment of Questions}
-
-Question semantics are an interesting subproblem of modern research on natural
-language semantics. While traditional approaches based on \abbr{IL} and newer
-accounts also employ plain extensional logic, % todo: which ones 
-the treatment
-of questions seems to demand that we introduce a new kind of logic formalism.
-But as interesting topics go, the semantics of questions has attracted many
-different views on this issue.
-
-This section is going to elaborate on some of the approaches in greater detail
-and also explain \acurt's notion of a question in particular. Note that the
-latter differs from the literature, but most differences are superficial.
-
-\subsubsection{The Question of Domain}
-
-Often question semantics has been reduced to a problem of
-pragmatics, treating direct interrogatives as nothing more than a
-paraphrase of embedded interrogatives.\cite{tichy}
-even goes as far as claiming equality in semantic (truth) value between
-indicatives and interrogatives. The benefit of such an approach would be the
-strict reliance on traditional analyses, where interrogatives would only infer a
-different `attitude' in the speaker's utterance. Putting this attitude into
-context and making it yield the correct results would then be left to
-pragmatics.
+\section{On the Proper Treatment of Questions}
+
+The analysis of the meaning of questions poses an interesting task to modern
+research on natural language semantics. It has so far attracted many different
+theories on how to represent and compute with them.
+But as interesting topics go, the semantics of questions
+has attracted many different views on this issue.  It is important to first
+settle on what constitutes and characterizes a \emph{direct} question before one
+can turn to the treatment of \emph{embedded} questions and propositions. Most
+literature seems to agree that the latter is but a special case of the former,
+even though sometimes the notion of a direct question is abstracted to a
+\term{proto-question}.
+
+\subsection{The Question of Domain}
+
+Often question semantics has been reduced to a problem of pragmatics, treating
+direct interrogatives as nothing more than a paraphrase of embedded
+interrogatives.\cite{tichy} even goes as far as claiming equality in semantic
+(truth) value between indicatives and (embedded) interrogatives. His analysis
+permits omitting a new the introduction of a new phrasal category into the
+logical formalism that would otherwise be needed to account for questions.
+Interrogative sentences then merely infer a different `attitude' in the
+speaker's utterance. Putting this attitude into context and making it yield the
+correct results is then left to pragmatics, which treats them as interrogative
+acts in the setting of a dialogue.
 
 The issue is ultimately whether the answering of questions belongs in the realm
-of pragmatics or semantics. Tichy's approach has been challenged and proven
-unfeasible by \cite{gs:q}, who argue for a semantic interpretation of questions.
-They also propose certain adequacy criteria for theories of the semantics of
-questions in order to be able to compare different approaches made in the
-literature and also to defend their own solution. Their analysis is based on the
-introduction of another type of logical expression, an \term{interrogative}.
-In contrast to indicatives, interrogatives do not yield a truth value, but
-instead pose a \emph{question}. They will be formally introduced in section
-\ref{sec:protoq}.
-
-It is important to note that \cite{karttunen:1977} attempts a treatment of
-questions in a similar manner, but also \emph{denies} a difference in logical
-type for questions as such. According to his hypothesis, direct questions are a
-syntactic variant of indirect (embedded) questions and yes/no interrogatives are
-a ``syntactically degenerate variant'' of constituent interrogatives.
-
-\begin{quotes}A direct question can be treated as semantically equivalent to a
-certain kind of declarative sentence containing the corresponding indirect
-question embedded under  a suitable `performative' verb.\end{quotes}
-
-Using this strategy, Karttunen escapes the treatment of direct  questions and
-thus gives an approach for indirect (i.e. \emph{embedded}) questions  only.
-In this respect, his analysis is somewhat similar to the one presented in
-\cite{gs:sawhq}, but the latter goes on to challenge certain aspects of his
-hypothesis.
+of pragmatics or semantics. Tichy's approach, along with other, similar ones,
+has been challenged by \cite{gs:q}, who argue for a treatment of interrogatives
+on a semantic level, conceding a \emph{meaning} to interrogatives.  They also
+propose certain adequacy criteria for theories of the semantics of questions in
+order to be able to compare different approaches made in the literature and also
+to defend their own solution. Their analysis is based on the introduction of
+another type of logical expression, a \term{question}, which embeds a
+proposition (with possibly one or more free variables) under a question operator
+\term{?}.  In contrast to indicatives, questions do not yield a truth value, but
+are defined by means of their \term{possible answers}.  They are formally
+introduced in section \ref{sec:protoq}.
+
+%\cite{karttunen:1977} attempts a treatment of
+%questions in a similar manner, but also \emph{denies} a difference in logical
+%type for questions as such. According to his hypothesis, direct questions are a
+%syntactic variant of indirect (embedded) questions and yes/no interrogatives are
+%a ``syntactically degenerate variant'' of constituent interrogatives.
+
+%\begin{quotes}A direct question can be treated as semantically equivalent to a
+%certain kind of declarative sentence containing the corresponding indirect
+%question embedded under  a suitable `performative' verb.\end{quotes}
+
+%Using this strategy, Karttunen escapes the treatment of direct  questions and
+%thus gives an approach for indirect (i.e. \emph{embedded}) questions  only.
+%In this respect, his analysis is somewhat similar to the one presented in
+%\cite{gs:sawhq}, but the latter goes on to challenge certain aspects of his
+%hypothesis.
 
@@
 
-Their notion of questions will be extended for \acurt{} in order to account for
-yes/no interrogatives and larger sets of answers made possible by using model
-which allow for possible worlds.
-
-Constituent interrogatives also differ in what they ask for. One can assume the
-existence of at least three different types:
+Their notion of questions was modified for \acurt{} in order to account for
+yes/no interrogatives and larger sets of answers made possible by using a model
+representation which allows for possible worlds.
+
+\subsubsection{The Grammatical Domain of \acurt}
+
+Constituent interrogatives further differ in \emph{what} they ask for. One can
+assume the existence of at least three different types:
 
@@
   Butch (shot Vincent.)} or even more then one noun phrase at the same time, \example{Who had
-  what in the hash bar?\footnote{Multiple constituent interrogatives pose a
+  what in the hash bar?}\footnote{Multiple constituent interrogatives pose a
   somewhat greater challenge, since they come with an inherent ambiguity, the
   so-called \term{Baker Ambiguity}, as presented in \cite{baker:1970}.
-  Such ambiguities will not be concerned here.}}
+  Such ambiguities will not be concerned here.}
   \item PP-questions ask for prepositional constituent, e.g. \example{Where did
@@
 This classification of interrogatives can go on for quite some more entries,
-including local, modal, causal and all sorts of other questions. Out of all
-those categories, only NP-questions will be supported, with an option to
+including local, modal, causal and many sorts of other questions. Out of all
+those categories, only NP-questions are be supported by \acurt{}, with an option to
 integrate other types of interrogatives at a later stage.
+
+\subsubsection{Pragmatic Concerns}
 
@@
 of \term{answerhood}, i.e. what constitutes a valid answer to a given
-interrogative. The ground-laying work by \cite{hamblin:q} states that in knowing
-all possible answers to a question, one also knows the question itself. The
-notion of answerhood is going to be elaborated on in section \ref{sec:answerhood}
-
-One thing remains clear though, no matter what the notion of a question or
-finding an answer to it is; upon
-finding an answer, said answer has to be \emph{realized} in natural language
-using text or speech synthesis. When formulating an answer, many pragmatic
-considerations have to be taken into account.
-However, as the uttering of an embedded question does not prompt an
-answer\footnote{Unless, of course, the question is uttered as an
-\term{imperative} or \term{performative}-- but this is an entirely different
-class of problems that is not discussed here.}, the present
+interrogative. The different positions on this topic are elaborated on in
+section \ref{sec:answerhood}. Regardless of the concept of an optimal answer, it
+is clear that upon finding an answer, said answer has to be \emph{realized} in
+natural language using text or speech synthesis. When formulating an answer,
+many pragmatic considerations have to be taken into account.  However, as the
+uttering of an embedded question does not prompt an answer\footnote{Unless, of
+course, the question is uttered as an \term{imperative} -- but this is an
+entirely different class of problems that is not discussed here.}, the present
 work will not directly concern itself with a solution to this problem.
@@
 
-\subsubsection{The Semantics of Interrogatives: The ? Operator}\label{sec:protoq}
+\subsection{The Semantics of Interrogatives: The ? Operator}\label{sec:protoq}
 
@@
 present, only addresses the problem of direct constituent interrogatives and
-does neither treat yes/no interrogatives nor embedded questions. The
-interrogative operator, as it will be introduced shortly, is designed to cover
-the semantics of both types of questions simultaneously. It takes a proposition
-(a formula yielding a truth value as a result) but gives back a set of
-propositions. Computation on more than one proposition in parallel will be
-introduced into the \curt{} systems in section \ref{sec:possworlds}.  \acurt{}
+does not treat yes/no interrogatives or embedded questions. The interrogative
+operator, as it will be introduced shortly, is designed to cover the semantics
+of both types of questions simultaneously. It takes a proposition (a formula
+yielding a truth value as a result) and yields a set of propositions. \acurt{}
 represents embedded questions by embedding a special construct, a
-\term{proto-question}, under verbs of certain types (which will be discussed
-later on). Since the yes/no interrogative is considered a special case of the
-constituent interrogative the semantic formulae produced by the grammatical
-back end of \acurt{} do not differ for the two cases.
+\term{proto-question}, under verbs of certain types (discussed in section
+\ref{sec:ontology}). Since the yes/no interrogative is considered a special case of the
+constituent interrogative the semantic formulae produced by the grammatical back
+end of \acurt{} show no difference in structure for the two cases.
 
 \begin{definition}\label{def:qop}
-  \stress{The Question Operator}
+  \stress{The Question Operator}, following \cite{gs:q}
   \begin{enumerate}
@@
     considered a question associated wit a yes/no interrogative.
-    \item \label{def:outermost}The question operator does \emph{not combine} with any other operator.
+    \item * \label{def:outermost}The question operator does \emph{not combine} with any other operator.
   \end{enumerate}
@@
 
-Note that the last clause of definition \ref{def:qop} enforces the question operator to
-always be the last operator of a formula. This constraint will be overridden
-during parsing of questions to produce forms that do have nested questions.
-When calculating the models and readings of a formula, however, these operators
-will be resolved first. This is done in order to increase the efficiency during
-parsing, since the syntactic parser will sometimes have to backtrack.
-
-As was already mentioned, \pn{Helpful Curt} has a certain albeit limited
+Note that the last clause of definition \ref{def:qop} enforces the question
+operator to always be the last operator of a formula. This constraint will be
+overridden during parsing of questions to produce forms that do have nested
+questions.  When calculating the models and readings of a formula, however,
+these operators will be resolved first. This is done in order to increase the
+efficiency during parsing, since the syntactic parser will sometimes have to
+backtrack. In fact, the generation of \term{logical forms} and their
+interpretation are always two distinct steps in \acurt.
+
+As was already mentioned, \pn{Helpful Curt} has a certain, albeit limited
 support for evaluating NP constituent interrogatives. Blackburn and Bos adopt a
 sort of pseudo-notation of the question operator in what they call an
-extra-logical form which does not itself yield a truth value, but is
-internally evaluated in a special case.
-
-They treat a question as a \emph{quantification
-over a certain individual $X$, for which $R$ and $S$ must hold.} Thus a parse of
-a sentence like \ref{ex:bbqueexa} and \ref{ex:bbqueexb} would yield the representation given in
-\ref{ex:bbqueparsea} and \ref{ex:bbqueparseb}.
-
-\ex. \a. \label{ex:bbqueexa} Which boxer likes Jody?
+extra-logical form. It does not yield a truth value, but is internally evaluated
+in a special case.
+
+\curt{} sees a question as a \emph{quantification over a certain individual
+$X$}, for which two conditions, $R$ and $S$, must hold. \cite{bosgabdsil} call
+$R$ and $S$ the \term{domain} and \term{body} of the question. The domain of a
+question is usually ontological information, posing a restriction on the
+quantified variable. It is selected according to the \wh-pronoun that triggered
+it. So \example{who} prompts the domain to contain $person(x)$, whereas
+\example{which}, used as a quantifier in a phrase as \example{which boxer}, would
+impose the restriction $boxer(x)$. \ref{ex:bbqueexa} and \ref{ex:bbqueexb} give
+an example of how \curt{} parses constituent interrogatives.
+
+\ex.\label{ex:bbqueexa}
+\a. Which boxer likes Jody?
+\b. $?x.boxer(x) \wedge like(x,jody)$
 \b. \label{ex:bbqueparsea} \code{que(X, boxer(X), like(X, jody))}
 
-\ex. \a. \label{ex:bbqueexb} Who shot Vincent?
+\ex.\label{ex:bbqueexb}
+\a. Who shot Vincent?
+\b. $?x.person(x) \wedge shoot(x, vincent)$
 \b. \label{ex:bbqueparseb} \code{que(X, person(X), shoot(X, vincent))}
+
+Similarly, yes/no interrogatives, which are new to the \curt{} system, are
+parsed using the same sort of operator:\footnote{The actual representation in
+the code will differ a little in order to work around a bug in the system.}
+
+\ex.
+\a. (whether) Vincent snorts.
+\b. $?.snort(Vincent)$
+\b. \lstinline!que([], [], snort(vincent))!
+
+\ex. \a. (whether) Mia or Jody dances.
+\b. $?.dance(mia) \vee dance(jody)$
+\b. \lstinline!que([], [], or(dance(mia),dance(jody)))!
 
@@
 Groenendijk and Stokhof and also later approaches by \cite{g:is}. \acurt{}
-bridges the gap between both worlds.
-
-\subsubsection{What Answers the Question?}\label{sec:answerhood}
+aims to bridge the gap between both worlds.
+
+\subsection{What Answers the Question?}\label{sec:answerhood}
 
@@
 the relation of the question to its answer or answers with respect to the
-current discourse. It is important to note that an answer other than `no'
+current discourse. Note that an answer other than `no' or
 `nobody' cannot occur without a certain discourse or background knowledge already being present
-for the answer to be evaluated against.
+for the answer to be evaluated.
 
@@
 For yes/no interrogatives this means, that if the question is $?.\varphi$, the
-answer two it is either $\varphi$ or $\neg\varphi$. For constituent
+answer to it is either $\varphi$ or $\neg\varphi$. For constituent
 interrogatives, this question is not so easily answered.  There is one central
-issue to be addressed first: the issue of the definition of answerhood: should
-the set of answers be an \emph{exhaustive} set of answers, including the
-negative ones, or just the set of possible \stress{true} answers?
+issue to be addressed first: should the set of answers be an \emph{exhaustive}
+set of answers, including the negative ones, or just the set of possible
+\emph{true} answers?
 
 The following situation might serve as an example: Suppose Mia and Vincent both
-play air guitar. Butch and Marcellus do not play air guitar. When answering the
+play air guitar. Butch and Marsellus do not play air guitar. When answering the
 question \example{Who plays air guitar?}, one can observe three degrees of
@@
 guitar, but also of Butch, she would still know all possible true answers, and,
-according to this analysis, a correct answer.
-
-Groenendijk and Stokhof argue for two more degrees of exhaustiveness. The first
-one would require the answer not only to contain all possible \emph{true}
+according to this analysis, have a correct answer.
+
+\cite{gs:sqpa} argue for two more degrees of exhaustiveness. The first
+one requires the answer not only to contain all possible \emph{true}
 propositions, but also to \emph{not contain any \stress{false} propositions}. So
@@
 
-The second and strongest degree of exhaustiveness would
-require any set of answers not only to contain the set of propositions that
-hold, but also the set of propositions that \emph{do not hold}. In this case, for
-the sentence \example{Yolanda knows who plays air guitar.} to be true, the
-sentence \example{Yolanda knows who does not play air guitar.} would
-automatically follow.
-
+\footnotetext{
 There even more ways of categorizing the number of answers to a question. So far
@@
 for John to know in order for him to know where to actually buy one. This has
-been regarded as the \term{mention-some} reading.
-
-Even after having settled on the \emph{number} and \emph{value} of propositions
-that have to serve as an answer to a question, one still might be concerned with
-their kind. Taken the previously given example, if Butch and Marcellus were
-all boxers in this situation, would not an answer such as \example{Every boxer.}
-suffice? \cite{rooy:qar} elaborates on the many aspects of different analyses
-regarding the relevance of answers. Suffice to say, \acurt{} is only content
-with \emph{maximally specific} resolutions to answers.
-
-Moreover, as \acurt{} will only have to deal with answers to questions
+been regarded as the \term{mention-some} reading.}
+
+The second and strongest degree of exhaustiveness 
+requires any set of answers not only to contain the set of propositions that
+hold, but also the set of propositions that \emph{do not hold}. In this case, for
+the sentence \example{Yolanda knows who plays air guitar.} to be true, the
+sentence \example{Yolanda knows who does not play air guitar.} would
+automatically follow.
+
+\subsubsection{Combinatorial Problems}
+
+Thus the maximum degree of exhaustiveness bears several problems. The most
+obvious one is that of a considerable \emph{combinatoric explosion}\footnotemark. In the
+previous example, containing only four individuals, the question
+$?x.play\_air\_guitar(x)$ evaluates to a total set of $2^4$ propositions:
+\example{Nobody plays air guitar.}, \example{Only Butch plays air guitar},
+\example{Only Mia plays air guitar},\ldots{}, \example{Only Butch and Mia play
+air guitar},\ldots{}, \example{Everybody plays air guitar}. Generally, every
+question will evaluate to $2^{|D|}$ propositions. As \cite{bosgabdsil} show, to
+find a suitable answer to a given partition of logical space (i.e. a
+\emph{question}), the logical inference tools would have to perform consistency
+checks on the union \emph{every single} possible answer with the discourse.
+
+\footnotetext{
+There is at least one more problem with this analysis although \cite{gs:q} hint
+at an easy solution for this. Suppose Yolanda does not
+know Butch, but she knows Mia and Vincent, and that they play air guitar. But since
+she does not know about Butch, she has no chance of knowing a \emph{most
+specific} answer. Groenendijk and Stokhof's analysis thus requires every
+individual to have exhaustive knowledge about the discourse domain.}
+
+\cite{bosgabdsil} go on to propose a method for taming the computational
+complexity of Groenendijk and Stokhof's analysis, but they do so at the cost of
+a strong account for exhaustivity. Their system reduces the size of the
+partition to essentially four propositions: \example{Everybody plays air
+guitar.}, \example{Somebody plays air guitar.}, \example{Somebody does not play
+air guitar.}, \example{Nobody plays air guitar.} Now the inferencing task has
+shrunk down considerably, but it does not even require \emph{every true answer}
+to be known in the case that the body of the question does not hold for at least
+one individual, which will, for non-trivial domains, almost always be the case.
+
+As \acurt{} will only have to deal with answers to questions
 internally, it will also only regard maximally exhaustive interpretations of
 answers.
-Correctly answering a direct question with respect to relevance and other
+
+Furthermore, correctly answering a direct question with respect to relevance and other
 pragmatic factors is a task left as a task for turning \curt{} into a complete
@@
 
-\subsection{Embedded Questions}\label{sec:embed}
+\section{Embedded Questions}\label{sec:embed}
 
 After having agreed on a treatment of interrogatives, the next step is to
-construct a framework for embedding the result under certain verbs. The main
-issues of this section will thus be focused on providing a classification of
-embedding verbs and elaborating on the details of the relationship between
-embedded and embedding constituents of an utterance.
-
+construct a framework for embedding the result under certain verbs.  This
+section provides a classification of embedding verbs and elaborates on the
+details of the relationship between embedded and embedding constituents of an
+utterance.
+
+\subsection{Syntax}
 As was already mentioned, clause \ref{def:outermost} of definition \ref{def:qop}
@@
 \b.  \Tree
-[.$t$\\$know(?X_1.snort(X_1))(Mia)\wedge ?X_1.snort(X_1)$ [.$T$ { $\lambda X.X(Mia)$\\Mia } ]
-[.$IV$\\$\lambda X.know(?X_1.snort(X_1))(X)\wedge ?X_1.snort(X_1)$ [.{$IV/\bar{t}$} 
-{ $\lambda S.\lambda X.[  \lambda Y.[ know(Y)(X) \wedge Y ](S) ]$\\knows } ]
+[.$t$\\$know(?X_1.snort(X_1))(Mia)$ [.$T$ { $\lambda X.X(Mia)$\\Mia } ]
+[.$IV$\\$\lambda X.know(?X_1.snort(X_1))(X)$ [.{$IV/\bar{t}$} 
+{ $\lambda S.\lambda X.[  \lambda Y.[ know(Y)(X)](S) ]$\\knows } ]
 [.{$\bar{t}$\\$?X_1.snort(X_1)$ } [.$\bar{t}/t$ ? ] [.{ $t$\\$snort(X_1)$ }
 [.$T$ { $X_1$\\who } ]
-[.$IV$ { $\lambda X.snort(X)$\\snorts } ] ] ] ]!\qsetw{4cm} ]\normalsize
-\b. $know(?X_1.snort(X_1))(Mia)\wedge ?X_1.snort(X_1)$\label{ex:formula}
+[.$IV$ { $\lambda X.snort(X)$\\snorts } ] ] ] ]!\qsetw{5cm} ]\normalsize
+\b. $know(?X_1.snort(X_1))(Mia)$\label{ex:formula}
 
@@
 
-The reason for $?$ to be embeddable within more complex formulae is merely one
-of convenience, since one could transform $?X.\varphi(X)$ to mean
-
-\[\forall w \in W_M [ \forall x \in D_w [ \varphi(w)(x) = 1 ] ]\]
-
-assuming $W_M$ is the set of possible worlds in a model $M$, but this would
-require modifications to the current way of evaluating predicates with \curt{}
-as well as the definition of a quantification over possible worlds, which is
-currently not supported and would require a special treatment. Furthermore, in
-the case that the explicit formula does not yield any results, the question
-should return its correct answer
-
-\[ \neg\exists x [\varphi(x)] \]
-
+The reason for $?$ to be embeddable within more complex formulae is not merely one
+of convenience, since it cannot be evaluated without access to a particular
+world's set of individuals $D$. This would mix semantic interpretation with
+the parsing process and such side-effects were not deemed desirable.
 Instead, before calling its inferencing tools \acurt{} will look out only for
@@
 [.$t$\\$know(?.snort(Vincent))(Mia)\wedge ?.snort(Vincent)$ [.$T$ { $\lambda X.X(Mia)$\\Mia } ]
-[.$IV$\\$\lambda X.know(?.snort(Vincent))(X)\wedge ?.snort(Vincent)$ [.{$IV/\bar{t}$} 
-{ $\lambda S.\lambda X.[  \lambda Y.[ know(Y)(X) \wedge Y ](S) ]$\\knows } ]
+[.$IV$\\$\lambda X.know(?.snort(Vincent))(X)$ [.{$IV/\bar{t}$} 
+{ $\lambda S.\lambda X.[  \lambda Y.[ know(Y)(X) ](S) ]$\\knows } ]
 [.{$\bar{t}$\\$?.snort(Vincent)$} [.$\bar{t}/t$ whether ] [.{ $t$\\$snort(Vincent)$ } 
@@
 \normalsize
-\b. $know(?.snort(Vincent))(Mia)\wedge ?.snort(Vincent)$
+\b. $know(?.snort(Vincent))(Mia)$
 
@@
 
-This would again demand explicit quantification over worlds and a different
-treatment of predicates. Instead, it can be treated with the same procedure as
-before when using an embedded question operator.
-
-\subsubsection{Types of Facts, Assertions and Interrogatives, Towards an
+Obviously, the partitions for yes/no interrogatives is much smaller than it is
+for constituent interrogatives and it does not need access to a particular model
+in order to form its partition. Yet this is not reason enough to give up the
+more general approach already described.
+
+\subsection{Types of Facts, Assertions and Interrogatives, Towards an
 Ontology of Embedding}\label{sec:ontology}
@@
 
-From these examples it seems that, again, there is no difference between
-constituent interrogatives and yes/no interrogatives concerning the position
-they can occupy. Furthermore, it seems that the three different verbs behave
-differently: `know' can take both questions and propositions, `ask' can only
-take questions and `believe' can only take propositions.
+These examples show that there is no difference between
+constituent interrogatives and yes/no interrogatives concerning the syntactic
+position they can occupy. Furthermore, it seems that the three different verbs
+behave differently: `know' can take both, questions and propositions, `ask' can
+only take questions, and `believe' can only take propositions.
 
@@
 
-The matter is more complicated, however, with \term{factive predicates}, where
-one might wish to establish another distinction. The example
-\ref{ex:whembed} already used a denotation for \example{know}, which will now be
-introduced formally:
-
-\begin{postulate}\label{def:know}
-  Knowledge necessitates the belief of a proposition $p$ by the knowing
-  individual as well as the verity of $p$.
-
-  \sem{know} = $\lambda p\lambda y.[know(y)(p)\wedge p]$
-\end{postulate}
-
-While the argumentation about exhaustiveness in section \ref{sec:answerhood}
-already drew upon this statement, postulate \ref{def:know} puts this into a
-more formal perspective. The postulate is based on the fact that, clearly, one
-can not \emph{know} something that is not true. In that case one's knowledge
-would be wrong, and one would only \emph{believe to know}. This also holds for
+The matter is more complicated with \term{factive predicates}, where
+one might wish to establish another distinction. Recall the example from section
+\ref{sec:answerhood}. In order for Yolanda to \emph{know} who plays air guitar,
+her assumed knowledge and the real world have to agree on this fact. If Yolanda
+thought Butch played air guitar, but in reality he did not, clearly Yolanda
+wouldn't know who plays air guitar. She would only \emph{believe to know}.
+This also holds for
 verbs like \example{discover}, \example{reveal} and \example{forget}. But does
@@
 in the same \emph{syntactic} category, since one can \example{predict/report who will
-win/wins the elections}, they seem to demand different \emph{syntactic}
+win/wins the elections} and also \example{predict/report that a man will
+win/wins the elections}, they seem to demand different \emph{semantic}
 treatment, since both, predictions on, and reports of the results of elections are
@@
 
-\ex. \label{ex:ginzburg}
+\ex. 
 \a. Jill believed a certain hypothesis. Hence, Jill believed that hypothesis
 to be true.
-\b. Jill discovered a certain hypothesis. It \emph{does not follow} that Jill
+\b. \label{ex:ginzburg}Jill discovered a certain hypothesis. It \emph{does not follow} that Jill
 discovered that hypothesis to be true.
@@
 
-\subsubsection{Modeling Knowledge}\label{sec:epistemic}
-
-We have seen that some embedding verbs may not be interpreted without
-implementing a higher order formalism that is able to model the notion of
-\emph{possible} and \emph{accessible} worlds. While foobar %fixme! Nelken &
-% Francez 2001
-argue for an extensional treatment of questions, they already % n & f 2006
-retract their claims and admit that a ``minimal account of intensionality'' has
-to be assumed. While they propose a two-world system later adopted in
-\cite{g:is}, this is not going to suffice if we want to model each individual's
-knowledge in order to be able to discriminate people's knowledge and beliefs.
-%fixme bla bla
-
-The underlying theory for 
-TODO Discussion of Lewis' paper.
-
-\subsection{Treating Disjunction}
-
-Consider the following example dialogue with the \pn{Helpful Curt}
-system\footnote{In a real conversation with \curt, it will usually inform the
-user about the activity of external inferencing tools. This has been omitted and
-marked by \code{[…]}.}:
-
-\begin{verbatim}
-  > Mia or Jody dances.
-  […]
-  Curt: OK.
-
-  > Who dances?
-  […]
-  Curt: This question makes sense!
-
-  Curt: maybe mia
-\end{verbatim}
-
-Clearly, \curt{} should give a full answer in this situation, but instead it
-just guesses, and does not reply with something along the lines of \example{Either
-Mia or Jody}. Unfortunately, this result does not suffice for building a system
-that can properly treat embedded questions, since it does not reflect all
-available information about the discourse properly.
-
-A solution to this will again require the presence of the concept of
-\term{possible worlds}. When uttering the phrase \example{Mia or Jody dances.}
-then this can be understood as creating \emph{two possible scenarios}, one in
-which there is an individual named \example{Mia}, and this individual dances,
-and one in which there is an individual named \example{Jody} and the same holds
-for this individual. In effect, this can be translated to two possible worlds
-being present in the current discourse.
+\subsection{Modeling Knowledge}\label{sec:epistemic}
+
+So far, it has been suggested that questions can be interpreted as partitions of
+the logical space of the current discourse model. Furthermore, a means for
+evaluating questions has been presented, again, with respect to the current
+discourse model. Then these questions, can be embedded under certain verbs in
+certain ways. But what is the assertion a sentence as \example{Mia knows who
+snorts.} makes?
+
+When treating issues like this, the link between philosophy and language becomes
+most obvious. One can assume that the knowledge about the world will differ,
+sometimes greatly, between different individuals. One will also have to assume
+that every person has their own beliefs. In \emph{knowing} or \emph{believing},
+or even \emph{imagining}, one can create an image of what it could possibly look
+like.
+
+This concept of \term{possible worlds} probably dates back to Leibniz' theory,
+who assumed that we live in the best of such possibilities. \cite{lewis} gives
+an elaborate description of what the construction of such worlds might look like
+and what their nature is -- or might be. He states that every individual
+holds two sets of \term{epistemic} and \term{doxastic} alternative worlds. The
+former is what an individual thinks is true, the latter accounts for its
+beliefs.
+
+\wh-embedding verbs embed sets of propositions, and they establish a relation
+between those propositions and their subject, one or more individuals. Depending
+on the type of verb, it can be that this individual wonders about a certain
+fact, reports it,  knows about it or believes it. In the latter two cases,
+something more happens: the proposition is asserted to be a fact of the
+individual's \emph{epistemic} or \emph{doxastic} worlds, which are worlds, just
+like any other possible world.
+
+Discussing whether knowledge has to be \emph{true}, for any particular
+definition of truth, or whether beliefs have to be consistent, or even if an
+individual has to know about itself in its own alternative worlds, is a moot
+point from a computational perspective. Suffice to say, that \acurt{} treats its
+own models as representing the truth. Furthermore, it introduces two new sets of
+worlds, accounting for epistemic and doxastic possibilities, with respect to
+worlds of its own model and individuals thereof. The sets $\mathcal{E}$ and
+$\mathcal{D}$ are thus defined in the following way:
 
 \begin{definition}
-  Disjunction
-
-  For any formula $\rho$ holding a disjunction, this formula can be expressed as a set
-  of the constituents of the disjunction
-
-  \[ \rho = \varphi \vee \psi, \textrm{ then } \rho' = \set{\varphi, \psi}\]
-
-  %Every other operator or function can take a set of propositions of the type of
-  %$\rho'$ to produce a set of propositions containing \emph{all possible
-  %permutations} of $\rho'$. For every possible operator $\sem{$\cdot$}$ with arity
-  %$n$ and arguments $\sem{$\cdot$}(\set{\varphi,\psi})$
-
-  todo: more generic: just go from 1 to j, which is rho, then from j+1 to n
-
-  \begin{eqnarray*}
-    %\sem{$\cdot$}(\rho') \mapsto \set{\varphi,\psi}\\
-    %\sem{$\cdot$}(\sigma_1,\dots,\sigma_j,\rho )
-    \varphi
-    %\sigma \wedge \set{\varphi, \psi} \mapsto \set{\sigma\wedge\varphi,\sigma\wedge\psi}\\
-    %\sigma \to \set{\varphi, \psi} \mapsto \set{\sigma\to\varphi,\sigma\to\psi}\\
-  \end{eqnarray*}
+  Epistemic and Doxastic Models
+
+  An epistemic or Doxastic model is a triplet \pair{d,i,w'}, where
+  \begin{itemize}
+    \item i is a reference to a possible world $w = \pair{D,F,R}$ of the main
+    model of the system, $\mathcal{M}$
+    \item $d \in D$
+    \item $w'$ is a possible world, defined the same way as before: $w' =
+    \pair{D',F',R'}$
+  \end{itemize}
 \end{definition}
 
-\subsection{Inquisitive Semantics}\label{sec:altq}
+Whenever the relation of knowledge between an individual and a proposition is
+encountered, the system records it within the epistemic worlds of the
+individual it holds for, also noting the indexes of the possible worlds the
+proposition holds in. Almost the same will be done for beliefs, which end up in
+the doxastic worlds of an individual, but are not required to be true of a
+possible world in \curt's knowledge.
+
+\acurt{} is an experimental program. As such, it can serve as a ground for
+experimentation to find out if this definition is better suited, or any other,
+for representing the \example{true} state of affairs.
+
+%\section{Treating Disjunction}
+
+%Consider the following example dialogue with the \pn{Helpful Curt}
+%system\footnote{In a real conversation with \curt, it will usually inform the
+%user about the activity of external inferencing tools. This has been omitted and
+%marked by \code{[…]}.}:
+
+%\begin{verbatim}
+  %> Mia or Jody dances.
+  %[…]
+  %Curt: OK.
+
+  %> Who dances?
+  %[…]
+  %Curt: This question makes sense!
+
+  %Curt: maybe mia
+%\end{verbatim}
+
+%Clearly, \curt{} should give a full answer in this situation, but instead it
+%just guesses, and does not reply with something along the lines of \example{Either
+%Mia or Jody}. Unfortunately, this result does not suffice for building a system
+%that can properly treat embedded questions, since it does not reflect all
+%available information about the discourse properly.
+
+%A solution to this will again require the presence of the concept of
+%\term{possible worlds}. When uttering the phrase \example{Mia or Jody dances.}
+%then this can be understood as creating \emph{two possible scenarios}, one in
+%which there is an individual named \example{Mia}, and this individual dances,
+%and one in which there is an individual named \example{Jody} and the same holds
+%for this individual. In effect, this can be translated to two possible worlds
+%being present in the current discourse.
+
+%\begin{definition}
+  %Disjunction
+
+  %For any formula $\rho$ holding a disjunction, this formula can be expressed as a set
+  %of the constituents of the disjunction
+
+  %\[ \rho = \varphi \vee \psi, \textrm{ then } \rho' = \set{\varphi, \psi}\]
+
+  %%Every other operator or function can take a set of propositions of the type of
+  %%$\rho'$ to produce a set of propositions containing \emph{all possible
+  %%permutations} of $\rho'$. For every possible operator $\sem{$\cdot$}$ with arity
+  %%$n$ and arguments $\sem{$\cdot$}(\set{\varphi,\psi})$
+
+  %todo: more generic: just go from 1 to j, which is rho, then from j+1 to n
+
+  %\begin{eqnarray*}
+    %%\sem{$\cdot$}(\rho') \mapsto \set{\varphi,\psi}\\
+    %%\sem{$\cdot$}(\sigma_1,\dots,\sigma_j,\rho )
+    %\varphi
+    %%\sigma \wedge \set{\varphi, \psi} \mapsto \set{\sigma\wedge\varphi,\sigma\wedge\psi}\\
+    %%\sigma \to \set{\varphi, \psi} \mapsto \set{\sigma\to\varphi,\sigma\to\psi}\\
+  %\end{eqnarray*}
+%\end{definition}
+
+\section{Inquisitive Semantics}\label{sec:altq}
 
@@
 
-\section{Implementation} \label{sec:implementation}
+\chapter{Implementation} \label{sec:implementation}
 
@@
 
-\subsection{Advertent Curt}
+\section{Advertent Curt}
 
@@
 
-\subsection{The Syntax of Embedded Questions}
+\section{The Syntax of Embedded Questions}
 
@@
 
-The grammatic rules are extended by the introduction of three new categories,
-\code{whemb}, \code{tbar} and \code{ivtbar}.
-
-The \code{whemb} and \code{ivtbar} categories have a \code{type} parameter, that dictates
-their behaviour, although their respective type systems never cooperate with
-each other. The type of the \code{whemb} category can hold the values
-\code{propos}, for a proposition embedding complementizer, like \example{that},
-and \code{question}, for a complementizer that turns a proposition into a yes/no
-interrogative, like \example{if} and \example{whether} (listing \ref{whemb}).
-
 \begin{lstlisting}[label=whemb,caption={Syntactic and Lexical Rules for Complementizers},float]
@@
 
+The grammatic rules are extended by the introduction of three new categories,
+\code{whemb}, \code{tbar} and \code{ivtbar}.
+
+The \code{whemb} and \code{ivtbar} categories have a \code{type} parameter, that dictates
+their behaviour, although their respective type systems never cooperate with
+each other. The type of the \code{whemb} category can hold the values
+\code{propos}, for a proposition embedding complementizer, like \example{that},
+and \code{question}, for a complementizer that turns a proposition into a yes/no
+interrogative, like \example{if} and \example{whether} (listing \ref{whemb}).
+
 \begin{lstlisting}[label=whembsemLex,caption={Semantic Rules for Complementizers},float]
@@
 
-\subsection{Modality and Possible Worlds}
+\section{Modality and Possible Worlds in \prol{}}
 \label{sec:indices}
 
-\subsubsection{Representing Intensionality in \prol{}}
-
-\section{Discussion}\label{sec:discussion}
+\section{Interpretation}
+
+\chapter{Discussion}\label{sec:discussion}
 
@@
 
-\section{Appendix}
-
-\subsection{The Source Code}
-
-\subsection{Examples}
-
-\subsubsection{An Example Model} \label{sec:examplemodel}
+Nelken and Francesz, give something that might not need sets of propositions
+
+\appendix
+
+\chapter{Appendix: The Source Code}
+
+\section{\code{advertentCurt.pl}}
+
+\chapter{Appendix: Examples}
+
+\subsection{An Example Model} \label{sec:examplemodel}
 

### 2008-09-01 thesis/paper.tex
removed footnote about holesem

@@
 Blackburn and Bos present an approach utilizing underspecified representations and
-dominance relations, \pn{Hole Semantics}\footnote{Also presented in % todo:cite
-%Bos' paper
-}. 
-However none of the cases discussed so far would require the enhanced
+dominance relations, \pn{Hole Semantics}.
+However, none of the cases discussed so far would require the enhanced
 expressive power of \pn{Hole Semantics} it was decided to not change the \curt{}

### 2008-09-01 src/advertentCurt.pl
CURT WORKS !!!!!!!!!!

@@ -225,76 +225,88 @@ getDoxasticBG(X,I,world(D,F,R)) :-
 	, assert(doxastic(Rest))
 	.
 
-beAdvertent((Index,world(D,F,Background)),knowledge(X,P)) :-
+%beAdvertent((Index,world(D,F,R)),que([],[],B)) :-
+	%fail
+	%.
+
+checkPartitions(Reading,[X|Xs],F) :-
 	(
-		P = que(_,alt,Q)
-		, !
-		, (
-			getEpistemicBG(X,I,world(_,_,EBG))
-			, !
-			, BG = and(Q,EBG)
-			, NBG = and(not(Q),EBG)
-		;
-			BG = Q
-			, NBG = not(Q)
-		)
-		, backgroundKnowledge(and(BG,Background),BK2)
-		, (
-			\+ check(and(and(Background,BK2),not(BG)),'yes/ interrogative: informativity',_)
-			, !
-			, backgroundKnowledge(BG,BK)
-			, check(and(BG,BK),'preparing world: consistency', model(D2,F2))
-			, World = (X,Index,world(D2,F2,BG))
-		;
-			\+ check(and(and(Background,BK2),BG),'/no interrogative: informativity',_)
-			, backgroundKnowledge(NBG,BK)
-			, check(and(NBG,BK),'preparing world: consistency', model(D2,F2))
-			, World = (X,Index,world(D2,F2,NBG))
-		)
-		, addEpistemic(World)
+		backgroundKnowledge(and(Reading,X),BK)
+		, \+ check(and(and(Reading,BK),X),'~npartition checking: informativity',_), !
+		, checkPartitions(Reading,Xs,F)
 	;
-		P = que(Y,Domain,Body)
-		, !
-		, satisfy(some(Y,and(Domain,Body)),model(D,F),[],Result)
-		, \+ Result = undef
-		, !
-		, findall(A,satisfy(and(Domain,Body),model(D,F),[g(Y,A)],pos),Answers)
-		, realiseAnswer(Answers,que(Y,Domain,Body),world(D,F,Background),String)
-		,
-		(
-			String = none, !
-			, beAdvertent((Index,world(D,F,Background)),knowledge(X,que([],alt,some(Y,Body))))
-		;
-			Y = String
-			, (
-				getEpistemicBG(X,I,world(_,_,EBG))
-				, !
-				, BG = and(Body,EBG)
-				, NBG = and(not(Body),EBG)
[…]

### 2008-09-01 src/or.pl

@@ -30,6 +30,30 @@ disjunk(Formula,Result) :-
 	, maplist(term(Sym),PDArgs,Result)
 	.
 
+% only handles terms with ONE free variable
+% fails if there is more than one
+formulate(Variable,Term,Name,Term2) :-
+	Term =.. [Sym|Args]
+	, findVars(Variable,Name,Args,NArgs)
+	, Term2 =.. [Sym|NArgs]
+	.
+
+findVars(Variable,Name,[X|Xs],[L|Ls]) :-
+	(
+		var(X), !
+		, Variable == X
+		, findVars(Variable,Name,Xs,Ls)
+		, L = Name
+	;
+		compound(X), !
+		, formulate(Variable,X,Name,L)
+		, findVars(Variable,Name,Xs,Ls)
+	;
+		X = L
+	)
+	.
+
+findVars(_,_,[],[]).
 
 % args/2
 % takes a list of lists and returns all permutations of each list's elements

### 2008-09-01 src/advertentCurt.pl
whitespace

@@ -130,21 +130,16 @@ curtUpdate([history],[],run):- !,
    printRepresentations(H).
 
 curtUpdate(Input,Moves,run) :-
-	kellerStorage(Input,Rs)
+	kellerStorage(Input,Readings)
 	, !
-	, maplist(disjunk,Rs,Nested)
-	, flatten(Nested,Readings)
 	, updateHistory(Input)
 	, interpretReadings(Readings,Model)
-	,
-	(
+	, (
 		\+ Model = []
 		, updateModels(Model), !
 		, Moves = [accept]
 	;
-		Moves = [reject]
-	)
-	.
+		Moves = [reject]).
 
 curtUpdate(_,[noparse],run).
 

### 2008-09-01 thesis/paper.tex
FINAL

@@
 \begin{document}
-\lstset{language=Prolog,numbers=left,numberstyle=\tiny,numbersep=5pt,stepnumber=2,basicstyle=\ttfamily,identifierstyle=\textbf,keywordstyle=\underbar}
+\lstset{language=Prolog,numbers=left,numberstyle=\tiny,numbersep=5pt,stepnumber=2,basicstyle=\small\ttfamily,identifierstyle=\textbf,keywordstyle=\underbar}
 \renewcommand{\thepage}{\roman{page}}
@@
 \end{titlepage}
-\mbox{}
-\newpage
 \tableofcontents
@@
 internally, it will also only regard maximally exhaustive interpretations of
-answers.
+answers, since a maximum degree of specificity is desired. While this impairs
+the performance of  the system considerably, \acurt{} what never meant to be a
+particularly graceful implementation, but only to provide a basic proof-of-concept.
 
@@
 \a. Mia knows who snorts.\footnotesize
-\b.  \Tree
-[.$t$\\$know(?X_1.snort(X_1))(Mia)$ [.$T$ { $\lambda X.X(Mia)$\\Mia } ]
-[.$IV$\\$\lambda X.know(?X_1.snort(X_1))(X)$ [.{$IV/\bar{t}$} 
-{ $\lambda S.\lambda X.[  \lambda Y.[ know(Y)(X)](S) ]$\\knows } ]
-[.{$\bar{t}$\\$?X_1.snort(X_1)$ } [.$\bar{t}/t$ ? ] [.{ $t$\\$snort(X_1)$ }
-[.$T$ { $X_1$\\who } ]
-[.$IV$ { $\lambda X.snort(X)$\\snorts } ] ] ] ]!\qsetw{5cm} ]\normalsize
 \b. $know(?X_1.snort(X_1))(Mia)$\label{ex:formula}
@@
 but different meaning postulates. Grammar rules, later explained in
-\ref{sec:curtgrammar}, are
+\ref{sec:Syntax}, are
 responsible for determining the correct order of functional application in the
@@
 \a. Mia knows whether Vincent snorts.  \footnotesize
-\b. \Tree
-[.$t$\\$know(?.snort(Vincent))(Mia)\wedge ?.snort(Vincent)$ [.$T$ { $\lambda X.X(Mia)$\\Mia } ]
-[.$IV$\\$\lambda X.know(?.snort(Vincent))(X)$ [.{$IV/\bar{t}$} 
-{ $\lambda S.\lambda X.[  \lambda Y.[ know(Y)(X) ](S) ]$\\knows } ]
-[.{$\bar{t}$\\$?.snort(Vincent)$} [.$\bar{t}/t$ whether ] [.{ $t$\\$snort(Vincent)$ } 
-[.$T$ { $\lambda X.X(Vincent)$\\Vincent } ]
-[.$IV$ { $\lambda X.snort(X)$\\snorts } ] ] ] ]!\qsetw{5cm} ]
-\normalsize
 \b. $know(?.snort(Vincent))(Mia)$
@@
 evaluating questions has been presented, again, with respect to the current
-discourse model. Then these questions, can be embedded under certain verbs in
-certain ways. But what is the assertion a sentence as \example{Mia knows who
-snorts.} makes?
+discourse model. These questions can be embedded under certain verbs in a number
+of possible combinations. But what is the assertion a sentence such as \example{Mia
+knows who snorts.} makes?
 
@@
 that every person has their own beliefs. In \emph{knowing} or \emph{believing},
-or even \emph{imagining}, one can create an image of what it could possibly look
-like.
+or even \emph{imagining}, one can create an image of what the world could
+possibly look like.
 
@@
 
-\wh-embedding verbs embed sets of propositions, and they establish a relation
-between those propositions and their subject, one or more individuals. Depending
-on the type of verb, it can be that this individual wonders about a certain
-fact, reports it,  knows about it or believes it. In the latter two cases,
-something more happens: the proposition is asserted to be a fact of the
-individual's \emph{epistemic} or \emph{doxastic} worlds, which are worlds, just
-like any other possible world.
+Wh-embedding verbs embed sets of propositions, and they establish a relation
+between those propositions and their subject, which is typically one or more
+individuals. Depending on the type of verb, it can be that this individual
+wonders about a certain fact, reports it,  knows about it or believes it. In the
+latter two cases, something interesting happens: the proposition is asserted to be a
+fact of the individual's \emph{epistemic} or \emph{doxastic} worlds, which are
+worlds, just like any other possible world, but usually only their `owner' can
+see them. Others might only guess what they look like.
 
@@
     model of the system, $\mathcal{M}$
-    \item $d \in D$
+    \item $d \in D$, the individual this world belongs to
     \item $w'$ is a possible world, defined the same way as before: $w' =
@@
 
-\acurt{} is an experimental program. As such, it can serve as a ground for
-experimentation to find out if this definition is better suited, or any other,
-for representing the \example{true} state of affairs.
-
-%\section{Treating Disjunction}
-
-%Consider the following example dialogue with the \pn{Helpful Curt}
-%system\footnote{In a real conversation with \curt, it will usually inform the
-%user about the activity of external inferencing tools. This has been omitted and
-%marked by \code{[…]}.}:
-
-%\begin{verbatim}
-  %> Mia or Jody dances.
-  %[…]
-  %Curt: OK.
-
-  %> Who dances?
-  %[…]
-  %Curt: This question makes sense!
-
-  %Curt: maybe mia
-%\end{verbatim}
-
-%Clearly, \curt{} should give a full answer in this situation, but instead it
-%just guesses, and does not reply with something along the lines of \example{Either
-%Mia or Jody}. Unfortunately, this result does not suffice for building a system
-%that can properly treat embedded questions, since it does not reflect all
-%available information about the discourse properly.
-
-%A solution to this will again require the presence of the concept of
-%\term{possible worlds}. When uttering the phrase \example{Mia or Jody dances.}
-%then this can be understood as creating \emph{two possible scenarios}, one in
-%which there is an individual named \example{Mia}, and this individual dances,
-%and one in which there is an individual named \example{Jody} and the same holds
-%for this individual. In effect, this can be translated to two possible worlds
-%being present in the current discourse.
-
-%\begin{definition}
-  %Disjunction
-
-  %For any formula $\rho$ holding a disjunction, this formula can be expressed as a set
-  %of the constituents of the disjunction
-
-  %\[ \rho = \varphi \vee \psi, \textrm{ then } \rho' = \set{\varphi, \psi}\]
-
-  %%Every other operator or function can take a set of propositions of the type of
-  %%$\rho'$ to produce a set of propositions containing \emph{all possible
-  %%permutations} of $\rho'$. For every possible operator $\sem{$\cdot$}$ with arity
-  %%$n$ and arguments $\sem{$\cdot$}(\set{\varphi,\psi})$
-
-  %todo: more generic: just go from 1 to j, which is rho, then from j+1 to n
-
-  %\begin{eqnarray*}
-    %%\sem{$\cdot$}(\rho') \mapsto \set{\varphi,\psi}\\
-    %%\sem{$\cdot$}(\sigma_1,\dots,\sigma_j,\rho )
-    %\varphi
-    %%\sigma \wedge \set{\varphi, \psi} \mapsto \set{\sigma\wedge\varphi,\sigma\wedge\psi}\\
-    %%\sigma \to \set{\varphi, \psi} \mapsto \set{\sigma\to\varphi,\sigma\to\psi}\\
-  %\end{eqnarray*}
-%\end{definition}
-
-\section{Inquisitive Semantics}\label{sec:altq}
-
-The very latest research now focusses on providing an adequate logical semantics
-for treating questions and even discourse within a dialogue. Groenendijk, in his
-theory he calls ``Inquisitive Semantics'', argues for a completely new of of
-treating dialogues (``Dialogue Management'') -- which naturally includes a
-treatment of questions. His work on this topic has been prepared in \cite{g:is}
-and is further developed in the notes to his course held at ESSLLI in Hamburg
-(\cite{g:isdm}), just two weeks before work on this paper was to be finished.
-Because of this time constraint, and because the latter paper is to date an
-unfinished document, his theory could not be considered as a ground for the
-implementation of \acurt. His findings, however, may very well serve as a ground
-for further research on dialogue systems.
-
-\subsubsection{The Rest}
-
-\prol{} programs can be self-modifying, or dynamic. In this way the program is
-able to change its own state and thus store information.
-
-\subsubsection{Disjunction and Ambiguity}
-Consider the following sentences:
-
-\ex. \label{ex:disjunction} \a. Mia or Jody dances
-\b. Either Butch snores or Vincent is using the chainsaw
-
-\pn{Helpful Curt}, with its unpartitioned logical space and its lack of
-the awareness of more than one (possible) world, would generate only one model
-to represent the facts of \ref{ex:disjunction}. It is therefore only able to
-generate \emph{one} possible answer to both questions, effectively accounting only
-for one possible constituent of the disjunction, prefixing it with a `maybe' to
-indicate its uncertainty.
-
-Another example where \curt{} has to sacrifice exhaustiveness over its one-world
-representation of the world is given in \ref{ex:ambig}
-
-\ex.\label{ex:ambig} Every boxer likes a woman.
-
-This will generally generate two readings, but only one model -- with only one
-world. And further discourse will have to choose between one of the readings and
-stick with the minimal model constructed by the model builder. This is not
-really satisfactory and extending the dimensions of knowledge within \curt{} seems
-necessary in order to treat such examples.
-
-The previous work on \acurt{} enables it to generate a \term{choice point} -- and
-moreover, not pursue only one path of it, but both at the same time, by
-generating more than one world in its model.
-Unfortunately, this will build up very large models over time and computations
-with it might prove to be very complex and resource demanding. Adding such a
-functionality to the system, however, is not a hard task and hopefully the
-technique employed here can be further refined. %todo: maybe add a footnote
-%about future prospects or so
-
-We first derive know(john,all(X,imp(woman(X),love(X,john)))) from the syntax of
-john knows whether every woman loves john, then we see if
-all(X,imp(woman(X),love(X,john))) is valid (uninformative) and ridicule John, if
-it's not valid and not consistent, assert his knowledge of the fact that it's
-not true. If it's true, we assert he knows that.
-
-For whether we can ridicule him if it's not true (inconsistent).
-The MB/TP do a comparison \Disc to all(X,imp(woman(X),love(X,john)))
-
-An important observation to be made here is that only \emph{assertive} verbs can
-embed real questions. \emph{Stipulating} verbs only embed \emph{propositions} as
-in \ref{ex:stipul}.
-Nevertheless, both are treated in the following analysis, but with a focus on
-the treatment of assertive \wh-embedding verbs.
-
-When treating questions one will be confronted with many different
-categorizations of questions. One of the first distinctions to make is the one
-between \term{embedded questions} as in \ref{ex:embed} and \term{direct
-questions} shown in \ref{ex:direct}.
-
-As indicated already by the title, this paper will concern itself with the
-treatment of the former and only briefly comment on the latter. The reasons for
-this are numerous  %fixme I need a synonym for that
-and this section is going to name the most important ones.
-
-Early literature, such as \cite{karttunen:1977} only regards embedded
-\wh-phrases, treating direct questions as semantically equivalent. To quote
-Karttunen:
-
-\ex. \a. John knows \emph{whether this sentence contains an embedded
-question}.\label{ex:embed}
-\b. Does this sentence pose a direct question?\label{ex:direct}
-
-The present work will rely on \cite{gs:q} for providing the main guidelines for
-the theoretic aspects and also lend from earlier works, most importantly
-\cite{gs:sqpa}\footnotemark. Later on, in section \ref{sec:altq} we will also
-discuss the very newest (at the time of writing) developments in question
-semantics, depicted in \cite{g:is}.
-
-\footnotetext{\cite{gs:sqpa} is actually a series of three articles
-that have appeared before. Among them, we will mostly consider \cite{gs:sawhq}.
-Both, the earlier paper and the later dissertation (and many other papers
-originating in Amsterdam) have been kindly made
-available electronically, free of charge, by the Universiteit van Amsterdam at
-the following address: \url{http://dare.uva.nl}}
-
 \chapter{Implementation} \label{sec:implementation}
-
-Originally, \prol{} was conceptualized as a programming language for a subset of
-first-order logic. This entails the predominant usage of \abbr{FOL} in section
-\ref{sec:implementation}, but some higher order predicates introduced to the
-language in recent years will also be put to use.
-
-\section{Advertent Curt}
-
-\acurt{} lends its name from the fact that it is not only aware of its own
-knowledge and presuppositions, but `knows' about what individuals in its
-model of the world are aware of. It is not in this sense a question
-\emph{answering} system, but it may be seen as an early effort towards providing
-such a system on a more sophisticated scale. While section \ref{sec:altq} shows
-how direct questions can be integrated into the system with relative ease, focus
-is still on providing a feasible method of implementing intensionality in what
-is mostly a first-order logic computing environment.
-
-It is also not constrained to being useful in a particular domain, like
-\pn{CHAT-80} is.
-
-\section{The Syntax of Embedded Questions}
-
-The internal structure of the subset of interrogative sentences that will be
+%\section{Advertent Curt}
+
+%\acurt{} lends its name from the fact that it is not only aware of its own
+%knowledge, but `knows' about what individuals in its
+%model of the world are aware of. It is not in this sense a question
+%\emph{answering} system, but it may be seen as an early effort towards providing
+%such a system on a more sophisticated scale.
+
+With the theoretical background of the previous chapter in mind, the discussion
+can now turn to the implementation of some of the more interesting concepts
+presented. In particular, this chapter provides reference implementations of a
+grammar that is able to parse simple embedded questions in section
+\ref{sec:Syntax}, and then goes on to explain \curt's refaced discourse model in
+section \ref{sec:indices}. Finally, section \ref{sec:inference} explains how the
+inferencing system deals with the new extensions.
+
+\section{The Syntax of Embedded Questions}\label{sec:Syntax}
+
+The internal structure of the subset of interrogative sentences that is
 considered here is not a very complicated one. Since yes/no interrogatives  embed
@@
 
-\pn{Keller Storage} is already quite capable
-of dealing with more complex constituents such as nested noun phrases, but it
-has its limitations. In particular, it can only handle quantifier scope
-ambiguities and does not take negation into account. Moreover, it overgenerates
-with some quantifier scope ambiguities\footnote{\cite{blackburnbos:cl1} mention
-sentences such as \ref{ex:kellersucks}, where Keller Storage produces six
-instead of the desired five readings.
-\ex. One criminal knows every owner of a hash bar\label{ex:kellersucks}
+\pn{Keller Storage} is already quite capable of dealing with more complex
+constituents such as nested noun phrases, but it has its limitations. In
+particular, it can only handle \emph{quantifier} scope ambiguities and does not
+take other ambiguities, such as negation into account. Moreover, it
+overgenerates with some quantifier scope
+ambiguities\footnote{\cite{blackburnbos:cl1} mention sentences such as
+\ref{ex:kellersucks}, where Keller Storage produces six instead of the desired
+five readings.  \ex. One criminal knows every owner of a hash
+bar\label{ex:kellersucks}
 
@@
 
-The \code{whemb} and \code{ivtbar} categories have a \code{type} parameter, that dictates
-their behaviour, although their respective type systems never cooperate with
-each other. The type of the \code{whemb} category can hold the values
-\code{propos}, for a proposition embedding complementizer, like \example{that},
-and \code{question}, for a complementizer that turns a proposition into a yes/no
-interrogative, like \example{if} and \example{whether} (listing \ref{whemb}).
-
 \begin{lstlisting}[label=whembsemLex,caption={Semantic Rules for Complementizers},float]
@@
 
+The \code{whemb} and \code{ivtbar} categories have a \code{type} parameter that dictates
+their behaviour, although their respective type systems never cooperate with
+each other. The type of the \code{whemb} category can hold the values
+\code{propos}, for a proposition embedding complementizer, like \example{that},
+and \code{question}, for a complementizer that turns a proposition into a yes/no
+interrogative, like \example{if} and \example{whether} (listing \ref{whemb}).
+
+\begin{lstlisting}[label=tbar,caption={Semantic Rules for $\bar{t}$},float]
+tbar([sem:TBar]) -->
+    whemb([sem:WHemb])
+    , s([coord:no,sem:S])
+    , { combine(tbar:TBar,[whemb:WHemb,s:S]) } .
+
+tbar([sem:TBar]) -->
+    s([coord:no,sem:S])
+    , { combine(tbar:TBar,[s:S]) } .
+
+tbar([sem:TBar]) -->
+    q([sem:Q])
+    , { combine(tbar:TBar,[s:Q]) } .
+\end{lstlisting}
+
 The types now determine whether to produce an embedded yes/no
@@
 
-\begin{lstlisting}[label=tbar,caption={Semantic Rules for $\bar{t}$},float]
-tbar([sem:TBar]) -->
-    whemb([sem:WHemb])
-    , s([coord:no,sem:S])
-    , { combine(tbar:TBar,[whemb:WHemb,s:S]) } .
-
-tbar([sem:TBar]) -->
-    s([coord:no,sem:S])
-    , { combine(tbar:TBar,[s:S]) } .
-
-tbar([sem:TBar]) -->
-    q([sem:Q])
-    , { combine(tbar:TBar,[s:Q]) } .
-\end{lstlisting}
-
 \begin{lstlisting}[label=tbarCombine,caption={$\bar{t}$ is responsible for beta-converting the embedded sentence.},float]
@@
     betaConvert(S,SRed).
-
 combine(tbar:[SRed],[s:S]) :-
@@
 
-\section{Modality and Possible Worlds in \prol{}}
-\label{sec:indices}
-
-\section{Interpretation}
+\section{Modality and Possible Worlds in \prol{}} \label{sec:indices}
+
+\prol{} programs can be self-modifying. In this way the program is able to
+change its own state and thus store information. The \lstinline!assert/1!,
+\lstinline+abolish/1+ and \lstinline!retract/1! goals are typically used to access a
+program's logic directly. \curt{} uses these goals to store information about
+the current discourse, models and the dialogue's history.
+
+Furthermore, the internal representations of models and readings have been
+folded into one entity. Every world is assigned the reading that
+originally generated it via the model builder that was used during inferencing
+tasks.
+
+A slightly refurbished notion of discourse has been installed. It allows
+\acurt{} to perform computations on arbitrarily large numbers of possible worlds
+without losing information by having to decide on one reading. Thus the readings
+of ambiguous sentences such as \example{Every boxer likes a woman} can be
+pursued in parallel.  As soon as a new input sentence conflicts with one
+possible world, this world is \example{dropped}, with computation continuing on
+the other worlds. Listing \ref{interpret} shows how this works for one
+particular world and one particular reading. The \code{interpretReadings/3}
+predicate is responsible for assigning subtasks to \code{interpret/3}, and can
+be looked up in the appendix. The \code{check/3} predicate is a trivial abstraction
+over the original calls to the inferencing engines and is  also available in the
+appendix.
+
+\begin{lstlisting}[label=interpret,caption={The interpretation code for one
+  particular world and reading},float]
+interpret((Index,Old),New,World) :-
+  (
+    beAdvertent((Index,Old),New), !
+    , World = (Index,Old)
+  ;
+    getKnowledge(Old,New,BK,Reading) , (
+      check(and(BK,New),'consistency',BBModel), !  , (
+        check(and(BK,not(New)),'informativity',_), !
+        , BBModel = model(D,F)
+        , World = (Index,world(D,F,Reading))
+      ;
+        format('~nFound uninformative reading. Dropping reading.',[])
+        , World = (Index,Old)
+      )
+    ;
+      format('~nFound inconsistency. Dropping world.',[])
+      , World = [])) .
+\end{lstlisting}
+
+\subsection{Epistemic and Doxastic Alternatives}\label{sec:inference}
+
+While the original
+implementation provided only one set of models, using \lstinline!models/1!, this
+implementation introduces two
+new dynamic goals, \lstinline!epistemic/1! and \lstinline!doxastic/1!. They
+serve the obvious purpose of accumulating what \curt{} knows about other
+individual's knowledge. Listing \ref{getEpistemicBG} shows the code for the
+retrieval of epistemic backgrounds. The one for doxastic backgrounds is very
+similar and omitted here. The goal fails if it cannot unify with any particular
+background given its instantiations, which are the name of the entity in
+question (\code{X}) and the index of a possible world in the main curt storage
+(\code{I}).
+
+\begin{lstlisting}[label=getEpistemicBG,caption={A Predicate to Retrieve Epistemic
+  Backgrounds},float]
+getEpistemicBG(X,I,world(D,F,R)) :-
+  epistemic(E), !
+  , delete(E,(X,I,world(D,F,R)),Rest)
+  , \+ Rest = E
+  , retract(epistemic(_)), !
+  , assert(epistemic(Rest)) .
+\end{lstlisting}
+
+\subsection{Interpretation of \wh-embedded Verbs}
+
+The heart of the reasoning of \acurt{} lies within the predicate
+\code{beAdvetent/3}. Listing \ref{beAdvertent} reproduces one of its clauses,
+this one specifically dealing with answers to constituent interrogatives. First,
+\code{findBuddies/2} returns a list of all individuals in the current discourse.
+Then \code{formulate/4} is called on each name for the body and domain of the
+question. This predicate instantiates an unsaturated formula with a name it is
+given, to produce, e.g. $boxer(butch)$ from $butch$ and $boxer(X)$. It is
+important to note that to date, \acurt{} can only deal with named individuals,
+and not with quantifications over individuals, such as $\exists x[boxer(x)]$.
+
+\begin{lstlisting}[label=beAdvertent,caption={The Core of Advertent Curt Dealing
+  with Constituent Interrogatives},float]
+beAdvertent((_Index,world(_D,F,R)),que(X,Domain,Body),Answer) :-
+  findBuddies(F,Buddies)
+  , maplist(or:formulate(X,and(Domain,Body)),Buddies, Formulae)
+  , partition(Formulae,Parted)
+  , checkPartitions(R,Parted,Answer) .
+\end{lstlisting}
+
+Next, the combinatorial bomb is armed with \code{partition/2}, which generates
+all possible logical partitions of the domain and body of the question with
+respect to the discourse referents it is given. Afterwards,
+\code{checkPartitions/3} is called and ignites the inferencing tools by calling
+them once for each partition, unified with the current discourse of the world
+(held in the variable \code{R}) to find the one most specific partition, which
+serves as the only true answer to the question.
+The code for the helper predicates is again available in the appendix. Other,
+more simple clauses of \code{beAdvertent/3} deal with embedded propositions and
+yes/no interrogatives and are not reproduced here.
+
+Listing \ref{beA2} shows how \acurt{} handles embedded interrogatives. It first
+calls the predicate shown above, then retrieves the individual's knowledge
+background and takes that as a basis for further computation. First, the
+background knowledge is computed, then the resulting formula is checked by the
+inferencing tools -- this time with the individual's epistemic background. This
+ensures that an individual cannot hold contradictory beliefs. The check also
+produces a model, which is then passed on to the epistemic background of
+\code{X}.
+
+\begin{lstlisting}[label=beA2,caption={Treatment of Embedded Interrogatives of
+  Knowledge},float]
+beAdvertent((Index,world(D,F,Background)),knowledge(X,que(Y,Domain,Body))) :-
+	! , beAdvertent((Index,world(D,F,Background)),que(Y,Domain,Body),Answer)
+	, (
+		getEpistemicBG(Y,_,world(_,_,EBG))
+		, !  , Q = and(Answer,EBG)
+	;
+		Q = Answer
+	)
+	, backgroundKnowledge(Q,BK)
+	, check(and(Q,BK),'preparing world: consistency', model(D2,F2))
+	, World = (X, Index,world(D2,F2,Q))
+	, addEpistemic(World) .
+\end{lstlisting}
 
@@
 
-We have seen, the \curt{} system is quite capable of treating questions this way.
-Having implemented modality and possible worlds in such a fashion, let's go on
-and implement some even cooler stuff (tense, aspect, negation, comparatives,
-anaphora). Basing all this on a nice implementation of DRT (discussed in
-\cite{kampreyle:drt}) like the one given in \cite{blackburnbos:cl2} might be
-interesting. The Pronto \abbr{QA} system of \cite{prontoqa} already defines a
-means of representing questions and queries against a Discourse Representation
-Structure.
-
-Burhans and Shapiro
-
-Nelken and Francesz, give something that might not need sets of propositions
+This thesis presented a method of approaching question answering in a
+light-weight prototype of a dialog system. As such, not all results are
+satisfactory, and in order to cover the semantics of at least some basic human
+conversation, \acurt{} will have to grow considerably. Yet it is debatable
+whether the \curt{} system can serve as a ground to a full-fledged dialogue
+system, since it was only planned an designed as a teaching system for
+\cite{blackburnbos:cl1}. It would have to be modified at many places, provided
+with a more exhaustive ontology, and better parser. \curt{} uses nothing more
+than the \prol{} in-built \abbr{DCG} parser, which was not designed to cover a
+wide range of natural language syntax.
+
+\acurt{} can deal with basic dialogues such as the following:
+\begin{quote}
+  Mia and Vincent snort. Butch does not snort. Butch is not Vincent. Butch is
+  not Mia.
+
+  Who snorts? Vincent knows who snorts. Mia knows if Butch snorts.
+\end{quote}
+
+In this scenario, \curt{} will correctly answer that Mia and Vincent snort, but
+not Butch. It will also acknowledge that Vincent knows about this fact. Finally,
+it will also record that Mia knows that Butch does not snort. Note that when
+using its ontology, the system will not need the information of the third and
+fourth sentence. The sentences will then still show how \curt{} can distinguish
+between informative and uninformative readings even with respect to its
+background knowledge.
+
+The extension of the system could still use some improvements to address its
+present poor usability. Currently, \acurt{} will believe everything it is told,
+and when it finds a contradiction, it will just eliminate the offending worlds
+from its storage. Thus, when the user inadvertently enters a contradiction,
+\acurt{} will end up in a state of ignorance.
+
+Furthermore, while \curt{} does \emph{acknowledge} the epistemic and doxastic
+background of the individuals it keeps track of, it does so far not do anything
+useful with them, other than display them. However, it should not pose a
+significant challenge to implement a functionality that would be able to answer
+a question such as \example{What does Vincent believe?}
+
+A fully-capable dialogue system would have to encompass a much wider range of
+functionality. At least a rudimentary notion of tenses and events, and a solid way of
+treating anaphoric references seem to be a bare minimum of what such a system
+should be capable of when talking to a human. But as it stands, \curt{} is not
+even able to cope with negation properly. For example, it cannot interpret a
+sentence such as \example{Vincent does not know who snorts.} And what is this
+sentence supposed to mean? This could be an interesting future topic in the
+research of embedded questions.
+
+Discourse Representation Theory, as introduced in
+\cite{kampreyle:drt} could probably solve some of the issues with \curt{}. It
+provides a solid treatment of anaphoric references, is directly translatable
+into \abbr{FOL} and Kamp \& Reyle also propose methods of dealing with plurals,
+tense and negation in \abbr{DRT}.
+\cite{blackburnbos:cl2} presents an approach to implementing \abbr{DRT} in
+\prol{}. Furthermore, the \pn{Pronto QA System} of \cite{prontoqa} already uses
+\abbr{DRT} for its inferencing tasks, and so does the implementation proposed by
+\cite{bosgabdsil}.
+
+\subsection{Inquisitive Semantics}\label{sec:altq}
+
+The very latest research now focusses on providing an adequate logical semantics
+for treating questions and even discourse within a dialogue. Groenendijk, in his
+theory he calls ``Inquisitive Semantics'', argues for a completely new way of
+treating dialogues (``Dialogue Management'') -- which naturally includes a
+treatment of questions. His work on this topic has been prepared in \cite{g:is}
+and is further developed in the notes to his course held at ESSLLI in Hamburg
+(\cite{g:isdm}), just two weeks before work on this paper was to be finished.
+Because of this time constraint, and because the latter paper is to date an
+unfinished document, his theory could not be considered as a ground for the
+implementation of \acurt. His findings, however, may very well serve as a ground
+for further research on dialogue systems.
+
+%Burhans and Shapiro
+
+%Nelken and Francesz, give something that might not need sets of propositions
 
@@
 
-\section{\code{advertentCurt.pl}}
-
-\chapter{Appendix: Examples}
-
-\subsection{An Example Model} \label{sec:examplemodel}
-
-\begin{eqnarray*}
-  M &=& \set{\pair{1,\pair{D_1,F_1,\varphi_1}},\pair{2,D_2,F_2,\varphi_2}} \\
-  D_1 &=& \set{d_1,d_2,d_3} \\
-  F_1 &=& \{f(0,mia,d_1), f(0,vincent,d_2), f(0,jody,d_3),\\
-  && f(1,woman,\set{ d_1,d_3 }), f(1,man,\set{ d_2 }),
-  f(2,like,\set{\pair{d_2,d_1},\pair{d_3,d_2}})\} \\
-  \phi &=& \exists x,y,z [mia(x)\wedge woman(x)\wedge \wedge vincent(y)\wedge
-  man(y)\wedge jody(z)\wedge woman(z) \\&&\wedge x \neq y \wedge x\neq z \wedge z
-  \neq y\wedge like(x)(y)\wedge like(y)(z)]\\
-  D_2 &=& 
-\end{eqnarray*}
+\lstset{basicstyle=\tiny\ttfamily}
+
+\section{Curt and Helpers}
+\subsection{\code{advertentCurt.pl}}
+
+\lstinputlisting{../src/advertentCurt.pl}
+
+\subsection{\code{or.pl}}
+
+\lstinputlisting{../src/or.pl}
+\section{Grammar}
+\subsection{\code{semLexStorage.pl}}
+\lstinputlisting{../src/semLexStorage.pl}
+
+\subsection{\code{semRulesKeller.pl}}
+\lstinputlisting{../src/semRulesKeller.pl}
+
+\subsection{\code{englishGrammar.pl}}
+\lstinputlisting{../src/englishGrammar.pl}
+
+\subsection{\code{englishLexicon.pl}}
+\lstinputlisting{../src/englishLexicon.pl}
 

### 2008-09-01 src/advertentCurt.pl
FINAL

@@ -178,12 +178,8 @@ interpret((Index,Old),New,World) :-
 		beAdvertent((Index,Old),New), !
 		, World = (Index,Old)
 	;
-		getKnowledge(Old,New,BK,Reading)
-		,
-		(
-			check(and(BK,New),'consistency',BBModel), !
-			,
-			(
+		getKnowledge(Old,New,BK,Reading) , (
+			check(and(BK,New),'consistency',BBModel), !  , (
 				check(and(BK,not(New)),'informativity',_), !
 				, BBModel = model(D,F)
 				, World = (Index,world(D,F,Reading))
@@ -193,10 +189,7 @@ interpret((Index,Old),New,World) :-
 			)
 		;
 			format('~nFound inconsistency. Dropping world.',[])
-			, World = []
-		)
-	)
-	.
+			, World = [])) .
 
 getKnowledge([],New,BackgroundKnowledge,New) :- 
 	backgroundKnowledge(New,BackgroundKnowledge).
@@ -234,8 +227,6 @@ checkPartitions(Reading,[X|Xs],F) :-
 		, F = X
 	)
 	.
-
-
 
 beAdvertent((_Index,world(_D,F,R)),que(X,Domain,Body),Answer) :-
 	findBuddies(F,Buddies)
